### 2023-10-27 article.qmd
Add article.qmd

---
title: "SimBio: a static typing, modern Python package to simulate reaction networks"
format:
  html:
    number-sections: true
    anchor-sections: true
  pdf:
    documentclass: scrartcl
    papersize: a4
    geometry:
        - margin=20mm
    pdf-engine: tectonic
    number-sections: true
    colorlinks: true
---

# Introduction

Chemical Reaction Networks (CRN) models
species and rates
Several modalities (ODE, SDE and jump process)
writing equations directly is error-prone and not easily reusable
Several packages exist to simulate them

Biosimulators.org, a registry of simulation tools, list at least 15 Python softwares.
(some) wrappers to external libraries
script the model definition
avoid exporting to csv and loading in Python
non-extendable from python
large community

Writing models
GUI (e.g., COPASI)
DSL (e.g, Antimony)
Python (e.g, PySB)
Benefits of IDE integration (syntax highlighting, etc)

Here we present SimBio, a Python package
that leverages standard modern Python syntax to integrate well with IDEs and static analysis tools.
It is written in pure Python such that it can be easily introspected and extended from Python.

It focused on providing a user experience, maintainability, and scalability

Clean API to create tools (graphs, tables, etc)

# Results

The code to define and simulate reaction systems
is split into three Python packages:
symbolite, to create symbolic expressions;
poincaré, to define dynamical systems;
and simbio, to define reaction systems
and interface with systems biology standards such as SBML.

## Symbolite

Symbolite is a lightweight symbolics package
to create and inspect mathematical expressions,
which can be compiled to different backends.
Currently, we have implementations for NumPy, SymPy and JAX,
but it is easy to add new backends.

## Poincaré

Poincare is a package to define and simulate dynamical systems.
It provides a `System` class,
where we can define `Constant`s, `Parameter`s and `Variable`s,
and create equations linking a variable's derivative with an expression:

```python
class Decay(System):
    x: Variable = initial(default=1)  # Variable with initial condition
    eq = x.derive() << -x  # equation assigning dx/dt = -x
```

It allows to define higher-order systems
by assigning an initial condition to a derivative:

```python
class Oscillator(System):
    x: Variable = initial(default=1)
    v: Derivative = x.derive(initial=0)
    eq = v.derive() << -x
```

The benefits of using a class to define the system are multiple:

1. the variable name to which a component is assigned can be automatically saved in the component for introspection (i.e., `Oscilator.x.name == "x"`),
2. it provides a namespace such that we can easily define multiple independent models in the same script,
3. it allows IDEs to provide autocomplete and refactoring capabilities (`Oscillator.<TAB>` shows `x`, `v` and `eq`),
4. it allows to create instances which can be composed into a bigger model.

For the latter,
it is important to add the type annotations,
as it can provide a signature autocompletion in IDEs supporting the `dataclass_transform` (PEP 681).

```python
class BigModel(System):
    x: Variable = initial(default=1)
    subsystem = Oscillator(x=x)
    independent = Oscillator(x=1)  # independent variable created inside
```

Using type annotations also allows to statically check for typing errors.
For instance, `Parameter`s can be assigned either a number or a time-dependent expression.
But can also be linked to an outside variable when composing models.
As such, they cannot be used to parametrize the initial conditions of variables,
for which we have to use a `Constant`:

```python
class Model(System):
    k: Constant = assign(default=1, constant=True)
    # two variables sharing initial condition
    x: Variable = initial(default=c)
    y: Variable = initial(default=c)

    p: Parameter = assign(default=1)
    z: Variable = initial(default=p)  # typing error
```

To simulate a system,
we create a `Simulator` instance,
which compiles a given system and interfaces with solvers.
By default, it creates a first-order ordinary differential equation (ODE) system
using `numpy` as a backend.
The Simulator wraps the output in a `pandas.DataFrame`,
which can be easily plotted with the standard `plot` method.

```python
sim = Simulator(Oscillator)
df = sim.solve(save_at=np.linspace(0, 5, 100))

df
```

```python
df.plot()
```

## SimBio

simbio: define and simulate reaction networks, mass-action, SBML, SED-ML
simulate system with predefined reactions, and stoichiometries
load from biomodels and run
Cool stuff
Interactive
Pandas
Generate from sbml?
Speed leveraging backends

# Conclusion

Modular and extensible set of python libraries
Leverage IDE integration and static typing for model creation

Introduce 3 layers: symbolite, poincare and simbio
easier to understand and develop
usable, extendable, and maintainable by a broader community
protocols
change in the future

### 2023-10-30 article.qmd
Improve writing

@@
 
-The code to define and simulate reaction systems
-is split into three Python packages:
+Modular code architecture enhances reusability, extendability and maintainability for the community.
+Hence, we split the code to define and simulate reaction systems into three Python packages:
 symbolite, to create symbolic expressions;
@@
 
-Symbolite is a lightweight symbolics package
-to create and inspect mathematical expressions,
-which can be compiled to different backends.
-Currently, we have implementations for NumPy, SymPy and JAX,
+Symbolite is a lightweight symbolics package to create algebraic mathematical expressions.
+They can be inspected and compiled to different backends.
+Currently, we have implementations for NumPy (and Numba), SymPy and JAX,
 but it is easy to add new backends.
@@
 It provides a `System` class,
-where we can define `Constant`s, `Parameter`s and `Variable`s,
+where we can define `Constant`s, `Parameter`s, `Variable`s,
 and create equations linking a variable's derivative with an expression:
@@
 It allows to define higher-order systems
-by assigning an initial condition to a derivative:
+by assigning an initial condition to a `Derivative`:
 

### 2023-10-30 article.qmd
Write simbio section

@@
 
-simbio: define and simulate reaction networks, mass-action, SBML, SED-ML
-simulate system with predefined reactions, and stoichiometries
-load from biomodels and run
+For reaction networks,
+we are usually interested in first-order differential equations
+relating the rate of change of the network's species.
+
+To facilitate defining these models,
+SimBio defines `Species`,
+which comprises of a `poincare.Variable` and an stoichiometry,
+and `Reaction`s,
+which transform reactants into products taking into account the stoichiometry.
+
+```python
+class Model(Copmartment):
+    """2A -> B"""
+    A: Species = initial(default=1)
+    B: Species = initial(default=0)
+    r = Reaction(reactants=[2 * A], product=[B], rate=1)
+```
+
+It also includes a `MassAction` variant,
+which automatically adds the reactants as part of the rate.
+`MassAction(reactants=[2 * A], product=[B], rate=1)`
+is equivalent to
+`Reaction(reactants=[2 * A], product=[B], rate=1 * A**2)`,
+and many predefined commonly used reactions reactions.
+
+SimBio also supports importing from and exporting to SBML,
+or directly from BioModels:
+
+```python
+from simbio import Simulator
+from simbio.io import biomodels, sbml
+
+model = sbml.load("repressilator.sbml")
+model = biomodels.load_model("BIOMD12")  # repressilator model
+Simulator(model).solve(...)
+```
+
 Cool stuff

### 2023-11-10 article.qmd
Mention numba backend in poincaré section
Change title

@@
 ---
-title: "SimBio: a static typing, modern Python package to simulate reaction networks"
+title: "SimBio: a flexible and extensible Python package for systems modelling"
 format:
@@
 
+Switching backends to `"numba"` can improve model runtime
+on a factor upto x30 or more for big or long running models,
+but incurs in a compile time penalty for the first run
+which must be taken into account.
+
 ## SimBio

### 2023-11-10 article.qmd
Extend introduction

@@
 
-Chemical Reaction Networks (CRN) models
-species and rates
-Several modalities (ODE, SDE and jump process)
-writing equations directly is error-prone and not easily reusable
-Several packages exist to simulate them
+Chemical Reaction Networks (CRN)
+are the basis of many models used across the biological sciences and epidemiology.
+They consist of a set of species and reactions that mediate conversions between them.
+Several modalities exist to simulate these systems:
+deterministic Ordinary Differential Equations (ODE) which model a macroscopic behavior,
+Stochastic Differential Equations (SDE) which model microscopic fluctuations,
+and jump processes (Gillespie-like simulations) which take into account the discreteness of the populations.
+Hence, it is useful to write these models in a higher level description
+which can be reused for these different types of simulations.
+equations directly is error-prone and not easily reusable
 
-Biosimulators.org, a registry of simulation tools, list at least 15 Python softwares.
-(some) wrappers to external libraries
-script the model definition
+Several tools already exist to define and simulate these systems.
+In particular,
+Biosimulators.org,
+a registry of simulation tools,
+list at least 15 Python softwares
+among which are included
+COPASI,
+Tellurium,
+GillesPy2
+PySB.
+However,
+some are only Python wrappers to external (C, C++) libraries.
+They allow to "control" the model definition and simulation in a Python script,
+but it is not possible to use existing Python packages
+to change the way the simulation is carried out.
+In other words, they are not extensible from Python.
+
 avoid exporting to csv and loading in Python
-non-extendable from python
 large community
 
-Writing models
-GUI (e.g., COPASI)
-DSL (e.g, Antimony)
-Python (e.g, PySB)
+To write these models,
+there are three main approaches.
+Tools such as COPASI provide a Graphical User Interface (GUI),
+where the components, relations and their values are defined.
+Others use a Domain Specific Language (DSL),
+such as Antimony,
+which is a less verbose version of the System Biology Markup Language (SBML).
+The last approach is to use a programming language such as Python,
+which is the approach used by PySB.
+
 Benefits of IDE integration (syntax highlighting, etc)
 
-Here we present SimBio, a Python package
-that leverages standard modern Python syntax to integrate well with IDEs and static analysis tools.
-It is written in pure Python such that it can be easily introspected and extended from Python.
+Here we present SimBio, a Python package to define and simulate reaction networks.
+From the end-user point of view,
+it is focused on providing an ergonomic experience
+by integrating well with IDEs and static analysis tools
+through the use of standard modern Python syntax.
+From the power-user point of view,
+it can be extended from Python,
+since its coded in pure Python.
+Finally,
+from the developer/maintainer point of view,
+it has a modular architecture with a clear separation of concerns,
+making it easier to maintain or to contribute new code.
 
-It focused on providing a user experience, maintainability, and scalability
-
+the models can be easily introspected,
+and simulation extended from Python.
 Clean API to create tools (graphs, tables, etc)

### 2023-11-10 article.qmd
Remove covered points
Add comments to SBML code

@@
 
-model = sbml.load("repressilator.sbml")
-model = biomodels.load_model("BIOMD12")  # repressilator model
+model = sbml.load("repressilator.sbml")  # from existing SBML file
+model = biomodels.load_model("BIOMD12")  # from BioModels
 Simulator(model).solve(...)
@@
 
-Cool stuff
-Interactive
-Pandas
-Generate from sbml?
-Speed leveraging backends
+## Other things
 
-# Conclusion
+units -> Pint
+interactive
+
+# Discussion
 

### 2023-11-10 article.qmd
Include "error-prone" into first paragraph

@@
 Several modalities exist to simulate these systems:
-deterministic Ordinary Differential Equations (ODE) which model a macroscopic behavior,
-Stochastic Differential Equations (SDE) which model microscopic fluctuations,
-and jump processes (Gillespie-like simulations) which take into account the discreteness of the populations.
-Hence, it is useful to write these models in a higher level description
-which can be reused for these different types of simulations.
-equations directly is error-prone and not easily reusable
+deterministic Ordinary Differential Equations (ODE) to model a macroscopic behavior,
+Stochastic Differential Equations (SDE) to model microscopic fluctuations,
+and jump processes (Gillespie-like simulations) to take into account the discreteness of the populations.
+Instead of writing the equations directly for each of these formulations
+which is error-prone and hard to reuse,
+these models can be written in a higher level description
+which can be compiled for these different types of simulations.
 

### 2023-11-10 article.qmd
Add references

@@
 title: "SimBio: a flexible and extensible Python package for systems modelling"
+bibliography: bibliography.bib
 format:
@@
 among which are included
-COPASI,
-Tellurium,
-GillesPy2
-PySB.
+COPASI [@hoopsCOPASICOmplexPAthway2006],
+Tellurium [@choiTelluriumExtensiblePythonbased2018],
+GillesPy2 [@matthewGillesPy2BiochemicalModeling2023]
+PySB [@lopezProgrammingBiologicalModels2013].
 However,
@@
 change in the future
+
+# References
+
+::: {#refs}
+:::

### 2023-11-10 article.qmd
Rewrite of the introduction

@@
 In particular,
-Biosimulators.org,
+BioSimulators.org [@shaikhBioSimulatorsCentralRegistry2022],
 a registry of simulation tools,
-list at least 15 Python softwares
-among which are included
-COPASI [@hoopsCOPASICOmplexPAthway2006],
-Tellurium [@choiTelluriumExtensiblePythonbased2018],
-GillesPy2 [@matthewGillesPy2BiochemicalModeling2023]
-PySB [@lopezProgrammingBiologicalModels2013].
-However,
-some are only Python wrappers to external (C, C++) libraries.
-They allow to "control" the model definition and simulation in a Python script,
-but it is not possible to use existing Python packages
+list at least 15 Python softwares.
+Some of them are:
+
+1. COPASI [@hoopsCOPASICOmplexPAthway2006],
+a stand-alone software
+which provides a Graphical User Interface (GUI) to define and simulate models.
+2. Tellurium [@choiTelluriumExtensiblePythonbased2018],
+which defines itself as a Python-based modeling environment
+and uses a C++ library called libRoadRunner [@welshLibRoadRunnerHighPerformance2023] to simulate them.
+3. PySB [@lopezProgrammingBiologicalModels2013],
+a Python library
+that created a DSL using standard Python to define models
+which are then compiled to ODE using a Perl library called BioNetGen [@harrisBioNetGenAdvancesRulebased2016].
+
+A limitation of these tools is their extensibility from Python,
+as they are or use wrapped libraries in other languages.
+As such, they allow to "control" the model definition and simulation in a Python script,
+but we cannot leverage other packages from the huge Python community
 to change the way the simulation is carried out.
-In other words, they are not extensible from Python.
+In some sense, we could say that
+they facilitate not having to export the simulation result to a CSV file
+and then loading it in Python to continue the analysis.
 
-avoid exporting to csv and loading in Python
-large community
+Another weak point is related to the way the models are written.
+While most tools support standards such as the System Biology Markup Language (SBML) [@huckaSBMLL3V2],
+no one is expected to write an SBML file directly.
+Tellurium uses a Domain Specific Language (DSL) called Antimony [@smithAntimonyModularModel2009],
+which can be translated to SBML.
+A DSL permits to reuse of the code *as is* in different programming environments,
+but will not be recognized in Integrated Development Enviromnents (IDE)
+and, then, cannot provide the development experience we have come to expect:
+syntax highlighting, code completion, refactoring, static analysis, among others.
+This has been recognized by the community,
+and an extension for Visual Studio Code has been created as part of a master thesis [@vscode-antimony].
+But maintaining it into the future could be burdensome for the smaller systems biology community.
 
-To write these models,
-there are three main approaches.
-Tools such as COPASI provide a Graphical User Interface (GUI),
-where the components, relations and their values are defined.
-Others use a Domain Specific Language (DSL),
-such as Antimony,
-which is a less verbose version of the System Biology Markup Language (SBML).
-The last approach is to use a programming language such as Python,
-which is the approach used by PySB.
-
-Benefits of IDE integration (syntax highlighting, etc)
-
-Here we present SimBio, a Python package to define and simulate reaction networks.
+To address these two issues,
+we developed SimBio,
+a Python package to define and simulate reaction networks.
 From the end-user point of view,

### 2023-11-10 article.qmd
Improve the last paragraph of the introduction

@@
 From the power-user point of view,
-it can be extended from Python,
-since its coded in pure Python.
+since its coded in pure Python,
+it can be extended from Python.
+Its models can be introspected to create other representations
+such as graphs connecting species and/or reactions,
+or tables with parameters or equations.
 Finally,
 from the developer/maintainer point of view,
-it has a modular architecture with a clear separation of concerns,
+it has a modular architecture with a clear separation of concerns into three packages,
 making it easier to maintain or to contribute new code.
-
-the models can be easily introspected,
-and simulation extended from Python.
-Clean API to create tools (graphs, tables, etc)
 
@@
 
+We have written a set of Python packages
+that allow to define and simulate dynamical systems and reaction networks.
+Thay have a modular architecture that makes them easy to extend in different points.
+
+Poincaré and SimBio are focused on their integration with IDE
+
 Modular and extensible set of python libraries

### 2023-11-10 article.qmd
Convert code cells to executable cells
Fix code errors and add imports

@@
 
-```python
+```{python}
+from poincare import Variable, System, initial
+
 class Decay(System):
@@
 
-```python
+```{python}
+from poincare import Derivative
+
 class Oscillator(System):
@@
 
-```python
+```{python}
 class BigModel(System):
     x: Variable = initial(default=1)
-    subsystem = Oscillator(x=x)
-    independent = Oscillator(x=1)  # independent variable created inside
+    subsystem = Decay(x=x)
+    independent = Decay(x=1)  # independent variable created inside
 ```
@@
 
-```python
+```{python}
+from poincare import Constant, Parameter, assign
+
 class Model(System):
-    k: Constant = assign(default=1, constant=True)
+    c: Constant = assign(default=1, constant=True)
     # two variables sharing initial condition
@@
 
-```python
+```{python}
+import numpy as np
+from poincare import Simulator
+
 sim = Simulator(Oscillator)
-df = sim.solve(save_at=np.linspace(0, 5, 100))
+df = sim.solve(save_at=np.linspace(0, 5, 101))
 
@@
 
-```python
-df.plot()
+```{python}
+df.plot();
 ```
@@
 
-```python
-class Model(Copmartment):
+```{python}
+from simbio import Compartment, Species, Reaction, initial
+
+class Model(Compartment):
     """2A -> B"""
@@
     B: Species = initial(default=0)
-    r = Reaction(reactants=[2 * A], product=[B], rate=1)
+    r = Reaction(reactants=[2 * A], products=[B], rate_law=1)
 ```

### 2023-11-10 article.qmd
Add citations for symbolite backends

@@
 They can be inspected and compiled to different backends.
-Currently, we have implementations for NumPy (and Numba), SymPy and JAX,
+Currently, we have implementations for
+NumPy [@harrisArrayProgrammingNumPy2020],
+Numba [@lamNumbaLLVMbasedPython2015],
+SymPy [@meurerSymPySymbolicComputing2017],
+and JAX [@jax2018github],
 but it is easy to add new backends.

### 2023-11-10 article.qmd
Explain the symbolite backends

--- 2023-11-10 article.qmd
@@
 They can be inspected and compiled to different backends.
-Currently, we have implementations for NumPy (and Numba), SymPy and JAX,
+Currently, we have implementations for
+NumPy [@harrisArrayProgrammingNumPy2020];
+Numba [@lamNumbaLLVMbasedPython2015], a Just-in-Time (JIT) compiler to LLVM;
+SymPy [@meurerSymPySymbolicComputing2017], a library for symbolic mathematics;
+and JAX [@jax2018github], a library that support automatic differentation
+and compilation to GPUs and TPUs;
 but it is easy to add new backends.

### 2023-11-10 article.qmd
Cite dataclass_transform

@@
 it is important to add the type annotations,
-as it can provide a signature autocompletion in IDEs supporting the `dataclass_transform` (PEP 681).
+as it can provide a signature autocompletion in IDEs supporting the `dataclass_transform` [@debontePEP681Data2021]:
 

### 2023-11-10 article.qmd
Small changes

@@
 
-It allows to define higher-order systems
+It also allows to define higher-order systems
 by assigning an initial condition to a `Derivative`:
@@
     p: Parameter = assign(default=1)
-    z: Variable = initial(default=p)  # typing error
+    z: Variable = initial(default=p)  # type error
 ```

### 2023-11-10 article.qmd
Write discussion

@@
 
-We have written a set of Python packages
+In this article,
+we have described a set of Python packages we developed
 that allow to define and simulate dynamical systems and reaction networks.
-Thay have a modular architecture that makes them easy to extend in different points.
+They are focused on a deep integration with IDEs
+such that code analysis tools can flag errors before running the code,
+and help with refactoring and code completion.
+For this,
+we chose to use standard modern Python syntax,
+since the integration with IDEs will be maintained by the big players in the huge Python community.
 
-Poincaré and SimBio are focused on their integration with IDE
+In contrast to previous tools,
+both model definition and compilation into an ODE function is implemented in Python.
+This lowers the barrier to develop different simulation modalities,
+such as perfomance optimizations by leveraging some structure in the model.
+But being in Python does not mean less performance than C/C++ tools,
+as the resulting ODE function can be *jitted* using numba.
 
-Modular and extensible set of python libraries
-Leverage IDE integration and static typing for model creation
-
-Introduce 3 layers: symbolite, poincare and simbio
-easier to understand and develop
-usable, extendable, and maintainable by a broader community
-protocols
-change in the future
+These set of packages have a modular architecture with a clear separation of concerns.
+As such, they can be reused, improved and extended by the broader Python community.
+For example, people outside the systems biology community could
+add a stochastic integrator to poincaré,
+which could then be used from SimBio.
+This architecture also makes them easier to understand,
+reducing the difficulty to contribute fixes or new code,
+assuring their maintainance and continued development into the future.
 

### 2023-11-14 article.qmd
Add docx format

@@
     colorlinks: true
+  docx:
+    number-sections: true
 ---

### 2023-11-14 article.qmd
Add Hernan's changes

--- 2023-11-10 article.qmd
@@
 ---
-title: "SimBio: a flexible and extensible Python package for systems modelling"
+title: "SimBio: a versatile and expandable Python package for modeling systems."
 bibliography: bibliography.bib
@@
 Chemical Reaction Networks (CRN)
-are the basis of many models used across the biological sciences and epidemiology.
-They consist of a set of species and reactions that mediate conversions between them.
-Several modalities exist to simulate these systems:
-deterministic Ordinary Differential Equations (ODE) to model a macroscopic behavior,
-Stochastic Differential Equations (SDE) to model microscopic fluctuations,
-and jump processes (Gillespie-like simulations) to take into account the discreteness of the populations.
+are the foundation of many models used in biological sciences and epidemiology.
+They are made up of a set of species and reactions that mediate conversions between them.
+There are several ways to simulate these systems:
+deterministic Ordinary Differential Equations (ODEs) to model macroscopic behavior,
+Stochastic Differential Equations (SDEs) to model microscopic fluctuations,
+and jump processes (Gillespie-like simulations) to account for the discreteness of populations.
 Instead of writing the equations directly for each of these formulations
-which is error-prone and hard to reuse,
-these models can be written in a higher level description
-which can be compiled for these different types of simulations.
+which is error-prone and difficult to reuse,
+these models can be written in a higher-level description
+that can be compiled for these different types of simulations.
 
@@
 
-A limitation of these tools is their extensibility from Python,
-as they are or use wrapped libraries in other languages.
-As such, they allow to "control" the model definition and simulation in a Python script,
-but we cannot leverage other packages from the huge Python community
-to change the way the simulation is carried out.
-In some sense, we could say that
-they facilitate not having to export the simulation result to a CSV file
-and then loading it in Python to continue the analysis.
+One limitation of these tools is that they are not extensible from Python,
+as they are either built on or use wrapped libraries in other languages.
+This means that they allow you to "control" the model definition and simulation in a Python script,
+but you cannot use other packages from the vast Python community to change the way the simulation is carried out.
+In a way, we could say that
+they make it easier to avoid having to export the simulation results to a CSV file
+and then load them into Python for further analysis.
+TODO: GIVE EXAMPLES
 
-Another weak point is related to the way the models are written.
+Another weakness is related to the way models are written.
 While most tools support standards such as the System Biology Markup Language (SBML) [@huckaSBMLL3V2],
@@
 which can be translated to SBML.
-A DSL permits to reuse of the code *as is* in different programming environments,
-but will not be recognized in Integrated Development Enviromnents (IDE)
-and, then, cannot provide the development experience we have come to expect:
-syntax highlighting, code completion, refactoring, static analysis, among others.
+A DSL allows code to be reused in different programming environments,
+but it will not be recognized by Integrated Development Environments (IDEs)
+and therefore cannot provide the development experience we have come to expect, such as
+syntax highlighting, code completion, refactoring, and static analysis.
 This has been recognized by the community,
 and an extension for Visual Studio Code has been created as part of a master thesis [@vscode-antimony].
-But maintaining it into the future could be burdensome for the smaller systems biology community.
+However, maintaining it in the future could be burdensome for the smaller systems biology community.
 
-To address these two issues,
-we developed SimBio,
-a Python package to define and simulate reaction networks.
-From the end-user point of view,
-it is focused on providing an ergonomic experience
+We developed SimBio,
+a Python package for defining and simulating reaction networks,
+to address these two issues.
+It focuses on providing an ergonomic experience for end-users
 by integrating well with IDEs and static analysis tools
 through the use of standard modern Python syntax.
-From the power-user point of view,
-since its coded in pure Python,
-it can be extended from Python.
-Its models can be introspected to create other representations
+It can be extended from Python
+since it is coded in pure Python,
+which is beneficial for power users.
+Its models can be introspected to create other representations,
 such as graphs connecting species and/or reactions,
@@
 Finally,
-from the developer/maintainer point of view,
 it has a modular architecture with a clear separation of concerns into three packages,
-making it easier to maintain or to contribute new code.
+making it easier to maintain or to contribute new code,
+which is beneficial for developers and maintainers.
 
@@
 
-Modular code architecture enhances reusability, extendability and maintainability for the community.
-Hence, we split the code to define and simulate reaction systems into three Python packages:
+Modular code architecture makes code reusable, extensible, and easier to maintain.
+Therefore, we split the code to define and simulate reaction systems into three Python packages:
 symbolite, to create symbolic expressions;
 poincaré, to define dynamical systems;
-and simbio, to define reaction systems
-and interface with systems biology standards such as SBML.
+and simbio, to define reaction systems and interface with systems biology standards such as SBML.
+TODO: NOT SURE SUBSECTIONS ARE A GOOD IDEA. MAYBE JUST BOLD
 
@@
 
-In contrast to previous tools,
-both model definition and compilation into an ODE function is implemented in Python.
-This lowers the barrier to develop different simulation modalities,
-such as perfomance optimizations by leveraging some structure in the model.
-But being in Python does not mean less performance than C/C++ tools,
-as the resulting ODE function can be *jitted* using numba.
+Unlike previous tools,
+both model definition and compilation into an ODE function are implemented in Python.
+This lowers the barrier to developing different simulation modalities,
+such as performance optimizations by leveraging some structure in the model.
+However, being in Python does not mean that performance is less than that of C/C++ tools,
+as the resulting ODE function can be JITted using Numba.
 
-These set of packages have a modular architecture with a clear separation of concerns.
-As such, they can be reused, improved and extended by the broader Python community.
-For example, people outside the systems biology community could
-add a stochastic integrator to poincaré,
+These packages are designed with a modular architecture that clearly separates different concerns.
+This makes them easier to reuse, improve, and extend by the broader Python community.
+For example, someone outside the systems biology community could add a stochastic integrator to poincaré,
 which could then be used from SimBio.
 This architecture also makes them easier to understand,
-reducing the difficulty to contribute fixes or new code,
-assuring their maintainance and continued development into the future.
+which reduces the difficulty of contributing fixes or new code,
+and ensures their maintenance and continued development into the future.
 

### 2023-11-23 article.qmd
Add Sheriff suggestions

@@
 This has been recognized by the community,
-and an extension for Visual Studio Code has been created as part of a master thesis [@vscode-antimony].
+and an extension for Visual Studio Code has been created [@vscode-antimony].
 However, maintaining it in the future could be burdensome for the smaller systems biology community.
+TODO: this is missing for other tools
 
 We developed SimBio,
-a Python package for defining and simulating reaction networks,
+an open-source Python package for defining and simulating reaction networks,
 to address these two issues.
@@
 which is beneficial for power users.
+TODO: high degree of flexibility,  large potential to leverage existing python tools.  For example, ....
 Its models can be introspected to create other representations,

### 2023-11-27 article.qmd
Update bibliography for vscode-antimony
Change title and introduction to include poincaré

@@
 ---
-title: "SimBio: a versatile and expandable Python package for modeling systems."
+title: "Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems."
 bibliography: bibliography.bib
@@
 This has been recognized by the community,
-and an extension for Visual Studio Code has been created [@vscode-antimony].
+and an extension for Visual Studio Code has been created [@maTestingDynamicalSystems2021].
 However, maintaining it in the future could be burdensome for the smaller systems biology community.
@@
 
-We developed SimBio,
-an open-source Python package for defining and simulating reaction networks,
-to address these two issues.
-It focuses on providing an ergonomic experience for end-users
+To address these issues,
+we developed poincaré and SimBio,
+open-source Python packages for defining and simulating systems.
+Poincaré allows to define differential equation systems
+and SimBio builds on top of it to define reaction networks.
+They are focused on providing an ergonomic experience for end-users
 by integrating well with IDEs and static analysis tools
 through the use of standard modern Python syntax.
-It can be extended from Python
-since it is coded in pure Python,
+They can be extended from Python
+since they are coded in pure Python,
 which is beneficial for power users.
 TODO: high degree of flexibility,  large potential to leverage existing python tools.  For example, ....
-Its models can be introspected to create other representations,
+Their models can be introspected to create other representations,
 such as graphs connecting species and/or reactions,
@@
 Finally,
-it has a modular architecture with a clear separation of concerns into three packages,
+they have a modular architecture with a clear separation of concerns,
 making it easier to maintain or to contribute new code,

### 2023-11-27 article.qmd
Remove subsection
Add example of limitation of just controlling from Python
Add CSS to justify

@@
     anchor-sections: true
+    css: article.css
   pdf:
@@
 but you cannot use other packages from the vast Python community to change the way the simulation is carried out.
+For instance,
+to do a Random Parameter Scan in COPASI,
+we must choose between one of the predefined distributions,
+but cannot use one of the meny distributions from `scipy.stats`.
 In a way, we could say that
@@
 and then load them into Python for further analysis.
-TODO: GIVE EXAMPLES
 
@@
 and simbio, to define reaction systems and interface with systems biology standards such as SBML.
-TODO: NOT SURE SUBSECTIONS ARE A GOOD IDEA. MAYBE JUST BOLD
-
-## Symbolite
 

### 2023-11-27 article.qmd
Add performance figure

@@
 which is beneficial for power users.
-TODO: high degree of flexibility,  large potential to leverage existing python tools.  For example, ....
+For instance, it is easy to reuse integrators defined in other packages,
+but it is also possible to change the compilation itself
+(the )
 Their models can be introspected to create other representations,
@@
 
-## Other things
+## Reproducibility and performance
 
-units -> Pint
-interactive
+![Performance of the different softwares to solve a subset of BioModels models.
+(left) Run time for model BIOMD3 as a function of the number of output points.
+(right) Run time for different models ordered according to the geometric mean of the run times.](src/performance/figures/performance.png){#fig-runtime}
 

### 2023-11-27 article.qmd
Move code blocks to a figure and add mathematical expression

@@
 
-Poincare is a package to define and simulate dynamical systems.
-It provides a `System` class,
-where we can define `Constant`s, `Parameter`s, `Variable`s,
-and create equations linking a variable's derivative with an expression:
+```{python}
+# | echo: false
+import numpy as np
+import seaborn
+from poincare import Derivative, Variable, System, initial
+from poincare import Simulator
+
+def plot(model, figheight):
+    ax = (
+        Simulator(model)
+        .solve(save_at=np.linspace(0, 5, 101))
+        .plot(figsize=(2, figheight))
+    )
+    ax.legend(frameon=False)
+    seaborn.despine(ax.figure)
+
+```
+
+:::: {#fig-code layout-nrow=3 .column-body-outset}
+
+::: {#fig-first-order layout-ncol=3}
+```{python}
+class Decay(System):
+    x: Variable = initial(default=1)
+    eq = x.derive() << -x
+
+```
+
+$$
+\begin{align}
+    \dot{x} &= -x
+    \\\\
+    x(0) &= 1
+\end{align}
+$$
 
 ```{python}
-from poincare import Variable, System, initial
+# | echo: false
+plot(Decay, 1)
 
-class Decay(System):
-    x: Variable = initial(default=1)  # Variable with initial condition
-    eq = x.derive() << -x  # equation assigning dx/dt = -x
 ```
 
-It also allows to define higher-order systems
-by assigning an initial condition to a `Derivative`:
+First-order system of an exponential decay.
+:::
 
+::: {#fig-second-order layout-ncol=3}
 ```{python}
-from poincare import Derivative
-
 class Oscillator(System):
@@
     eq = v.derive() << -x
+
 ```
+
+$$
+\begin{align}
+    \ddot{x} = -x
+    \\\\
+    \begin{cases}
+        x(0) &= 1 \\
+        \dot{x}(0) &= 0
+    \end{cases}
+\end{align}
+$$
+
+```{python}
+# | echo: false
+plot(Oscillator, 1.5)
+
+```
+
+Second-order system of an harmonic oscillator.
+:::
+
+::: {#fig-composition layout-ncol=3}
+```{python}
+class BigModel(System):
+    x: Variable = initial(default=1)
+    subsystem = Decay(x=x)
+    independent = Decay(x=2)
+
+```
+
+$$
+\begin{align}
+    \begin{cases}
+        \dot{x} = -x \\
+        \dot{y} = -y
+    \end{cases}
+    \\\\
+    \begin{cases}
+        x(0) &= 1 \\
+        y(0) &= 2
+    \end{cases}
+\end{align}
+$$
+
+```{python}
+# | echo: false
+plot(BigModel, 1.5)
+
+```
+
+First-order system of two exponential decays.
+:::
+::::
+
+Poincare is a package to define and simulate dynamical systems.
+It provides a `System` class,
+where we can define `Constant`s, `Parameter`s, `Variable`s,
+and create equations linking a variable's derivative with an expression (@fig-first-order).
+It also allows to define higher-order systems
+by assigning an initial condition to a `Derivative` (@fig-second-order).
 
@@
 3. it allows IDEs to provide autocomplete and refactoring capabilities (`Oscillator.<TAB>` shows `x`, `v` and `eq`),
-4. it allows to create instances which can be composed into a bigger model.
+4. it allows to create instances which can be composed into a bigger model (@fig-composition).
 
@@
 as it can provide a signature autocompletion in IDEs supporting the `dataclass_transform` [@debontePEP681Data2021]:
-
-```{python}
-class BigModel(System):
-    x: Variable = initial(default=1)
-    subsystem = Decay(x=x)
-    independent = Decay(x=1)  # independent variable created inside
-```
 

### 2023-11-27 article.qmd
Remove plots in figure 1

@@
 # | echo: false
-import numpy as np
-import seaborn
-from poincare import Derivative, Variable, System, initial
-from poincare import Simulator
-
-def plot(model, figheight):
-    ax = (
-        Simulator(model)
-        .solve(save_at=np.linspace(0, 5, 101))
-        .plot(figsize=(2, figheight))
-    )
-    ax.legend(frameon=False)
-    seaborn.despine(ax.figure)
+from poincare import Derivative, System, Variable, initial
 
@@
 
-:::: {#fig-code layout-nrow=3 .column-body-outset}
+:::: {#fig-code layout-nrow=3}
 
@@
 $$
-\begin{align}
-    \dot{x} &= -x
-    \\\\
-    x(0) &= 1
-\end{align}
+\dot{x} = -x
 $$
 
-```{python}
-# | echo: false
-plot(Decay, 1)
-
-```
+$$
+x(0) = 1
+$$
 
@@
 $$
-\begin{align}
 \ddot{x} = -x
-    \\\\
+$$
+
+$$
 \begin{cases}
@@
 \end{cases}
-\end{align}
 $$
-
-```{python}
-# | echo: false
-plot(Oscillator, 1.5)
-
-```
 
@@
 $$
-\begin{align}
 \begin{cases}
@@
 \end{cases}
-    \\\\
+$$
+
+$$
 \begin{cases}
@@
 \end{cases}
-\end{align}
 $$
-
-```{python}
-# | echo: false
-plot(BigModel, 1.5)
-
-```
 
@@
 :::
+
+Code and corresponding mathematical expressions for different systems.
 ::::

### 2023-11-27 article.qmd
Simulation as figure

@@
 To simulate a system,
-we create a `Simulator` instance,
+we create a `Simulator` instance (@fig-sim),
 which compiles a given system and interfaces with solvers.
@@
 
+::: {#fig-sim layout=[[1],[1,1]]}
 ```{python}
@@
 df = sim.solve(save_at=np.linspace(0, 5, 101))
-
-df
 ```
@@
 ```{python}
-df.plot();
+df.head()
 ```
+
+```{python}
+df.plot(figsize=(4, 2))
+```
+
+Simulation of the `Oscillator` system from @fig-second-order,
+which outputs a `pandas.DataFrame`.
+:::
 

### 2023-11-27 article.qmd
Constant and Parameter figure

@@
 As such, they cannot be used to parametrize the initial conditions of variables,
-for which we have to use a `Constant`:
+for which we have to use a `Constant` (@fig-constant).
+
+::: {#fig-constant layout-nrow=2}
+```{python}
+# | echo: false
+from poincare import Constant, Parameter, assign
+
+```
 
 ```{python}
-from poincare import Constant, Parameter, assign
-
 class Model(System):
     c: Constant = assign(default=1, constant=True)
-    # two variables sharing initial condition
     x: Variable = initial(default=c)
@@
 
+```
+
+```{python}
+class BadModel(System):
     p: Parameter = assign(default=1)
     z: Variable = initial(default=p)  # type error
+
 ```
+
+Constant and Parameter allow to connect initial conditions and equations, respectively.
+Using Parameters as initial conditions raises an error,
+and can be flagged by static type checkers.
+:::
 

### 2023-11-27 article.qmd
Format Compartment code
Expand legend for simulation figure

@@
 
-Simulation of the `Oscillator` system from @fig-second-order,
-which outputs a `pandas.DataFrame`.
+Simulation of the `Oscillator` system from @fig-second-order.
+The output is a `pandas.DataFrame` with a column for each variable
+and the time as index.
+It is inspected and plotted with the `pandas` methods.
 :::
@@
     """2A -> B"""
+
     A: Species = initial(default=1)
@@
     r = Reaction(reactants=[2 * A], products=[B], rate_law=1)
+
 ```

### 2023-11-27 article.qmd
Add imports as part of poincare code figure
SimBio figure

@@
 
+:::: {#fig-poincare layout-nrow=4}
+
 ```{python}
-# | echo: false
 from poincare import Derivative, System, Variable, initial
@@
 ```
-
-:::: {#fig-code layout-nrow=3}
 
@@
 
-For reaction networks,
-we are usually interested in first-order differential equations
-relating the rate of change of the network's species.
-
-To facilitate defining these models,
-SimBio defines `Species`,
-which comprises of a `poincare.Variable` and an stoichiometry,
-and `Reaction`s,
-which transform reactants into products taking into account the stoichiometry.
+:::: {#fig-simbio layout-nrow=3}
 
 ```{python}
-from simbio import Compartment, Species, Reaction, initial
+from simbio import Compartment, MassAction, Species, Reaction, initial
 
+```
+
+::: {#fig-reaction layout-ncol=2}
+```{python}
 class Model(Compartment):
@@
     B: Species = initial(default=0)
-    r = Reaction(reactants=[2 * A], products=[B], rate_law=1)
+    r = Reaction(
+        reactants=[2 * A],
+        products=[B],
+        rate_law=1,
+    )
 
@@
 
-It also includes a `MassAction` variant,
-which automatically adds the reactants as part of the rate.
-`MassAction(reactants=[2 * A], product=[B], rate=1)`
-is equivalent to
-`Reaction(reactants=[2 * A], product=[B], rate=1 * A**2)`,
-and many predefined commonly used reactions reactions.
+$$
+\begin{align}
+&\begin{cases}
+    \dot{A} = -2 \\
+    \dot{B} = +1
+\end{cases}
+\\\\
+&\begin{cases}
+    A(0) &= 1 \\
+    B(0) &= 0
+\end{cases}
+\end{align}
+$$
 
-SimBio also supports importing from and exporting to SBML,
-or directly from BioModels:
+Reaction system.
+:::
 
-```python
-from simbio import Simulator
+::: {#fig-mass-action layout-ncol=2}
+```{python}
+class Model(Compartment):
+    """2A -> B"""
+
+    A: Species = initial(default=1)
+    B: Species = initial(default=0)
+    r = MassAction(
+        reactants=[2 * A],
+        products=[B],
+        rate=1,
+    )
+
+```
+
+$$
+\begin{align}
+&\begin{cases}
+    \dot{A} = -2 A^2 \\
+    \dot{B} = +1 A^2
+\end{cases}
+\\\\
+&\begin{cases}
+    A(0) &= 1 \\
+    B(0) &= 0
+\end{cases}
+\end{align}
+$$
+
+Mass-Action system.
+:::
+
+::::
+
+For reaction networks,
+we are interested in first-order differential equations
+relating the rate of change of the species.
+To facilitate defining these models,
+SimBio defines `Species`,
+which comprises of a `poincare.Variable` and an stoichiometry,
+and `Reaction`,
+which transform reactants into products taking into account the stoichiometry (@fig-reaction).
+It also includes `MassAction`, a `Reaction` subclass,
+which automatically adds the reactants as part of the rate law (@fig-mass-action).
+Several commonly used reactions are predefined as `MassAction` subclasses.
+Additionally, SimBio supports importing from and exporting to SBML,
+or directly from BioModels (@fig-simbio-io).
+
+::: {#fig-simbio-io layout-ncol=1}
+```{python}
 from simbio.io import biomodels, sbml
@@
 model = biomodels.load_model("BIOMD12")  # from BioModels
-Simulator(model).solve(...)
 ```
+
+Creation of a model from a local SBML file or one uploaded to BioModels.
+:::
 

### 2023-11-27 article.qmd
Add text for the reproducibility and runtime sections

@@
 (left) Run time for model BIOMD3 as a function of the number of output points.
-(right) Run time for different models ordered according to the geometric mean of the run times.](src/performance/figures/performance.png){#fig-runtime}
+(right) Run time for different models ordered according to the geometric mean of the run times, using 300 output points.](src/performance/figures/performance.png){#fig-runtime}
+
+As poincaré and SimBio are coded in pure Python,
+an expected concern is its runtime performance when integrating a model.
+To test it,
+we downloaded 28 models from BioModels,
+and simulated them with COPASI, Tellurium/RoadRunner and simbio.
+For simbio, we considered both the pure Python and numba-compiled code.
+We used the LSODA solver with $10^-6$ relative and absolute tolerances,
+with the exception of tellurium,
+which provides the similar CVODE solver.
+We measured separately the loading, first/cold run and subsequent/warm runs of each model.
+
+For COPASI and Tellurium/RoadRunner,
+we noted that its runtime depended on the number of evaluation points,
+something that does not happen with SimBio (@fig-runtime, left).
+Using numba to compile the right-hand-side (RHS) code generated by SimBio
+results in an order of magnitude speed-up,
+putting its performance on par with COPASI and RoadRunner.
+If we switch the LSODA `scipy` solver for a more efficient `numbalsoda` implementation,
+we get another speed-up in the SimBio runtime,
+beating all other methods.
+This is also true for the other models we tested (@fig-runtime, right).
 

### 2023-11-28 article.qmd
Add variable names in caption for poincare
Change dot to explicit derivatives and expand caption for SimBio example

@@
 
-Second-order system of an harmonic oscillator.
+Second-order system of an harmonic oscillator,
+where $\dot{x}$ is named $v$.
 :::
@@
 
-First-order system of two exponential decays.
+First-order system of two exponential decays,
+where $y$ is named $\text{independent.x}$.
 :::
@@
 &\begin{cases}
-    \dot{A} = -2 \\
-    \dot{B} = +1
+    \frac{dA}{dt} = -2 \\
+    \frac{dB}{dt} = +1
 \end{cases}
@@
 
-Reaction system.
+Reaction system $2A \rightarrow B$ with rate $1$.
 :::
@@
 &\begin{cases}
-    \dot{A} = -2 A^2 \\
-    \dot{B} = +1 A^2
+    \frac{dA}{dt} = -2 A^2 \\
+    \frac{dB}{dt} = +1 A^2
 \end{cases}
@@
 
-Mass-Action system.
+Mass-Action system $2A \rightarrow B$ with rate $1 A^2$.
 :::
 
+Examples of similar reaction systems transforming $2A \rightarrow B$.
+For mass-action systems, the rate is proportional to the reactants.
 ::::

### 2023-11-28 article.qmd
Expand reproducibility and performance

--- 2023-11-27 article.qmd
@@
 (left) Run time for model BIOMD3 as a function of the number of output points.
-(right) Run time for different models ordered according to the geometric mean of the run times.](src/performance/figures/performance.png){#fig-runtime}
+(right) Run time for different models ordered according to the geometric mean of the run times, using 300 output points.](src/performance/figures/performance.png){#fig-runtime}
+
+To test the reproducibility of results of SimBio,
+we considered the first 100 curated models from BioModels.
+As SimBio does not yet support the full SBML standard,
+for instance, events,
+we simulated only 60 of those.
+Using the results from COPASI as ground truth,
+SimBio reproduces the results within the solver tolerances.
+
+To test the performance,
+we simulated them with COPASI, Tellurium/RoadRunner and simbio.
+For simbio, we considered both the NumPy and numba backends.
+For COPASI and simbio, we used the LSODA solver,
+while using the similar CVODE solver for tellurium.
+In all cases, we used relative and absolute tolerances of  $10^-6$.
+We measured separately three stages of the simulation:
+the loading, first/cold run and subsequent/warm runs of each model.
+
+For COPASI and Tellurium/RoadRunner,
+we noted that its runtime depended on the number of evaluation points,
+something that does not seem to happen with SimBio (@fig-runtime, left).
+While SimBio's NumPy backend is slower than both COPASI and RoadRunner,
+we obtained an order of magnitude speed-up using the numba backend putting it on par with them.
+A user might have to consider the trade-off between compilation and run times,
+as the compilation of the right-hand-side (RHS) code might take longer than the runtime itself,
+and not be worth it for running the model only once.
+Another speed-up in the runtime can be had
+by switching the LSODA `scipy` solver for a more efficient `numbalsoda` implementation,
+which avoids the Python interpreter between each of the integration steps.
+This last combination beats all other methods,
+which is also true for the other models we tested (@fig-runtime, right).
 

### 2023-11-28 article.qmd
Rewriting the introduction

@@
 One limitation of these tools is that they are not extensible from Python,
-as they are either built on or use wrapped libraries in other languages.
-This means that they allow you to "control" the model definition and simulation in a Python script,
-but you cannot use other packages from the vast Python community to change the way the simulation is carried out.
+as they are not fully Python packages
+but wrap libraries in other languages to do the simulation.
+As such, they allow you to "control" the model definition and simulation in a Python script.
+But it is not possible to use other packages from the vast Python community to change the way the simulation is carried out.
 For instance,
 to do a Random Parameter Scan in COPASI,
-we must choose between one of the predefined distributions,
-but cannot use one of the meny distributions from `scipy.stats`.
-In a way, we could say that
-they make it easier to avoid having to export the simulation results to a CSV file
-and then load them into Python for further analysis.
+we must choose between one of 4 predefined distributions,
+but cannot use one of the many distributions from `scipy.stats`.
 
@@
 However, maintaining it in the future could be burdensome for the smaller systems biology community.
-TODO: this is missing for other tools
 

### 2023-11-28 article.qmd
Fix typos

@@
 SymPy [@meurerSymPySymbolicComputing2017], a library for symbolic mathematics;
-and JAX [@jax2018github], a library that support automatic differentation
+and JAX [@jax2018github], a library that support automatic differentiation
 and compilation to GPUs and TPUs;
@@
 Switching backends to `"numba"` can improve model runtime
-on a factor upto x30 or more for big or long running models,
+on a factor up to x30 or more for big or long running models,
 but incurs in a compile time penalty for the first run

### 2023-11-29 article.qmd
Add writing models with PySB

@@
 However, maintaining it in the future could be burdensome for the smaller systems biology community.
+In the case of PySB,
+it leveraged the highly dynamic nature of Python to create a DSL within it.
+By default, it uses global state to create species and parameters,
+without assigning them to Python variables
+or adding them explicitly to the model.
+But IDEs cannot understand it and, hence, provide help, as this is not the way they expect Python code to be written.
 

### 2023-11-29 article.qmd
Introduction

@@
 but cannot use one of the many distributions from `scipy.stats`.
+In Tellurium, we cannot choose other Python solvers,
+as adding new integrators has to be done in C++.
 
@@
 A DSL allows code to be reused in different programming environments,
-but it will not be recognized by Integrated Development Environments (IDEs)
+but it will not be recognized by default in Integrated Development Environments (IDEs)
 and therefore cannot provide the development experience we have come to expect, such as
@@
 open-source Python packages for defining and simulating systems.
-Poincaré allows to define differential equation systems
+Poincaré allows to define systems of differential equations
 and SimBio builds on top of it to define reaction networks.
@@
 through the use of standard modern Python syntax.
-They can be extended from Python
-since they are coded in pure Python,
-which is beneficial for power users.
-For instance, it is easy to reuse integrators defined in other packages,
-but it is also possible to change the compilation itself
-(the )
+Since they are coded in pure Python,
+every part from model definition to simulation can be extended from Python itself.
+For example,
+easier changes such as reusing integrators defined in other packages,
+to more complex ones such as changing the compilation process itself
+to leverage some structure in the equations.
 Their models can be introspected to create other representations,

### 2023-11-29 article.qmd
Extend figure captions

@@
 and compilation to GPUs and TPUs;
-but it is easy to add new backends.
+and it is easy to add new backends.
 
@@
 First-order system of an exponential decay.
+The variable `x` stores the initial condition for $x$,
+and the variable `eq` stores the rate equation for $x$.
 :::
@@
 Second-order system of an harmonic oscillator,
-where $\dot{x}$ is named $v$.
+where the variable `v` stores the derivative $\dot{x}$
+and the rate equation is specified for the derivative `v`.
 :::
@@
     x: Variable = initial(default=1)
-    subsystem = Decay(x=x)
+    linked = Decay(x=x)
     independent = Decay(x=2)
@@
 
-First-order system of two exponential decays,
-where $y$ is named $\text{independent.x}$.
+First-order system of two exponential decays by composition of the `Decay` system.
+The subsystem `linked` has a reference to the outer variable `x`,
+while the subsystem `independent` defines a new variable with initial condition §2$,
+which on the corresponding mathematical expression was named $y$.
 :::
@@
 
-::: {#fig-reaction layout-ncol=2}
+::: {layout-ncol=2}
 ```{python}
@@
 $$
-
-Reaction system $2A \rightarrow B$ with rate $1$.
 :::
 
-::: {#fig-mass-action layout-ncol=2}
+::: {layout-ncol=2}
 ```{python}
@@
 $$
-
-Mass-Action system $2A \rightarrow B$ with rate $1 A^2$.
 :::
 
-Examples of similar reaction systems transforming $2A \rightarrow B$.
-For mass-action systems, the rate is proportional to the reactants.
+A reaction system for species $A$ and $B$ with initial conditions $1$ and $0$, respectively.
+A single reaction transforming $2A$ into $B$ is saved in variable `r`.
+The rate $1$ is specified directly for `Reaction`,
+and is proportional to the reactants for `MassAction`.
 ::::
@@
 and `Reaction`,
-which transform reactants into products taking into account the stoichiometry (@fig-reaction).
+which transform reactants into products taking into account the stoichiometry (@fig-simbio).
 It also includes `MassAction`, a `Reaction` subclass,
-which automatically adds the reactants as part of the rate law (@fig-mass-action).
+which automatically adds the reactants as part of the rate law (@fig-simbio).
 Several commonly used reactions are predefined as `MassAction` subclasses.

### 2023-11-29 article.qmd
Add screenshots

@@
 
-::: {#fig-constant layout-nrow=2}
-```{python}
-# | echo: false
-from poincare import Constant, Parameter, assign
+::: {#fig-constant layout-nrow=1}
 
-```
+::: {#fig-parameter layout-nrow=2}
+![](src/ide/constant.png)
 
-```{python}
-class Model(System):
-    c: Constant = assign(default=1, constant=True)
-    x: Variable = initial(default=c)
-    y: Variable = initial(default=c)
+![](src/ide/type_error.png)
 
-```
+`Constant` are assigned with `assign(..., constant=True)` and allow to link `Variable`s initial conditions,
+while trying to use a `Parameter` instead is flagged as type errors (red underlining).
+:::
 
-```{python}
-class BadModel(System):
-    p: Parameter = assign(default=1)
-    z: Variable = initial(default=p)  # type error
+::: {#fig-composition layout-nrow=1}
+![](src/ide/composition_tooltip.png)
 
-```
+A tooltip is shown when composing models,
+which show the expected variables and their default values.
+The IDE highlights mismatched types (`my_var` is `Variable` and `c` must be a number or a `Constant`)
+and wrong names (`z` is not a name in `Model`).
+:::
 
-Constant and Parameter allow to connect initial conditions and equations, respectively.
-Using Parameters as initial conditions raises an error,
-and can be flagged by static type checkers.
+Screenshots of Visual Studio Code showing tooltips and highlighted type errors.
 :::

### 2023-11-29 src/ide/composition_tooltip.py
IDE screenshots code

@@ -1,0 +1,15 @@
+from poincare import System, Variable, initial  # noqa: I001
+
+from .constant import Model
+
+
+
+
+
+
+
+class BigModel(System):
+    my_var: Variable = initial(default=1)
+    subsystem = Model()
+    wrong_type = Model(c=my_var)
+    wrong_variable_name = Model(z=my_var)

### 2023-11-29 src/ide/constant.py

@@ -1,0 +1,7 @@
+from poincare import Constant, System, Variable, assign, initial
+
+
+class Model(System):
+    c: Constant = assign(default=1, constant=True)
+    x: Variable = initial(default=c)
+    y: Variable = initial(default=c)

### 2023-11-29 src/ide/type_error.py

@@ -1,0 +1,8 @@
+from poincare import Parameter, System, Variable, assign, initial
+
+
+class BadModel(System):
+    p: Parameter = assign(default=1)
+    z: Variable = initial(default=p)
+    eq = z.derive() << p * z
+

### 2023-11-30 article.qmd
Fix exponent
Change dot to frac derivatives

@@
 $$
-\dot{x} = -x
+\frac{dx}{dt} = -x
 $$
@@
 $$
-\ddot{x} = -x
+\frac{d^2x}{dt^2} = -x
 $$
@@
     x(0) &= 1 \\
-    \dot{x}(0) &= 0
+    \frac{dx}{dt}(0) &= 0
 \end{cases}
@@
 Second-order system of an harmonic oscillator,
-where the variable `v` stores the derivative $\dot{x}$
+where the variable `v` stores the derivative $\frac{dx}{dt}$
 and the rate equation is specified for the derivative `v`.
@@
 \begin{cases}
-    \dot{x} = -x \\
-    \dot{y} = -y
+    \frac{dx}{dt} = -x \\
+    \frac{dy}{dt} = -y
 \end{cases}
@@
 while using the similar CVODE solver for tellurium.
-In all cases, we used relative and absolute tolerances of  $10^-6$.
+In all cases, we used relative and absolute tolerances of $10^{-6}$.
 We measured separately three stages of the simulation:

### 2023-11-30 article.qmd
Clarify download from BioModels
Implemented reactions

@@
 which automatically adds the reactants as part of the rate law (@fig-simbio).
-Several commonly used reactions are predefined as `MassAction` subclasses.
+Several commonly used reactions are predefined as `MassAction` subclasses,
+such as
+`MichaelisMenten` ($S + E \rightarrow ES -> P + E$)
+and its approximate form without the intermediate species $ES$,
+and it is also simple to implement your own as subclasses of `Reaction` or `MassAction`.
 Additionally, SimBio supports importing from and exporting to SBML,
-or directly from BioModels (@fig-simbio-io).
+and downloading them directly from BioModels (@fig-simbio-io).
 

### 2023-11-30 article.qmd
Show model repr from BioModels figure
Align equations in figures

@@
 
-::: {#fig-first-order layout-ncol=3}
+::: {#fig-first-order layout-ncol=3 layout-valign="center"}
 ```{python}
@@
 
-::: {#fig-second-order layout-ncol=3}
+::: {#fig-second-order layout-ncol=3 layout-valign="center"}
 ```{python}
@@
 
-::: {#fig-composition layout-ncol=3}
+::: {#fig-composition layout-ncol=3 layout-valign="center"}
 ```{python}
@@
 
-::: {#fig-parameter layout-nrow=2}
+::: {#fig-parameter layout-nrow=2 layout-valign="bottom"}
 ![](src/ide/constant.png)
@@
 
-::: {#fig-composition layout-nrow=1}
+::: {#fig-composition layout-nrow=1 layout-valign="bottom"}
 ![](src/ide/composition_tooltip.png)
@@
 
-::: {layout-ncol=2}
+::: {layout-ncol=2 layout-valign="center"}
 ```{python}
@@
 model = biomodels.load_model("BIOMD12")  # from BioModels
+model
 ```

### 2023-12-01 src/ide/autocomplete.py
Add autocomplete example

@@ -1,0 +1,10 @@
+from poincare import Constant, Parameter, System, Variable, assign, initial
+
+
+class Model(System):
+    A: Variable = initial(default=0)
+    b: Constant = assign(default=1, constant=True)
+    c: Parameter = assign(default=2)
+
+
+Model.

### 2023-12-01 article.qmd
Improve IDE figures

@@
 
-::: {#fig-constant layout-nrow=1}
+::: {#fig-constant layout-nrow=1 layout-valign="bottom"}
 
-::: {#fig-parameter layout-nrow=2 layout-valign="bottom"}
-![](src/ide/constant.png)
+::: {#fig-parameter}
+![](src/ide/ide1.png)
 
-![](src/ide/type_error.png)
-
-`Constant` are assigned with `assign(..., constant=True)` and allow to link `Variable`s initial conditions,
+`Constant` are assigned with `assign(..., constant=True)` and can be used to link `Variable`s initial conditions,
 while trying to use a `Parameter` instead is flagged as type errors (red underlining).
+The IDE automatically recognizes `e` as an `Equation`,
+and provides autocompletion of the variables.
 :::
@@
 ::: {#fig-composition layout-nrow=1 layout-valign="bottom"}
-![](src/ide/composition_tooltip.png)
+![](src/ide/ide2.png)
 
@@
 which show the expected variables and their default values.
-The IDE highlights mismatched types (`my_var` is `Variable` and `c` must be a number or a `Constant`)
-and wrong names (`z` is not a name in `Model`).
+The IDE highlights
+wrong names (`z` is not a name in `Model`)
+and
+mismatched types (`x` is `Variable` and `a` must be a number or a `Constant`)
 :::

### 2023-12-01 src/ide/autocomplete.py

@@ -2,9 +2,10 @@ from poincare import Constant, Parameter, System, Variable, assign, initial
 
 
 class Model(System):
-    A: Variable = initial(default=0)
-    b: Constant = assign(default=1, constant=True)
+    a: Constant = assign(default=1, constant=True)
+    b: Variable = initial(default=a)
     c: Parameter = assign(default=2)
-
+    d: Variable = initial(default=c)
+    e = d.derive() << a + c * d
 
 Model.

### 2023-12-01 src/ide/composition_tooltip.py

@@ -1,6 +1,8 @@
 from poincare import System, Variable, initial  # noqa: I001
 
-from .constant import Model
+from .autocomplete import Model
+
+
 
 
 
@@ -9,7 +11,6 @@ from .constant import Model
 
 
 class BigModel(System):
-    my_var: Variable = initial(default=1)
-    subsystem = Model()
-    wrong_type = Model(c=my_var)
-    wrong_variable_name = Model(z=my_var)
+    x: Variable = initial(default=1)
+    wrong_name = Model(z=x)
+    wrong_type = Model(a=x)

### 2023-12-01 article.qmd
Improve the legend of the performance

@@
 
-![Performance of the different softwares to solve a subset of BioModels models.
-(left) Run time for model BIOMD3 as a function of the number of output points.
-(right) Run time for different models ordered according to the geometric mean of the run times, using 300 output points.](src/performance/figures/performance.png){#fig-runtime}
+![Performance of different softwares to solve models from the cureted section of BioModels.
+(left) Run time for the model BIOMD3 as a function of the number of output points.
+(right) Run time for different models for 300 output points, using the geometric mean of the different softwares to order them.](src/performance/figures/performance.png){#fig-runtime}
 

### 2023-12-01 article.qmd
Remove pdf and docx outputs

@@
     css: article.css
-  pdf:
-    documentclass: scrartcl
-    papersize: a4
-    geometry:
-        - margin=20mm
-    pdf-engine: tectonic
-    number-sections: true
-    colorlinks: true
-  docx:
-    number-sections: true
 ---

### 2023-12-01 article.qmd
Change phrasing regading solver tolerance

@@
 we considered the first 100 curated models from BioModels.
-As SimBio does not yet support the full SBML standard,
-for instance, events,
-we simulated only 60 of those.
+As SimBio does not yet support the full SBML standard (for instance, events),
+we could only simulate 60 of those.
 Using the results from COPASI as ground truth,
-SimBio reproduces the results within the solver tolerances.
+SimBio reproduces the results with numerical differences depending on the solver tolerance.
 

### 2023-12-01 article.qmd
Join captions in IDE figure

@@
 As such, they cannot be used to parametrize the initial conditions of variables,
-for which we have to use a `Constant` (@fig-constant).
+for which we have to use a `Constant` (@fig-ide).
 
-::: {#fig-constant layout-nrow=1 layout-valign="bottom"}
-
-::: {#fig-parameter}
+::: {#fig-ide layout-ncol=2 layout-valign="bottom"}
 ![](src/ide/ide1.png)
 
+![](src/ide/ide2.png)
+
+Screenshots of Visual Studio Code showing tooltips and highlighted type errors.
 `Constant` are assigned with `assign(..., constant=True)` and can be used to link `Variable`s initial conditions,
@@
 and provides autocompletion of the variables.
-:::
-
-::: {#fig-composition layout-nrow=1 layout-valign="bottom"}
-![](src/ide/ide2.png)
-
 A tooltip is shown when composing models,
@@
 mismatched types (`x` is `Variable` and `a` must be a number or a `Constant`)
-:::
-
-Screenshots of Visual Studio Code showing tooltips and highlighted type errors.
 :::

### 2023-12-13 article.qmd
Hernan's suggested changes
- Remove plot output
- Paraphrase line
Sheriff's changes

@@
 
+```{python}
+# | echo: false
+import matplotlib
+
+matplotlib.rc("figure", figsize=(4, 2))
+```
+
 # Introduction
 
-Chemical Reaction Networks (CRN)
-are the foundation of many models used in biological sciences and epidemiology.
-They are made up of a set of species and reactions that mediate conversions between them.
-There are several ways to simulate these systems:
+Chemical Reaction Networks (CRNs)
+are a fundamental concept of modelling in numerous fields including systems biology, biochemistry, chemical engineering and epidemeology.
+They are comprised of a set of chemical species or biological entities and represntation of complex interaction and transformation between them through a series reactions.
+These systems can be modelled and simulated through multiple approaches:
 deterministic Ordinary Differential Equations (ODEs) to model macroscopic behavior,
@@
 and jump processes (Gillespie-like simulations) to account for the discreteness of populations.
-Instead of writing the equations directly for each of these formulations
+Instead of directly writing the equations for each of these formulations
 which is error-prone and difficult to reuse,
@@
 
-Several tools already exist to define and simulate these systems.
-In particular,
+Several tools already exist for defining and simulating CRNs.
 BioSimulators.org [@shaikhBioSimulatorsCentralRegistry2022],
 a registry of simulation tools,
-list at least 15 Python softwares.
-Some of them are:
-
-1. COPASI [@hoopsCOPASICOmplexPAthway2006],
-a stand-alone software
-which provides a Graphical User Interface (GUI) to define and simulate models.
-2. Tellurium [@choiTelluriumExtensiblePythonbased2018],
-which defines itself as a Python-based modeling environment
-and uses a C++ library called libRoadRunner [@welshLibRoadRunnerHighPerformance2023] to simulate them.
-3. PySB [@lopezProgrammingBiologicalModels2013],
-a Python library
+list at least 15 Python software including COPASI [@hoopsCOPASICOmplexPAthway2006], Tellurium [@choiTelluriumExtensiblePythonbased2018] and pySB [@lopezProgrammingBiologicalModels2013]. 
+COPASI is a standalone software with a GUI for defining and simulating models.
+It's widely used for its user-friendly interface and comprehensive features.
+There are also python bindings and interfaces for COPASI to allow advanced scripting.
+Tellurium is a Python-based modeling environment,
+however, it uses a C++ library called libRoadRunner in the backend to simulate models.
+PySB is a Python library
 that created a DSL using standard Python to define models
@@
 but wrap libraries in other languages to do the simulation.
-As such, they allow you to "control" the model definition and simulation in a Python script.
-But it is not possible to use other packages from the vast Python community to change the way the simulation is carried out.
-For instance,
-to do a Random Parameter Scan in COPASI,
-we must choose between one of 4 predefined distributions,
-but cannot use one of the many distributions from `scipy.stats`.
-In Tellurium, we cannot choose other Python solvers,
-as adding new integrators has to be done in C++.
+While they enable model definition and simulation control via Python scripts,
+they don't fully leverage Python's extensive package ecosystem.
+For example, in COPASI, users are restricted to predefined distributions for Random Parameter Scans
+and cannot utilize the diverse distributions available in `scipy.stats`
+Similarly, Tellurium doesn’t allow the use of Python solvers,
+as adding new integrators requires C++ implementation.
 
-Another weakness is related to the way models are written.
-While most tools support standards such as the System Biology Markup Language (SBML) [@huckaSBMLL3V2],
-no one is expected to write an SBML file directly.
+Another challenge is the way models are written.
+Many tools use a Domain Specific Language (DSL) for coding,
+and support the System Biology Markup Language (SBML) as an exchange format [@huckaSBMLL3V2],
+as direct SBML coding is impractical.
+A DSL allows reuse of code in different programming environments,
+but it will not be recognized by default in Integrated Development Environments (IDEs).
+Therefore DSLs cannot provide the development experience such as syntax highlighting, code completion, refactoring, and static analysis,
+unless such support is specifcally developed.
 Tellurium uses a Domain Specific Language (DSL) called Antimony [@smithAntimonyModularModel2009],
 which can be translated to SBML.
-A DSL allows code to be reused in different programming environments,
-but it will not be recognized by default in Integrated Development Environments (IDEs)
-and therefore cannot provide the development experience we have come to expect, such as
-syntax highlighting, code completion, refactoring, and static analysis.
-This has been recognized by the community,
-and an extension for Visual Studio Code has been created [@maTestingDynamicalSystems2021].
-However, maintaining it in the future could be burdensome for the smaller systems biology community.
-In the case of PySB,
-it leveraged the highly dynamic nature of Python to create a DSL within it.
+An extension for Visual Studio Code was developed, but its maintenance could be demanding task for the systems biology community.
+PySB, using Python’s dynamic nature, created a DSL within Python.
 By default, it uses global state to create species and parameters,
 without assigning them to Python variables
-or adding them explicitly to the model.
-But IDEs cannot understand it and, hence, provide help, as this is not the way they expect Python code to be written.
+or adding them explicitly to the model,
+but this approach is not fully compatible with IDEs,
+affecting the development experience.
 
-To address these issues,
+To overcome these limitations,
 we developed poincaré and SimBio,
 open-source Python packages for defining and simulating systems.
-Poincaré allows to define systems of differential equations
-and SimBio builds on top of it to define reaction networks.
-They are focused on providing an ergonomic experience for end-users
-by integrating well with IDEs and static analysis tools
+Poincaré allows defining differential equation systems,
+while SimBio builds on it for defining reaction networks.
+They are focused on providing an ergonomic experience to end-users by integrating well with IDEs and static analysis tools
 through the use of standard modern Python syntax.
-Since they are coded in pure Python,
-every part from model definition to simulation can be extended from Python itself.
-For example,
-easier changes such as reusing integrators defined in other packages,
-to more complex ones such as changing the compilation process itself
-to leverage some structure in the equations.
-Their models can be introspected to create other representations,
+Since they are coded in Python,
+every part from model definition to simulation can be extended from Python syntax.
+Being ?first-ever? pure Python packages for systems modelling,
+they offer extensive extensibility,
+from simple tasks like reusing integrators defined in other packages,
+to complex ones like altering the compilation process to leverage some structure in the equations (For example...).
+The models built using these packages can be introspected to create other representations,
 such as graphs connecting species and/or reactions,
 or tables with parameters or equations.
-Finally,
+Futhermore,
 they have a modular architecture with a clear separation of concerns,
@@
 which is beneficial for developers and maintainers.
+We showcased the reliability of these tools by benchmarking them against the simulation results from other tools.
+We also highlighted the substantial performance improvements our tools offer,
+as this is crucial for construction and simulaiton of models of whole cells and organisms,
+which necessitate the simulation of significantly large-scale models.
 
@@
 Symbolite is a lightweight symbolics package to create algebraic mathematical expressions.
-They can be inspected and compiled to different backends.
+Symbolite expressions can be inspected and compiled to various backends.
 Currently, we have implementations for
@@
 and JAX [@jax2018github], a library that support automatic differentiation
-and compilation to GPUs and TPUs;
-and it is easy to add new backends.
+and compilation to GPUs and TPUs.
+Symbolite is designed to facilitate the easy integration of new backends. The source code, documentation to install and use these packages and tutorials are available at the `github page link`.
 
-## Poincaré
+## Versatile modelling and simulation of dynamical systems with Poincaré
 
@@
 It provides a `System` class,
-where we can define `Constant`s, `Parameter`s, `Variable`s,
+where one can define `Constant`s, `Parameter`s, `Variable`s,
 and create equations linking a variable's derivative with an expression (@fig-first-order).
@@
 
-The benefits of using a class to define the system are multiple:
+Utilizing classes for system definition offers several advantages:
 
-1. the variable name to which a component is assigned can be automatically saved in the component for introspection (i.e., `Oscilator.x.name == "x"`),
-2. it provides a namespace such that we can easily define multiple independent models in the same script,
-3. it allows IDEs to provide autocomplete and refactoring capabilities (`Oscillator.<TAB>` shows `x`, `v` and `eq`),
-4. it allows to create instances which can be composed into a bigger model (@fig-composition).
+1. The variable name to which a component is assigned can be automatically saved in the component for introspection (i.e., `Oscilator.x.name == "x"`),
+2. It provides a namespace such that allows easily definition of multiple independent models in the same script,
+3. It allows IDEs to provide autocomplete and refactoring capabilities (`Oscillator.<TAB>` shows `x`, `v` and `eq`),
+4. It allows creatoion of instances which can be composed into a bigger model (@fig-composition).
 
-For the latter,
+For the latter (???, not clear rephrase, probably after the next sentence),
 it is important to add the type annotations,
@@
 
-Using type annotations also allows to statically check for typing errors.
+Type annotations play a significant role in static type checking, helping to identify typing errors.
 For instance, `Parameter`s can be assigned either a number or a time-dependent expression.
-But can also be linked to an outside variable when composing models.
-As such, they cannot be used to parametrize the initial conditions of variables,
-for which we have to use a `Constant` (@fig-ide).
+But can also be linked to an external variable when composing models.
+As such, they (what?) cannot be used to parametrize the initial conditions of variables,
+for which we have to use a `Constant` (@fig-ide). (message not clear, rephrase)
 
@@
 :::
+(add arrows to the figure)
 
 To simulate a system,
-we create a `Simulator` instance (@fig-sim),
+we created a `Simulator` instance (@fig-sim),
 which compiles a given system and interfaces with solvers.
@@
 using `numpy` as a backend.
+This can be easily switched to other solvers. 
 The Simulator wraps the output in a `pandas.DataFrame`,
@@
 ```{python}
-df.plot(figsize=(4, 2))
+df.plot();
 ```
@@
 on a factor up to x30 or more for big or long running models,
-but incurs in a compile time penalty for the first run
+however incurs in a compile time penalty for the first run
 which must be taken into account.
 
-## SimBio
+## Extensible definition of reaction networks using SimBio
 
@@
 
-For reaction networks,
-we are interested in first-order differential equations
-relating the rate of change of the species.
-To facilitate defining these models,
-SimBio defines `Species`,
-which comprises of a `poincare.Variable` and an stoichiometry,
+For the reaction networks,
+our focus is on first-order differential equations
+that describe the rate of change of species.
+SimBio simplifies the definition of these network models by introducing `Species`,
+which consists of a `poincare.Variable` and stoichiometry,
 and `Reaction`,
-which transform reactants into products taking into account the stoichiometry (@fig-simbio).
-It also includes `MassAction`, a `Reaction` subclass,
-which automatically adds the reactants as part of the rate law (@fig-simbio).
+a construct that converts reactants into products taking into account the stoichiometry (@fig-simbio).
+Additionally, SimBio features `MassAction`, a subclass of `Reaction`,
+which intuitively incorporates reactants into the rate law (@fig-simbio).
+(is `RateLaw` not better term than `MassAction` ?)
+
 Several commonly used reactions are predefined as `MassAction` subclasses,
@@
 
-To test the reproducibility of results of SimBio,
-we considered the first 100 curated models from BioModels.
-As SimBio does not yet support the full SBML standard (for instance, events),
-we could only simulate 60 of those.
-Using the results from COPASI as ground truth,
-SimBio reproduces the results with numerical differences depending on the solver tolerance.
+To evaluate SimBio's reproducibility,
+we analyzed curated SBML models from BioModels (ref needed).
+Among the first 100 curated models from BioModels we selected 60 which did not contains events,
+as SimBio doesn't support events yet.
+We simulated the selected models with COPASI
+and used the simulated results as ground truth,
+and demonstrated that SimBio reproduces the results with minor numerical differences depending on the solver tolerance.
 
-To test the performance,
-we simulated them with COPASI, Tellurium/RoadRunner and simbio.
-For simbio, we considered both the NumPy and numba backends.
-For COPASI and simbio, we used the LSODA solver,
-while using the similar CVODE solver for tellurium.
-In all cases, we used relative and absolute tolerances of $10^{-6}$.
-We measured separately three stages of the simulation:
-the loading, first/cold run and subsequent/warm runs of each model.
+For performance testing,
+we ran simulations using COPASI, Tellurium/RoadRunner, and SimBio.
+Within SimBio, both NumPy and Numba backends were considered.
+The LSODA solver was used for COPASI and SimBio,
+while for Tellurium the comparable CVODE solver was used.
+In all cases, we used relative and absolute tolerances of $10^{-6}$. We measured three simulation stages:
+loading, the initial (cold) run, and subsequent (warm) runs for each model.
 
@@
 In this article,
-we have described a set of Python packages we developed
-that allow to define and simulate dynamical systems and reaction networks.
-They are focused on a deep integration with IDEs
-such that code analysis tools can flag errors before running the code,
-and help with refactoring and code completion.
-For this,
-we chose to use standard modern Python syntax,
-since the integration with IDEs will be maintained by the big players in the huge Python community.
+we introduced a suite of Python packages we developed
+for defining and simulating dynamical systems and reaction networks.
+These packages are deeply integrated with Integrated Development Environments (IDEs),
+enabling code analysis tools to identify errors prior to execution and assist in refactoring and code completion.
+We adopted standard modern Python syntax to ensure seamless IDE integration, supported by the extensive Python community.
 
-Unlike previous tools,
-both model definition and compilation into an ODE function are implemented in Python.
-This lowers the barrier to developing different simulation modalities,
-such as performance optimizations by leveraging some structure in the model.
-However, being in Python does not mean that performance is less than that of C/C++ tools,
-as the resulting ODE function can be JITted using Numba.
+Our approach differs from previous tools in that both the model definition and its compilation into an Ordinary Differential Equation (ODE) function are entirely Python-based.
+This approach simplifies the development of various simulation methods,
+including performance enhancements that exploit specific model structures.
+Importantly, being Python-based does not compromise performance compared to C/C++ tools, as the ODE functions can be Just-In-Time (JIT) compiled using Numba.
 
-These packages are designed with a modular architecture that clearly separates different concerns.
-This makes them easier to reuse, improve, and extend by the broader Python community.
-For example, someone outside the systems biology community could add a stochastic integrator to poincaré,
-which could then be used from SimBio.
-This architecture also makes them easier to understand,
-which reduces the difficulty of contributing fixes or new code,
-and ensures their maintenance and continued development into the future.
+The inclusion of SBML support facilitates the effortless reuse of models created by the systems biology community,
+along with the vast collection of public models hosted in the BioModels repository.
+The modular architecture of these packages facilitates their reuse, enhancement, and extension by the wider Python community.
+For instance, an individual from outside the systems biology field could contribute a stochastic integrator to poincaré,
+which would then be available in SimBio.
+This clear separation of concerns also makes the packages more comprehensible,
+lowering the barrier for contributing improvements or new features.
+Such an architecture ensures their maintainability and ongoing development well into the future.
+
+`Gen Comment: I amnto sure whether the number of figures are already over the limit, you may have to combine figures, while doing so, if space permits create a small schematics / summary figure, with icons and texts showing the significance / and applicability of the tools developed.`
 

### 2023-12-13 article.qmd
Spellcheck

@@
 Chemical Reaction Networks (CRNs)
-are a fundamental concept of modelling in numerous fields including systems biology, biochemistry, chemical engineering and epidemeology.
-They are comprised of a set of chemical species or biological entities and represntation of complex interaction and transformation between them through a series reactions.
+are a fundamental concept of modelling in numerous fields including systems biology, biochemistry, chemical engineering and epidemiology.
+They are comprised of a set of chemical species or biological entities and representation of complex interaction and transformation between them through a series reactions.
 These systems can be modelled and simulated through multiple approaches:
@@
 a registry of simulation tools,
-list at least 15 Python software including COPASI [@hoopsCOPASICOmplexPAthway2006], Tellurium [@choiTelluriumExtensiblePythonbased2018] and pySB [@lopezProgrammingBiologicalModels2013]. 
+list at least 15 Python software including COPASI [@hoopsCOPASICOmplexPAthway2006], Tellurium [@choiTelluriumExtensiblePythonbased2018] and PySB [@lopezProgrammingBiologicalModels2013]. 
 COPASI is a standalone software with a GUI for defining and simulating models.
@@
 Therefore DSLs cannot provide the development experience such as syntax highlighting, code completion, refactoring, and static analysis,
-unless such support is specifcally developed.
+unless such support is specifically developed.
 Tellurium uses a Domain Specific Language (DSL) called Antimony [@smithAntimonyModularModel2009],
@@
 or tables with parameters or equations.
-Futhermore,
+Furthermore,
 they have a modular architecture with a clear separation of concerns,
@@
 We also highlighted the substantial performance improvements our tools offer,
-as this is crucial for construction and simulaiton of models of whole cells and organisms,
+as this is crucial for construction and simulation of models of whole cells and organisms,
 which necessitate the simulation of significantly large-scale models.
@@
 3. It allows IDEs to provide autocomplete and refactoring capabilities (`Oscillator.<TAB>` shows `x`, `v` and `eq`),
-4. It allows creatoion of instances which can be composed into a bigger model (@fig-composition).
+4. It allows creation of instances which can be composed into a bigger model (@fig-composition).
 
@@
 
-![Performance of different softwares to solve models from the cureted section of BioModels.
+![Performance of different softwares to solve models from the curated section of BioModels.
 (left) Run time for the model BIOMD3 as a function of the number of output points.

### 2023-12-13 src/ide/constant.py
Remove unused IDE figures

@@ -1,7 +1,0 @@
-from poincare import Constant, System, Variable, assign, initial
-
-
-class Model(System):
-    c: Constant = assign(default=1, constant=True)
-    x: Variable = initial(default=c)
-    y: Variable = initial(default=c)

### 2023-12-13 src/ide/ide2.py

@@ -1,6 +1,6 @@
 from poincare import System, Variable, initial  # noqa: I001
 
-from .autocomplete import Model
+from .ide1 import Model
 
 
 

### 2023-12-13 src/ide/type_error.py

@@ -1,8 +1,0 @@
-from poincare import Parameter, System, Variable, assign, initial
-
-
-class BadModel(System):
-    p: Parameter = assign(default=1)
-    z: Variable = initial(default=p)
-    eq = z.derive() << p * z
-

### 2023-12-13 article.qmd
Add arrows to figures

@@
 ::: {#fig-ide layout-ncol=2 layout-valign="bottom"}
-![](src/ide/ide1.png)
+![](src/ide/ide1.svg)
 
-![](src/ide/ide2.png)
+![](src/ide/ide2.svg)
 
@@
 :::
-(add arrows to the figure)
 

### 2023-12-13 article.qmd
First-ever pure Python

@@
 every part from model definition to simulation can be extended from Python syntax.
-Being ?first-ever? pure Python packages for systems modelling,
+Being the first-ever pure Python packages for systems modelling,
 they offer extensive extensibility,

### 2023-12-13 article.qmd
Add mention to PyPI

@@
 and simbio, to define reaction systems and interface with systems biology standards such as SBML.
+These are pure Python packages with standard dependencies from the PyData scientific stack such as NumPy [@harrisArrayProgrammingNumPy2020] and pandas [@mckinneyDataStructuresStatistical2010].
+They are published in PyPI (the Python Package Index),
+where links to the source code and documentation hosted in GitHub can be found,
+and can be easily installed with `pip install <package_name>`.
 
@@
 and compilation to GPUs and TPUs.
-Symbolite is designed to facilitate the easy integration of new backends. The source code, documentation to install and use these packages and tutorials are available at the `github page link`.
+Symbolite is designed to facilitate the easy integration of new backends.
 

### 2023-12-13 article.qmd
Paraphrase typing errors

@@
 1. The variable name to which a component is assigned can be automatically saved in the component for introspection (i.e., `Oscilator.x.name == "x"`),
-2. It provides a namespace such that allows easily definition of multiple independent models in the same script,
+2. It provides a namespace such that allows to easily define of multiple independent models in the same script,
 3. It allows IDEs to provide autocomplete and refactoring capabilities (`Oscillator.<TAB>` shows `x`, `v` and `eq`),
@@
 
-For the latter (???, not clear rephrase, probably after the next sentence),
-it is important to add the type annotations,
-as it can provide a signature autocompletion in IDEs supporting the `dataclass_transform` [@debontePEP681Data2021]:
-
-Type annotations play a significant role in static type checking, helping to identify typing errors.
-For instance, `Parameter`s can be assigned either a number or a time-dependent expression.
-But can also be linked to an external variable when composing models.
-As such, they (what?) cannot be used to parametrize the initial conditions of variables,
-for which we have to use a `Constant` (@fig-ide). (message not clear, rephrase)
+For this last point,
+IDEs that support `dataclass_transform` [@debontePEP681Data2021] can provide a tooltip with the expected signature (@fig-composition).
+This requires the use of type annotations
+which play a more significant role in static type checking,
+as they can help to identify errors before running the code.
+For instance, to parametrize the initial conditions of variables we have to use a `Constant`.
+If we try to use a `Parameter`, which could be a time-dependent expression, it is flagged as a type error (@fig-ide).
 

### 2023-12-13 article.qmd
Add biomodels reference

@@
 Additionally, SimBio supports importing from and exporting to SBML,
-and downloading them directly from BioModels (@fig-simbio-io).
+and downloading them directly from BioModels [@malik-sheriffBioModels15Years2020] (@fig-simbio-io).
 
@@
 To evaluate SimBio's reproducibility,
-we analyzed curated SBML models from BioModels (ref needed).
+we analyzed curated SBML models from BioModels [@malik-sheriffBioModels15Years2020].
 Among the first 100 curated models from BioModels we selected 60 which did not contains events,

### 2023-12-13 article.qmd
As TODO and bold to pending things

@@
 from simple tasks like reusing integrators defined in other packages,
-to complex ones like altering the compilation process to leverage some structure in the equations (For example...).
+to complex ones like altering the compilation process to leverage some structure in the equations.
+**TODO: For example...**
 The models built using these packages can be introspected to create other representations,
@@
 which intuitively incorporates reactants into the rate law (@fig-simbio).
-(is `RateLaw` not better term than `MassAction` ?)
+**TODO: is `RateLaw` not better term than `MassAction`?**
 
@@
 
-`Gen Comment: I amnto sure whether the number of figures are already over the limit, you may have to combine figures, while doing so, if space permits create a small schematics / summary figure, with icons and texts showing the significance / and applicability of the tools developed.`
+**TODO: I amnto sure whether the number of figures are already over the limit, you may have to combine figures, while doing so, if space permits create a small schematics / summary figure, with icons and texts showing the significance / and applicability of the tools developed.**
 

### 2023-12-13 article.qmd
Fix LaTeX math

@@
 such as
-`MichaelisMenten` ($S + E \rightarrow ES -> P + E$)
+`MichaelisMenten` ($S + E \leftrightarrow ES \rightarrow P + E$)
 and its approximate form without the intermediate species $ES$,

### 2023-12-14 article.qmd
Remove RateLaw comment
Rate law is an experimental law i.e. it is based on observations. On the other hand, law of mass action is a theoretical law based on the balanced chemical reaction.
Remove number of figure comment
For the application note it should be 2600 words,
or 2000 words plus 1 figure.

We have:
- 2914 words in total
- 2493 (-421) excluding references
- 2294 (-199) excluding code snippets
- 1947 (-347) excluding figure captions

@@
 which intuitively incorporates reactants into the rate law (@fig-simbio).
-**TODO: is `RateLaw` not better term than `MassAction`?**
 
@@
 
-**TODO: I amnto sure whether the number of figures are already over the limit, you may have to combine figures, while doing so, if space permits create a small schematics / summary figure, with icons and texts showing the significance / and applicability of the tools developed.**
-
 # References

### 2023-12-14 article.qmd
Add for-loop example in the introduction as a complex extensibility example

@@
 to complex ones like altering the compilation process to leverage some structure in the equations.
-**TODO: For example...**
+For example,
+using a for-loop in the compiled equations could improve the runtime performance
+if there is some repetitive structure in the system,
+as happens in spatial modelling.
 The models built using these packages can be introspected to create other representations,

### 2023-12-14 article.qmd
Spell-checker

@@
 as they can help to identify errors before running the code.
-For instance, to parametrize the initial conditions of variables we have to use a `Constant`.
+For instance, to parameterize the initial conditions of variables we have to use a `Constant`.
 If we try to use a `Parameter`, which could be a time-dependent expression, it is flagged as a type error (@fig-ide).
@@
 SimBio simplifies the definition of these network models by introducing `Species`,
-which consists of a `poincare.Variable` and stoichiometry,
+which consists of a `poincare.Variable` and a stoichiometric number,
 and `Reaction`,

### 2023-12-14 src/ide/ide2.py
Run pre-commit on all files

@@ -1,13 +1,6 @@
-from poincare import System, Variable, initial  # noqa: I001
+from poincare import System, Variable, initial
 
 from .ide1 import Model
-
-
-
-
-
-
-
 
 
 class BigModel(System):

### 2023-12-14 src/performance/plot.py
Add performance module

@@ -1,0 +1,69 @@
+from pathlib import Path
+
+import pandas as pd
+from matplotlib.figure import Figure
+
+
+def load(file):
+    return pd.read_csv(file).drop(columns="t_end").set_index("model_id")
+
+
+def plot_number_of_points(fig: Figure, df: dict[str, pd.DataFrame], model_id: str):
+    ax = fig.add_subplot()
+    ax.set_xlabel("Number of evaluated time points")
+    ax.set_ylabel("Run time [s]")
+    ax.set_ylim(1e-4, 2e-2)
+
+    times = (
+        pd.DataFrame(v.loc[model_id].rename(k) for k, v in df.items())
+        .drop(columns=["load", "first_run"])
+        .T
+    )
+    times.index = times.index.astype(float)
+    times.plot(ax=ax, logx=True, logy=True, marker="o", legend=False)
+
+
+def plot_runtime(fig: Figure, df: dict[str, pd.DataFrame]):
+    ax = fig.add_subplot()
+    ax.set_ylabel("Run time [s]")
+    ax.set_xlabel("Geometric mean run time [s]")
+
+    n_points = "300"
+    geom_average = 1 / sum(1 / v[n_points] for v in df.values())
+
+    ax.axline((1e-3, 1e-3), slope=1, color="black")
+    for k, v in df.items():
+        ax.scatter(geom_average, v[n_points], s=5)
+    ax.set(xscale="log", yscale="log")
+
+
+if __name__ == "__main__":
+    import sys
+
+    import matplotlib.pyplot as plt
+    import seaborn
+
+    path = Path(__file__).parent
+
+    files = {
+        "COPASI": "copasi.csv",
+        "RoadRunner": "roadrunner.csv",
+        "SimBio NumPy": "simbio-numpy-lsoda.csv",
+        "SimBio numba": "simbio-numba-lsoda.csv",
+        "SimBio numba\nand numbalsoda": "simbio-numba-numbalsoda.csv",
+    }
+
+    df = {k: load(path / f"results/{v}") for k, v in files.items()}
+
+    fig = plt.figure(layout="constrained", figsize=(8, 2.5))
+    figs = fig.subfigures(ncols=3, width_ratios=[1, 0.7, 1])
+    plot_number_of_points(figs[0], df, "BIOMD3")
+    plot_runtime(figs[2], df)
+    fig.legend(loc="center", bbox_to_anchor=(0.5, 0.5), frameon=False)
+    seaborn.despine(fig)
+
[…]

### 2023-12-14 src/performance/runners/copasi.py

@@ -1,0 +1,25 @@
+from dataclasses import dataclass
+
+import basico
+
+from ..timer import Runner
+
+
+@dataclass
+class COPASI(Runner):
+    atol: float
+    rtol: float
+
+    def load_model(self, sbml: str):
+        self.runner = basico.load_model_from_string(sbml)
+        return self
+
+    def run_model(self, *, stop: float, n_points: int):
+        return basico.run_time_course(
+            0,
+            stop,
+            n_points - 1,
+            a_tol=self.atol,
+            r_tol=self.rtol,
+            method="lsoda",
+        )

### 2023-12-14 src/performance/runners/roadrunner.py

@@ -1,0 +1,23 @@
+from dataclasses import dataclass
+
+import tellurium
+
+from ..timer import Runner
+
+
+@dataclass
+class Tellurium(Runner):
+    atol: float
+    rtol: float
+
+    def load_model(self, sbml: str):
+        self.runner = tellurium.loadSBMLModel(sbml)
+        return self
+
+    def run_model(self, *, stop: float, n_points: int):
+        self.runner.reset()
+        self.runner.integrator = "cvode"
+        self.runner.integrator.absolute_tolerance = self.atol
+        self.runner.integrator.relative_tolerance = self.rtol
+        self.runner.integrator.variable_step_size = True
+        return self.runner.simulate(0, stop, n_points)

### 2023-12-14 src/performance/runners/simbio.py

@@ -1,0 +1,56 @@
+from __future__ import annotations
+
+from dataclasses import dataclass
+from typing import Literal
+
+import numpy as np
+import scipy.integrate
+from poincare.simulator import Problem
+from poincare.solvers import LSODA, Solution
+from simbio import Simulator
+from simbio.io.sbml import loads
+
+from ..timer import Runner
+
+
+@dataclass
+class SimBio(Runner):
+    atol: float
+    rtol: float
+    solver: type[LSODA] | type[CVODE]
+    backend: Literal["numpy", "numba"]
+
+    def load_model(self, sbml: str):
+        self.runner = Simulator(loads(sbml), backend=self.backend)
+        return self
+
+    def run_model(self, *, stop: float, n_points: int):
+        return self.runner.solve(
+            save_at=np.linspace(0, stop, n_points),
+            solver=self.solver(atol=self.atol, rtol=self.rtol),
+        )
+
+
+@dataclass(frozen=True, kw_only=True)
+class CVODE:
+    atol: float
+    rtol: float
+
+    def __call__(self, problem: Problem, *, save_at: np.ndarray):
+        def func(t, y, p):
+            dy = np.empty_like(y)
+            return problem.rhs(t, y, p, dy)
+
+        solver = scipy.integrate.ode(func)
+        solver.set_initial_value(problem.y, float(problem.t[0]))
+        solver.set_f_params(problem.p)
+        solver.set_integrator("vode", atol=self.atol, rtol=self.rtol)
+        out = np.empty((save_at.size, len(problem.y)))
+        if save_at[0] == 0:
+            out[0] = problem.y
+        else:
+            out[0] = solver.integrate(save_at[0], relax=True)
+        for i, t in enumerate(save_at[1:-1], start=1):
+            out[i] = solver.integrate(t, relax=True)
+        out[-1] = solver.integrate(save_at[-1], relax=False)
+        return Solution(save_at, out)

### 2023-12-14 src/performance/timer.py

@@ -1,0 +1,131 @@
+from __future__ import annotations
+
+from dataclasses import dataclass
+from time import perf_counter
+from typing import Protocol
+
+import pandas as pd
+from biomodels import get_omex
+from tqdm import tqdm
+
+tqdm.pandas()
+
+
+class Runner(Protocol):
+    atol: float
+    rtol: float
+
+    def __init__(self, *, atol: float, rtol: float):
+        ...
+
+    def load_model(self, sbml: str):
+        ...
+
+    def run_model(self, *, stop: float, n_points: int):
+        ...
+
+
+@dataclass
+class Model:
+    sbml: str
+    stop: float
+
+
+def time(
+    *,
+    runner: Runner,
+    model: Model,
+    n_points: list[int],
+    n_runs: int,
+) -> dict[str, float]:
+    time_start = perf_counter()
+    runner.load_model(model.sbml)
+    time_loaded = perf_counter()
+    runner.run_model(stop=model.stop, n_points=n_points[0])
+    time_first_run = perf_counter()
+
+    output = {
+        "load": time_loaded - time_start,
+        "first_run": time_first_run - time_loaded,
+    }
+
+    for n in n_points:
+        start = perf_counter()
+        for _ in range(n_runs):
+            runner.run_model(stop=model.stop, n_points=n)
+        stop = perf_counter()
+        output[str(n)] = (stop - start) / n_runs
+    return output
+
+
+def run_model(model_id: str, t_end: float, *, runner: type[Runner]):
+    omex = get_omex(model_id)
+    model = Model(
+        sbml=omex.master.read_text(),
+        stop=t_end,
+    )
+    try:
+        result = time(
+            runner=runner(atol=1e-9, rtol=1e-6),
+            model=model,
+            n_points=[10, 30, 100, 300, 1000, 3000],
+            n_runs=100,
+        )
+
+    except Exception:
+        raise
+    else:
+        return {
+            "model_id": model_id,
+            "t_end": t_end,
+            **result,
+        }
+
+
+if __name__ == "__main__":
+    import sys
+    from functools import partial
+    from pathlib import Path
+
+    match sys.argv[1:]:
+        case ["copasi"]:
+            from .runners.copasi import COPASI as runner
+        case ["roadrunner"]:
+            from .runners.roadrunner import Tellurium as runner
+        case ["simbio", "numpy", "lsoda"]:
+            from .runners.simbio import LSODA, SimBio
+
[…]

### 2023-12-14 src/performance/test_results.py
Add test to compare results between COPASI and SimBio

@@ -1,0 +1,52 @@
+from pathlib import Path
+
+import biomodels
+import pandas as pd
+from pytest import mark
+from sbml.pydantic import loads as sbml_loads
+
+from .runners import copasi, simbio
+
+path = Path(__file__).parent
+t_end = pd.read_csv(path / "models.txt").set_index("model_id")["t_end"]
+
+tol = dict(atol=1e-6, rtol=1e-6)
+tol_compare = dict(atol=1e-2, rtol=1e-2)
+
+
+@mark.parametrize("model_id", t_end.index)
+def test_results(model_id: str):
+    t = t_end[model_id]
+    zipfile = biomodels.get_omex(model_id).master
+    sbml = zipfile.read_text()
+
+    runners = [
+        copasi.COPASI(**tol),
+        simbio.SimBio(**tol, solver=simbio.LSODA, backend="numpy"),
+    ]
+
+    result_copasi, result_simbio = (
+        r.load_model(sbml).run_model(stop=t, n_points=1000) for r in runners
+    )
+
+    result_copasi.index = result_copasi.index.rename("time").round(3)
+    result_simbio.index = result_simbio.index.rename("time").round(3)
+
+    sbml = sbml_loads(zipfile.read_bytes())
+    name_mapping = {s.id: s.name for s in sbml.model.species if s.name is not None}
+
+    def renamer_copasi(x: str):
+        if x.startswith("Values["):
+            return x.removeprefix("Values[").removesuffix("]")
+        return x
+
+    result_copasi = result_copasi.rename(columns=renamer_copasi)
+    result_simbio = result_simbio.rename(columns=name_mapping)
+
+    common_columns = result_simbio.columns.intersection(result_copasi.columns)
+
+    pd.testing.assert_frame_equal(
+        result_copasi[common_columns],
+        result_simbio[common_columns],
+        **tol_compare,
+    )

### 2023-12-15 article.qmd
Run pre-commit on all files
Remove SBML downloading message from pooch
Add missing dot

@@
 For example, in COPASI, users are restricted to predefined distributions for Random Parameter Scans
-and cannot utilize the diverse distributions available in `scipy.stats`
+and cannot utilize the diverse distributions available in `scipy.stats`.
 Similarly, Tellurium doesn’t allow the use of Python solvers,
@@
 
+```{python}
+# | echo: false
+# Pre-download SBML file, which is saved to a cache.
+# Avoids printing the "downloading" message
+from simbio.io import biomodels, sbml
+
+sbml.load("repressilator.sbml")
+```
+
 ::: {#fig-simbio-io layout-ncol=1}

### 2023-12-15 article.qmd
Add abstract

@@
 title: "Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems."
+abstract: >
+    This work introduces two novel Python packages:
+    Poincaré, for definition and simulation of dynamical systems,
+    and SimBio, which builds on top of Poincaré to
+    facilitate the definition of Chemical Reaction Networks (CRNs),
+    and add support for the Systems Biology Markup Language (SBML).
+
+    To ensure a user-friendly experience,
+    they use of standard typed modern Python syntax
+    that provide a seamless integration with Integrated Development Environments (IDEs),
+    enhances code analysis tools, enabling error identification before execution,
+    and facilitates code completion and refactoring.
+
+    In contrast to existing Python tools for CRN modeling,
+    that wrap external C/C++ libraries for simulation,
+    Poincaré and SimBio are pure Python packages,
+    which allows to extend their simulation capabilities by writing Python or leveraging other Python packages.
+    But this does not compromise performance,
+    as code can be Just-In-Time compiled with Numba,
+    and benchmarks show comparable or superior performance to other tools.
+
+    In conclusion, the Python-centric approach,
+    coupled with the packages' extensibiltiy,
+    positions Poincaré and SimBio as valuable tools for the modeling community.
 bibliography: bibliography.bib

### 2023-12-15 article.qmd
Add output false to code block
Fix pre-download of model
Improving abstract

--- 2023-12-15 article.qmd
@@
 title: "Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems."
+abstract: >
+    Chemical Reaction Networks (CRNs) are a fundamental concept of modelling in numerous fields,
+    including systems biology, biochemistry, chemical engineering and epidemiology.
+    Formulating them in a higher-level description
+    allows to solve them with different approaches,
+    such as deterministic or stochastic simulations.
+    Existing Python tools for CRN modeling wrap external C/C++ libraries for modelling and simulation,
+    which limits their extensibility to new simulation approaches
+    or to reuse packages from the vast Python community.
+
+    This work introduces two novel Python packages:
+    Poincaré, for definition and simulation of dynamical systems,
+    and SimBio, which builds on top of Poincaré to
+    facilitate the definition of Chemical Reaction Networks (CRNs),
+    and add support for the Systems Biology Markup Language (SBML).
+
+    Poincaré and SimBio are pure Python packages,
+    which allows to extend their simulation capabilities by writing Python or leveraging other Python packages.
+    But this does not compromise performance,
+    as code can be Just-In-Time compiled with Numba,
+    and benchmarks show comparable or superior performance to other tools.
+
+    Additionally,
+    to ensure a user-friendly experience,
+    they use of standard typed modern Python syntax
+    that provide a seamless integration with Integrated Development Environments (IDEs),
+    enhances code analysis tools, enabling error identification before execution,
+    and facilitates code completion and refactoring.
+
+    We believe that this Python-centric approach positions Poincaré and SimBio as valuable tools for the modeling community.
 bibliography: bibliography.bib
@@
 # | echo: false
+# | output: false
 # Pre-download SBML file, which is saved to a cache.
@@
 
-sbml.load("repressilator.sbml")
+biomodels.load_model("BIOMD12")
 ```

### 2023-12-15 article.qmd
Add authors

@@
     We believe that this Python-centric approach positions Poincaré and SimBio as valuable tools for the modeling community.
+
+authors:
+    -   name: Mauro Silberberg
+        orcid: 0000-0002-2402-1100
+        email: maurosilber@df.uba.ar
+        affiliations:
+        -   ref: UBA
+        -   ref: CONICET
+    -   name: Rahuman S. Malik-Sheriff
+        orcid: 0000-0003-0705-9809
+        email: sheriff@ebi.ac.uk
+        affiliations:
+        -   ref: EBI
+    -   name: Hernán E. Grecco
+        corresponding: true
+        orcid: 0000-0002-1165-4320
+        email: hgrecco@df.uba.ar
+        affiliations:
+        -   ref: UBA
+        -   ref: CONICET
+
+affiliations:
+    -   id: UBA
+        name: Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Física. Buenos Aires, Argentina.
+    -   id: CONICET
+        name: CONICET - Universidad de Buenos Aires, Instituto de Física de Buenos Aires (IFIBA). Buenos Aires, Argentina
+    -   id: EBI
+        name: European Bioinformatics Institute, European Molecular Biology Laboratory (EMBL-EBI), Wellcome Trust Genome Campus, Cambridge, UK
+        url: https://ebi.ac.uk/
+
 bibliography: bibliography.bib

### 2023-12-19 article.qmd
Changed abstract

@@
 abstract: >
-    Chemical Reaction Networks (CRNs) are a fundamental concept of modelling in numerous fields,
-    including systems biology, biochemistry, chemical engineering and epidemiology.
-    Formulating them in a higher-level description
-    allows to solve them with different approaches,
-    such as deterministic or stochastic simulations.
-    Existing Python tools for CRN modeling wrap external C/C++ libraries for modelling and simulation,
-    which limits their extensibility to new simulation approaches
-    or to reuse packages from the vast Python community.
-
-    This work introduces two novel Python packages:
-    Poincaré, for definition and simulation of dynamical systems,
-    and SimBio, which builds on top of Poincaré to
-    facilitate the definition of Chemical Reaction Networks (CRNs),
-    and add support for the Systems Biology Markup Language (SBML).
-
-    Poincaré and SimBio are pure Python packages,
-    which allows to extend their simulation capabilities by writing Python or leveraging other Python packages.
-    But this does not compromise performance,
-    as code can be Just-In-Time compiled with Numba,
-    and benchmarks show comparable or superior performance to other tools.
+    Chemical Reaction Networks (CRNs) play a pivotal role in diverse fields 
+    such as systems biology, biochemistry, chemical engineering, and epidemiology. 
+    High-level modelling of CRNs enables various simulation approaches, 
+    including deterministic and stochastic methods. 
+    However, existing Python tools for CRN modelling typically wrap external C/C++ libraries for modelling and simulation, 
+    limiting their extensibility and integration with the broader Python ecosystem.
+    In response, we developed Poincaré and SimBio, 
+    two novel Python packages for the definition and simulation of dynamical systems and CRNs. 
+    Poincaré serves as a foundation for dynamical system modelling, 
+    while SimBio extends this functionality to CRNs, 
+    including support for the Systems Biology Markup Language (SBML). 
+    Poincaré and SimBio are developed as pure Python packages 
+    enabling users to easily extend their simulation capabilities 
+    by writing new or leveraging other Python packages.
+    Moreover, this does not compromise the performance, 
+    as code can be Just-In-Time compiled with Numba. 
+    Our benchmark tests using curated models from the BioModels repository, 
+    demonstrate that these tools may provide a potentially superior performance advantage, 
+    compared to other existing tools.
 
@@
     to ensure a user-friendly experience,
-    they use of standard typed modern Python syntax
-    that provide a seamless integration with Integrated Development Environments (IDEs),
-    enhances code analysis tools, enabling error identification before execution,
-    and facilitates code completion and refactoring.
+    our packages use standard typed modern Python syntax
+    that provides a seamless integration with Integrated Development Environments (IDEs).
 
-    We believe that this Python-centric approach positions Poincaré and SimBio as valuable tools for the modeling community.
+    Python-centric approach significantly enhances code analysis, 
+    error detection, and refactoring capabilities, 
+    positioning Poincaré and SimBio as valuable tools for the modelling community.
 

### 2024-01-02 article.qmd
Add Henning to authors

@@
         email: sheriff@ebi.ac.uk
+        affiliations:
+        -   ref: EBI
+    -   name: Henning Hermjakob
+        orcid: 0000-0001-8479-0262
+        email: hhe@ebi.ac.uk
         affiliations:
@@
 ```{python}
-from simbio import Compartment, MassAction, Species, Reaction, initial
+from simbio import Compartment, MassAction, Species, RateLaw, initial
 
@@
     B: Species = initial(default=0)
-    r = Reaction(
+    r = RateLaw(
         reactants=[2 * A],
@@
 A single reaction transforming $2A$ into $B$ is saved in variable `r`.
-The rate $1$ is specified directly for `Reaction`,
+The rate $1$ is specified directly for `RateLaw`,
 and is proportional to the reactants for `MassAction`.
@@
 which consists of a `poincare.Variable` and a stoichiometric number,
-and `Reaction`,
+and `RateLaw`,
 a construct that converts reactants into products taking into account the stoichiometry (@fig-simbio).
-Additionally, SimBio features `MassAction`, a subclass of `Reaction`,
+Additionally, SimBio features `MassAction`, a subclass of `RateLaw`,
 which intuitively incorporates reactants into the rate law (@fig-simbio).
@@
 and its approximate form without the intermediate species $ES$,
-and it is also simple to implement your own as subclasses of `Reaction` or `MassAction`.
+and it is also simple to implement your own as subclasses of `RateLaw` or `MassAction`.
 Additionally, SimBio supports importing from and exporting to SBML,

### 2024-01-05 article.qmd
Rename Reaction to RateLaw

@@
         email: sheriff@ebi.ac.uk
-        affiliations:
-        -   ref: EBI
-    -   name: Henning Hermjakob
-        orcid: 0000-0001-8479-0262
-        email: hhe@ebi.ac.uk
         affiliations:

### 2024-01-08 article.tex
Add tex file

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{241,243,245}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.40,0.45,0.13}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\ExtensionTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.28,0.35,0.67}{#1}}
\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.46,0.62}{#1}}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.07,0.07,0.07}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\KOMAoption{captions}{tableheading}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems.},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Poincaré and SimBio: a versatile and extensible Python ecosystem
for modeling systems.}
\author{Mauro Silberberg \and Rahuman S. Malik-Sheriff \and Henning
Hermjakob \and Hernán E. Grecco}
\date{}

\begin{document}
\maketitle
\begin{abstract}
Chemical Reaction Networks (CRNs) play a pivotal role in diverse fields
such as systems biology, biochemistry, chemical engineering, and
epidemiology. High-level modelling of CRNs enables various simulation
approaches, including deterministic and stochastic methods. However,
existing Python tools for CRN modelling typically wrap external C/C++
libraries for modelling and simulation, limiting their extensibility and
integration with the broader Python ecosystem. In response, we developed
Poincaré and SimBio, two novel Python packages for the definition and
simulation of dynamical systems and CRNs. Poincaré serves as a
foundation for dynamical system modelling, while SimBio extends this
functionality to CRNs, including support for the Systems Biology Markup
Language (SBML). Poincaré and SimBio are developed as pure Python
packages enabling users to easily extend their simulation capabilities
by writing new or leveraging other Python packages. Moreover, this does
not compromise the performance, as code can be Just-In-Time compiled
with Numba. Our benchmark tests using curated models from the BioModels
repository, demonstrate that these tools may provide a potentially
superior performance advantage, compared to other existing tools.
Additionally, to ensure a user-friendly experience, our packages use
standard typed modern Python syntax that provides a seamless integration
with Integrated Development Environments (IDEs). Python-centric approach
significantly enhances code analysis, error detection, and refactoring
capabilities, positioning Poincaré and SimBio as valuable tools for the
modelling community.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, frame hidden, interior hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, enhanced]}{\end{tcolorbox}}\fi

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Chemical Reaction Networks (CRNs) are a fundamental concept of modelling
in numerous fields including systems biology, biochemistry, chemical
engineering and epidemiology. They are comprised of a set of chemical
species or biological entities and representation of complex interaction
and transformation between them through a series reactions. These
systems can be modelled and simulated through multiple approaches:
deterministic Ordinary Differential Equations (ODEs) to model
macroscopic behavior, Stochastic Differential Equations (SDEs) to model
microscopic fluctuations, and jump processes (Gillespie-like
simulations) to account for the discreteness of populations. Instead of
directly writing the equations for each of these formulations which is
error-prone and difficult to reuse, these models can be written in a
higher-level description that can be compiled for these different types
of simulations.

Several tools already exist for defining and simulating CRNs.
BioSimulators.org (Shaikh et al. 2022), a registry of simulation tools,
list at least 15 Python software including COPASI (Hoops et al. 2006),
Tellurium (Choi et al. 2018) and PySB (Lopez et al. 2013). COPASI is a
standalone software with a GUI for defining and simulating models. It's
widely used for its user-friendly interface and comprehensive features.
There are also python bindings and interfaces for COPASI to allow
advanced scripting. Tellurium is a Python-based modeling environment,
however, it uses a C++ library called libRoadRunner in the backend to
simulate models. PySB is a Python library that created a DSL using
standard Python to define models which are then compiled to ODE using a
Perl library called BioNetGen (L. A. Harris et al. 2016).

One limitation of these tools is that they are not extensible from
Python, as they are not fully Python packages but wrap libraries in
other languages to do the simulation. While they enable model definition
and simulation control via Python scripts, they don't fully leverage
Python's extensive package ecosystem. For example, in COPASI, users are
restricted to predefined distributions for Random Parameter Scans and
cannot utilize the diverse distributions available in
\texttt{scipy.stats}. Similarly, Tellurium doesn't allow the use of
Python solvers, as adding new integrators requires C++ implementation.

Another challenge is the way models are written. Many tools use a Domain
Specific Language (DSL) for coding, and support the System Biology
Markup Language (SBML) as an exchange format (Hucka et al., n.d.), as
direct SBML coding is impractical. A DSL allows reuse of code in
different programming environments, but it will not be recognized by
default in Integrated Development Environments (IDEs). Therefore DSLs
cannot provide the development experience such as syntax highlighting,
code completion, refactoring, and static analysis, unless such support
is specifically developed. Tellurium uses a Domain Specific Language
(DSL) called Antimony (Smith et al. 2009), which can be translated to
SBML. An extension for Visual Studio Code was developed, but its
maintenance could be demanding task for the systems biology community.
PySB, using Python's dynamic nature, created a DSL within Python. By
default, it uses global state to create species and parameters, without
assigning them to Python variables or adding them explicitly to the
model, but this approach is not fully compatible with IDEs, affecting
the development experience.

To overcome these limitations, we developed poincaré and SimBio,
open-source Python packages for defining and simulating systems.
Poincaré allows defining differential equation systems, while SimBio
builds on it for defining reaction networks. They are focused on
providing an ergonomic experience to end-users by integrating well with
IDEs and static analysis tools through the use of standard modern Python
syntax. Since they are coded in Python, every part from model definition
to simulation can be extended from Python syntax. Being the first-ever
pure Python packages for systems modelling, they offer extensive
extensibility, from simple tasks like reusing integrators defined in
other packages, to complex ones like altering the compilation process to
leverage some structure in the equations. For example, using a for-loop
in the compiled equations could improve the runtime performance if there
is some repetitive structure in the system, as happens in spatial
modelling. The models built using these packages can be introspected to
create other representations, such as graphs connecting species and/or
reactions, or tables with parameters or equations. Furthermore, they
have a modular architecture with a clear separation of concerns, making
it easier to maintain or to contribute new code, which is beneficial for
developers and maintainers. We showcased the reliability of these tools
by benchmarking them against the simulation results from other tools. We
also highlighted the substantial performance improvements our tools
offer, as this is crucial for construction and simulation of models of
whole cells and organisms, which necessitate the simulation of
significantly large-scale models.

\hypertarget{results}{%
\section{Results}\label{results}}

Modular code architecture makes code reusable, extensible, and easier to
maintain. Therefore, we split the code to define and simulate reaction
systems into three Python packages: symbolite, to create symbolic
expressions; poincaré, to define dynamical systems; and simbio, to
define reaction systems and interface with systems biology standards
such as SBML. These are pure Python packages with standard dependencies
from the PyData scientific stack such as NumPy (C. R. Harris et al.
2020) and pandas (McKinney 2010). They are published in PyPI (the Python
Package Index), where links to the source code and documentation hosted
in GitHub can be found, and can be easily installed with
\texttt{pip\ install\ \textless{}package\_name\textgreater{}}.

Symbolite is a lightweight symbolics package to create algebraic
mathematical expressions. Symbolite expressions can be inspected and
compiled to various backends. Currently, we have implementations for
NumPy (C. R. Harris et al. 2020); Numba (Lam, Pitrou, and Seibert 2015),
a Just-in-Time (JIT) compiler to LLVM; SymPy (Meurer et al. 2017), a
library for symbolic mathematics; and JAX (Bradbury et al. 2018), a
library that support automatic differentiation and compilation to GPUs
and TPUs. Symbolite is designed to facilitate the easy integration of
new backends.

\hypertarget{versatile-modelling-and-simulation-of-dynamical-systems-with-poincaruxe9}{%
\subsection{Versatile modelling and simulation of dynamical systems with
Poincaré}\label{versatile-modelling-and-simulation-of-dynamical-systems-with-poincaruxe9}}

\begin{figure}

\begin{minipage}[t]{\linewidth}

{\centering

\begin{Shaded}
\begin{Highlighting}[]
\ImportTok{from}\NormalTok{ poincare }\ImportTok{import}\NormalTok{ Derivative, System, Variable, initial}
\end{Highlighting}
\end{Shaded}

}

\end{minipage}%
\newline
\begin{minipage}[t]{\linewidth}

{\centering

\begin{figure}

\begin{minipage}[c]{0.33\linewidth}

{\centering

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{class}\NormalTok{ Decay(System):}
\NormalTok{    x: Variable }\OperatorTok{=}\NormalTok{ initial(default}\OperatorTok{=}\DecValTok{1}\NormalTok{)}
\NormalTok{    eq }\OperatorTok{=}\NormalTok{ x.derive() }\OperatorTok{\textless{}\textless{}} \OperatorTok{{-}}\NormalTok{x}
\end{Highlighting}
\end{Shaded}

}

\end{minipage}%
%
\begin{minipage}[c]{0.33\linewidth}

{\centering

\[
\frac{dx}{dt} = -x
\]

}

\end{minipage}%
%
\begin{minipage}[c]{0.33\linewidth}

{\centering

\[
x(0) = 1
\]

}

\end{minipage}%

\caption{First-order system of an exponential decay. The variable
\texttt{x} stores the initial condition for \(x\), and the variable
\texttt{eq} stores the rate equation for \(x\).}

\end{figure}

}

\subcaption{\label{}}
\end{minipage}%
\newline
\begin{minipage}[t]{\linewidth}

{\centering

\begin{figure}

\begin{minipage}[c]{0.33\linewidth}

{\centering

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{class}\NormalTok{ Oscillator(System):}
\NormalTok{    x: Variable }\OperatorTok{=}\NormalTok{ initial(default}\OperatorTok{=}\DecValTok{1}\NormalTok{)}
\NormalTok{    v: Derivative }\OperatorTok{=}\NormalTok{ x.derive(initial}\OperatorTok{=}\DecValTok{0}\NormalTok{)}
\NormalTok{    eq }\OperatorTok{=}\NormalTok{ v.derive() }\OperatorTok{\textless{}\textless{}} \OperatorTok{{-}}\NormalTok{x}
\end{Highlighting}
\end{Shaded}

}

\end{minipage}%
%
\begin{minipage}[c]{0.33\linewidth}

{\centering

\[
\frac{d^2x}{dt^2} = -x
\]

}

\end{minipage}%
%
\begin{minipage}[c]{0.33\linewidth}

{\centering

\[
\begin{cases}
    x(0) &= 1 \\
    \frac{dx}{dt}(0) &= 0
\end{cases}
\]

}

\end{minipage}%

\caption{Second-order system of an harmonic oscillator, where the
variable \texttt{v} stores the derivative \(\frac{dx}{dt}\) and the rate
equation is specified for the derivative \texttt{v}.}

\end{figure}

}

\subcaption{\label{}}
\end{minipage}%
\newline
\begin{minipage}[t]{\linewidth}

{\centering

\begin{figure}

\begin{minipage}[c]{0.33\linewidth}

{\centering

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{class}\NormalTok{ BigModel(System):}
\NormalTok{    x: Variable }\OperatorTok{=}\NormalTok{ initial(default}\OperatorTok{=}\DecValTok{1}\NormalTok{)}
\NormalTok{    linked }\OperatorTok{=}\NormalTok{ Decay(x}\OperatorTok{=}\NormalTok{x)}
\NormalTok{    independent }\OperatorTok{=}\NormalTok{ Decay(x}\OperatorTok{=}\DecValTok{2}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

}

\end{minipage}%
%
\begin{minipage}[c]{0.33\linewidth}

{\centering

\[
\begin{cases}
    \frac{dx}{dt} = -x \\
    \frac{dy}{dt} = -y
\end{cases}
\]

}

\end{minipage}%
%
\begin{minipage}[c]{0.33\linewidth}

{\centering

\[
\begin{cases}
    x(0) &= 1 \\
    y(0) &= 2
\end{cases}
\]

}

\end{minipage}%

\caption{First-order system of two exponential decays by composition of
the \texttt{Decay} system. The subsystem \texttt{linked} has a reference
to the outer variable \texttt{x}, while the subsystem
\texttt{independent} defines a new variable with initial condition §2\$,
which on the corresponding mathematical expression was named \(y\).}

\end{figure}

}

\subcaption{\label{}}
\end{minipage}%

\caption{\label{fig-poincare}Code and corresponding mathematical
expressions for different systems.}

\end{figure}

Poincare is a package to define and simulate dynamical systems. It
provides a \texttt{System} class, where one can define
\texttt{Constant}s, \texttt{Parameter}s, \texttt{Variable}s, and create
equations linking a variable's derivative with an expression
(Figure~\ref{fig-first-order}). It also allows to define higher-order
systems by assigning an initial condition to a \texttt{Derivative}
(Figure~\ref{fig-second-order}).

Utilizing classes for system definition offers several advantages:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The variable name to which a component is assigned can be
  automatically saved in the component for introspection (i.e.,
  \texttt{Oscilator.x.name\ ==\ "x"}),
\item
  It provides a namespace such that allows to easily define of multiple
  independent models in the same script,
\item
  It allows IDEs to provide autocomplete and refactoring capabilities
  (\texttt{Oscillator.\textless{}TAB\textgreater{}} shows \texttt{x},
  \texttt{v} and \texttt{eq}),
\item
  It allows creation of instances which can be composed into a bigger
  model (Figure~\ref{fig-composition}).
\end{enumerate}

For this last point, IDEs that support \texttt{dataclass\_transform} (De
Bonte and Traut 2021) can provide a tooltip with the expected signature
(Figure~\ref{fig-composition}). This requires the use of type
annotations which play a more significant role in static type checking,
as they can help to identify errors before running the code. For
instance, to parameterize the initial conditions of variables we have to
use a \texttt{Constant}. If we try to use a \texttt{Parameter}, which
could be a time-dependent expression, it is flagged as a type error
(Figure~\ref{fig-ide}).

\begin{figure}

\begin{minipage}[b]{0.50\linewidth}

{\centering

\includegraphics{src/ide/ide1.pdf}

}

\end{minipage}%
%
\begin{minipage}[b]{0.50\linewidth}

{\centering

\includegraphics{src/ide/ide2.pdf}

}

\end{minipage}%

\caption{\label{fig-ide}Screenshots of Visual Studio Code showing
tooltips and highlighted type errors. \texttt{Constant} are assigned
with \texttt{assign(...,\ constant=True)} and can be used to link
\texttt{Variable}s initial conditions, while trying to use a
\texttt{Parameter} instead is flagged as type errors (red underlining).
The IDE automatically recognizes \texttt{e} as an \texttt{Equation}, and
provides autocompletion of the variables. A tooltip is shown when
composing models, which show the expected variables and their default
values. The IDE highlights wrong names (\texttt{z} is not a name in
\texttt{Model}) and mismatched types (\texttt{x} is \texttt{Variable}
and \texttt{a} must be a number or a \texttt{Constant})}

\end{figure}

To simulate a system, we created a \texttt{Simulator} instance
(Figure~\ref{fig-sim}), which compiles a given system and interfaces
with solvers. By default, it creates a first-order ordinary differential
equation (ODE) system using \texttt{numpy} as a backend. This can be
easily switched to other solvers. The Simulator wraps the output in a
\texttt{pandas.DataFrame}, which can be easily plotted with the standard
\texttt{plot} method.

\begin{figure}

\begin{minipage}[t]{\linewidth}

{\centering

\begin{Shaded}
\begin{Highlighting}[]
\ImportTok{import}\NormalTok{ numpy }\ImportTok{as}\NormalTok{ np}
\ImportTok{from}\NormalTok{ poincare }\ImportTok{import}\NormalTok{ Simulator}

\NormalTok{sim }\OperatorTok{=}\NormalTok{ Simulator(Oscillator)}
\NormalTok{df }\OperatorTok{=}\NormalTok{ sim.solve(save\_at}\OperatorTok{=}\NormalTok{np.linspace(}\DecValTok{0}\NormalTok{, }\DecValTok{5}\NormalTok{, }\DecValTok{101}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

}

\end{minipage}%
\newline
\begin{minipage}[t]{0.50\linewidth}

{\centering

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{df.head()}
\end{Highlighting}
\end{Shaded}

\begin{longtable}[]{@{}lll@{}}
\toprule\noalign{}
& x & v \\
time & & \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
0.00 & 1.000000 & 0.000000 \\
0.05 & 0.998750 & -0.049979 \\
0.10 & 0.995005 & -0.099835 \\
0.15 & 0.988772 & -0.149440 \\
0.20 & 0.980069 & -0.198672 \\
\end{longtable}

}

\end{minipage}%
%
\begin{minipage}[t]{0.50\linewidth}

{\centering

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{df.plot()}\OperatorTok{;}
\end{Highlighting}
\end{Shaded}

\begin{figure}[H]

{\centering \includegraphics{article_files/figure-latex/cell-9-output-1.pdf}

}

\end{figure}

}

\end{minipage}%

\caption{\label{fig-sim}Simulation of the \texttt{Oscillator} system
from Figure~\ref{fig-second-order}. The output is a
\texttt{pandas.DataFrame} with a column for each variable and the time
as index. It is inspected and plotted with the \texttt{pandas} methods.}

\end{figure}

Switching backends to \texttt{"numba"} can improve model runtime on a
factor up to x30 or more for big or long running models, however incurs
in a compile time penalty for the first run which must be taken into
account.

\hypertarget{extensible-definition-of-reaction-networks-using-simbio}{%
\subsection{Extensible definition of reaction networks using
SimBio}\label{extensible-definition-of-reaction-networks-using-simbio}}

\begin{figure}

\begin{minipage}[t]{\linewidth}

{\centering

\begin{Shaded}
\begin{Highlighting}[]
\ImportTok{from}\NormalTok{ simbio }\ImportTok{import}\NormalTok{ Compartment, MassAction, Species, RateLaw, initial}
\end{Highlighting}
\end{Shaded}

}

\end{minipage}%
\newline
\begin{minipage}[t]{\linewidth}

{\centering

\begin{figure}

\begin{minipage}[c]{0.50\linewidth}

{\centering

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{class}\NormalTok{ Model(Compartment):}
    \CommentTok{"""2A {-}\textgreater{} B"""}

\NormalTok{    A: Species }\OperatorTok{=}\NormalTok{ initial(default}\OperatorTok{=}\DecValTok{1}\NormalTok{)}
\NormalTok{    B: Species }\OperatorTok{=}\NormalTok{ initial(default}\OperatorTok{=}\DecValTok{0}\NormalTok{)}
\NormalTok{    r }\OperatorTok{=}\NormalTok{ RateLaw(}
\NormalTok{        reactants}\OperatorTok{=}\NormalTok{[}\DecValTok{2} \OperatorTok{*}\NormalTok{ A],}
\NormalTok{        products}\OperatorTok{=}\NormalTok{[B],}
\NormalTok{        rate\_law}\OperatorTok{=}\DecValTok{1}\NormalTok{,}
\NormalTok{    )}
\end{Highlighting}
\end{Shaded}

}

\end{minipage}%
%
\begin{minipage}[c]{0.50\linewidth}

{\centering

\[
\begin{align}
&\begin{cases}
    \frac{dA}{dt} = -2 \\
    \frac{dB}{dt} = +1
\end{cases}
\\\\
&\begin{cases}
    A(0) &= 1 \\
    B(0) &= 0
\end{cases}
\end{align}
\]

}

\end{minipage}%

\end{figure}

}

\end{minipage}%
\newline
\begin{minipage}[t]{\linewidth}

{\centering

\begin{figure}

\begin{minipage}[c]{0.50\linewidth}

{\centering

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{class}\NormalTok{ Model(Compartment):}
    \CommentTok{"""2A {-}\textgreater{} B"""}

\NormalTok{    A: Species }\OperatorTok{=}\NormalTok{ initial(default}\OperatorTok{=}\DecValTok{1}\NormalTok{)}
\NormalTok{    B: Species }\OperatorTok{=}\NormalTok{ initial(default}\OperatorTok{=}\DecValTok{0}\NormalTok{)}
\NormalTok{    r }\OperatorTok{=}\NormalTok{ MassAction(}
\NormalTok{        reactants}\OperatorTok{=}\NormalTok{[}\DecValTok{2} \OperatorTok{*}\NormalTok{ A],}
\NormalTok{        products}\OperatorTok{=}\NormalTok{[B],}
\NormalTok{        rate}\OperatorTok{=}\DecValTok{1}\NormalTok{,}
\NormalTok{    )}
\end{Highlighting}
\end{Shaded}

}

\end{minipage}%
%
\begin{minipage}[c]{0.50\linewidth}

{\centering

\[
\begin{align}
&\begin{cases}
    \frac{dA}{dt} = -2 A^2 \\
    \frac{dB}{dt} = +1 A^2
\end{cases}
\\\\
&\begin{cases}
    A(0) &= 1 \\
    B(0) &= 0
\end{cases}
\end{align}
\]

}

\end{minipage}%

\end{figure}

}

\end{minipage}%

\caption{\label{fig-simbio}A reaction system for species \(A\) and \(B\)
with initial conditions \(1\) and \(0\), respectively. A single reaction
transforming \(2A\) into \(B\) is saved in variable \texttt{r}. The rate
\(1\) is specified directly for \texttt{RateLaw}, and is proportional to
the reactants for \texttt{MassAction}.}

\end{figure}

For the reaction networks, our focus is on first-order differential
equations that describe the rate of change of species. SimBio simplifies
the definition of these network models by introducing \texttt{Species},
which consists of a \texttt{poincare.Variable} and a stoichiometric
number, and \texttt{RateLaw}, a construct that converts reactants into
products taking into account the stoichiometry
(Figure~\ref{fig-simbio}). Additionally, SimBio features
\texttt{MassAction}, a subclass of \texttt{RateLaw}, which intuitively
incorporates reactants into the rate law (Figure~\ref{fig-simbio}).

Several commonly used reactions are predefined as \texttt{MassAction}
subclasses, such as \texttt{MichaelisMenten}
(\(S + E \leftrightarrow ES \rightarrow P + E\)) and its approximate
form without the intermediate species \(ES\), and it is also simple to
implement your own as subclasses of \texttt{RateLaw} or
\texttt{MassAction}. Additionally, SimBio supports importing from and
exporting to SBML, and downloading them directly from BioModels
(Malik-Sheriff et al. 2020) (Figure~\ref{fig-simbio-io}).

\begin{figure}

\begin{minipage}[t]{\linewidth}

{\centering

\begin{Shaded}
\begin{Highlighting}[]
\ImportTok{from}\NormalTok{ simbio.io }\ImportTok{import}\NormalTok{ biomodels, sbml}

\NormalTok{model }\OperatorTok{=}\NormalTok{ sbml.load(}\StringTok{"repressilator.sbml"}\NormalTok{)  }\CommentTok{\# from existing SBML file}
\NormalTok{model }\OperatorTok{=}\NormalTok{ biomodels.load\_model(}\StringTok{"BIOMD12"}\NormalTok{)  }\CommentTok{\# from BioModels}
\NormalTok{model}
\end{Highlighting}
\end{Shaded}

Elowitz2000 - Repressilator

type

total

names

variables

6

PX, PY, PZ, X, Y, Z

parameters

17

cell, beta, alpha0, alpha, eff, n, KM, tau\_mRNA, tau\_prot, ...

equations

12

Reaction1, Reaction2, Reaction3, Reaction4, Reaction5, ...

}

\end{minipage}%

\caption{\label{fig-simbio-io}Creation of a model from a local SBML file
or one uploaded to BioModels.}

\end{figure}

\hypertarget{reproducibility-and-performance}{%
\subsection{Reproducibility and
performance}\label{reproducibility-and-performance}}

\begin{figure}

{\centering \includegraphics{src/performance/figures/performance.png}

}

\caption{\label{fig-runtime}Performance of different softwares to solve
models from the curated section of BioModels. (left) Run time for the
model BIOMD3 as a function of the number of output points. (right) Run
time for different models for 300 output points, using the geometric
mean of the different softwares to order them.}

\end{figure}

To evaluate SimBio's reproducibility, we analyzed curated SBML models
from BioModels (Malik-Sheriff et al. 2020). Among the first 100 curated
models from BioModels we selected 60 which did not contains events, as
SimBio doesn't support events yet. We simulated the selected models with
COPASI and used the simulated results as ground truth, and demonstrated
that SimBio reproduces the results with minor numerical differences
depending on the solver tolerance.

For performance testing, we ran simulations using COPASI,
Tellurium/RoadRunner, and SimBio. Within SimBio, both NumPy and Numba
backends were considered. The LSODA solver was used for COPASI and
SimBio, while for Tellurium the comparable CVODE solver was used. In all
cases, we used relative and absolute tolerances of \(10^{-6}\). We
measured three simulation stages: loading, the initial (cold) run, and
subsequent (warm) runs for each model.

For COPASI and Tellurium/RoadRunner, we noted that its runtime depended
on the number of evaluation points, something that does not seem to
happen with SimBio (Figure~\ref{fig-runtime}, left). While SimBio's
NumPy backend is slower than both COPASI and RoadRunner, we obtained an
order of magnitude speed-up using the numba backend putting it on par
with them. A user might have to consider the trade-off between
compilation and run times, as the compilation of the right-hand-side
(RHS) code might take longer than the runtime itself, and not be worth
it for running the model only once. Another speed-up in the runtime can
be had by switching the LSODA \texttt{scipy} solver for a more efficient
\texttt{numbalsoda} implementation, which avoids the Python interpreter
between each of the integration steps. This last combination beats all
other methods, which is also true for the other models we tested
(Figure~\ref{fig-runtime}, right).

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

In this article, we introduced a suite of Python packages we developed
for defining and simulating dynamical systems and reaction networks.
These packages are deeply integrated with Integrated Development
Environments (IDEs), enabling code analysis tools to identify errors
prior to execution and assist in refactoring and code completion. We
adopted standard modern Python syntax to ensure seamless IDE
integration, supported by the extensive Python community.

Our approach differs from previous tools in that both the model
definition and its compilation into an Ordinary Differential Equation
(ODE) function are entirely Python-based. This approach simplifies the
development of various simulation methods, including performance
enhancements that exploit specific model structures. Importantly, being
Python-based does not compromise performance compared to C/C++ tools, as
the ODE functions can be Just-In-Time (JIT) compiled using Numba.

The inclusion of SBML support facilitates the effortless reuse of models
created by the systems biology community, along with the vast collection
of public models hosted in the BioModels repository. The modular
architecture of these packages facilitates their reuse, enhancement, and
extension by the wider Python community. For instance, an individual
from outside the systems biology field could contribute a stochastic
integrator to poincaré, which would then be available in SimBio. This
clear separation of concerns also makes the packages more
comprehensible, lowering the barrier for contributing improvements or
new features. Such an architecture ensures their maintainability and
ongoing development well into the future.

\hypertarget{references}{%
\section{References}\label{references}}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-jax2018github}{}}%
Bradbury, James, Roy Frostig, Peter Hawkins, Matthew James Johnson,
Chris Leary, Dougal Maclaurin, George Necula, et al. 2018. {``{JAX}:
Composable Transformations of {Python}+{NumPy} Programs.''}

\leavevmode\vadjust pre{\hypertarget{ref-choiTelluriumExtensiblePythonbased2018}{}}%
Choi, Kiri, J Kyle Medley, Matthias König, Kaylene Stocking, Lucian
Smith, Stanley Gu, and Herbert M Sauro. 2018. {``Tellurium: {An}
Extensible Python-Based Modeling Environment for Systems and Synthetic
Biology.''} \emph{Bio Systems} 171 (September): 74--79.
\url{https://doi.org/10.1016/j.biosystems.2018.07.006}.

\leavevmode\vadjust pre{\hypertarget{ref-debontePEP681Data2021}{}}%
De Bonte, Erik, and Eric Traut. 2021. {``{PEP} 681 {\textendash} {Data
Class Transforms}.''} \emph{Python Enhancement Proposals}.
https://peps.python.org/pep-0681/.

\leavevmode\vadjust pre{\hypertarget{ref-harrisArrayProgrammingNumPy2020}{}}%
Harris, Charles R., K. Jarrod Millman, Stéfan J. van der Walt, Ralf
Gommers, Pauli Virtanen, David Cournapeau, Eric Wieser, et al. 2020.
{``Array Programming with {NumPy}.''} \emph{Nature} 585 (7825): 357--62.
\url{https://doi.org/10.1038/s41586-020-2649-2}.

\leavevmode\vadjust pre{\hypertarget{ref-harrisBioNetGenAdvancesRulebased2016}{}}%
Harris, Leonard A, Justin S Hogg, José-Juan Tapia, John A P Sekar,
Sanjana Gupta, Ilya Korsunsky, Arshi Arora, Dipak Barua, Robert P
Sheehan, and James R Faeder. 2016. {``{BioNetGen} 2.2: Advances in
Rule-Based Modeling.''} \emph{Bioinformatics (Oxford, England)} 32 (21):
3366--68. \url{https://doi.org/10.1093/bioinformatics/btw469}.

\leavevmode\vadjust pre{\hypertarget{ref-hoopsCOPASICOmplexPAthway2006}{}}%
Hoops, Stefan, Sven Sahle, Ralph Gauges, Christine Lee, Jürgen Pahle,
Natalia Simus, Mudita Singhal, Liang Xu, Pedro Mendes, and Ursula
Kummer. 2006. {``{COPASI}{\textemdash}a {COmplex PAthway SImulator}.''}
\emph{Bioinformatics} 22 (24): 3067--74.
\url{https://doi.org/10.1093/bioinformatics/btl485}.

\leavevmode\vadjust pre{\hypertarget{ref-huckaSBMLL3V2}{}}%
Hucka, Michael, Frank Bergmann, Stefan Hoops, Sarah M Keating, Nicolas
Le Novère, Chris J Myers, Brett G Olivier, et al. n.d. {``{SBML
L3V2}.''}

\leavevmode\vadjust pre{\hypertarget{ref-lamNumbaLLVMbasedPython2015}{}}%
Lam, Siu Kwan, Antoine Pitrou, and Stanley Seibert. 2015. {``Numba: A
{LLVM-based Python JIT} Compiler.''} In \emph{Proceedings of the {Second
Workshop} on the {LLVM Compiler Infrastructure} in {HPC}}, 1--6. {LLVM}
'15. {New York, NY, USA}: {Association for Computing Machinery}.
\url{https://doi.org/10.1145/2833157.2833162}.

\leavevmode\vadjust pre{\hypertarget{ref-lopezProgrammingBiologicalModels2013}{}}%
Lopez, Carlos F, Jeremy L Muhlich, John A Bachman, and Peter K Sorger.
2013. {``Programming Biological Models in {Python} Using {PySB}.''}
\emph{Molecular Systems Biology} 9 (January): 646.
\url{https://doi.org/10.1038/msb.2013.1}.

\leavevmode\vadjust pre{\hypertarget{ref-malik-sheriffBioModels15Years2020}{}}%
Malik-Sheriff, Rahuman S, Mihai Glont, Tung V N Nguyen, Krishna Tiwari,
Matthew G Roberts, Ashley Xavier, Manh T Vu, et al. 2020.
{``{BioModels}{\textemdash}15 Years of Sharing Computational Models in
Life Science.''} \emph{Nucleic Acids Research} 48 (D1): D407--15.
\url{https://doi.org/10.1093/nar/gkz1055}.

\leavevmode\vadjust pre{\hypertarget{ref-mckinneyDataStructuresStatistical2010}{}}%
McKinney, Wes. 2010. {``Data {Structures} for {Statistical Computing} in
{Python}.''} In \emph{Python in {Science Conference}}, 56--61. {Austin,
Texas}. \url{https://doi.org/10.25080/Majora-92bf1922-00a}.

\leavevmode\vadjust pre{\hypertarget{ref-meurerSymPySymbolicComputing2017}{}}%
Meurer, Aaron, Christopher P. Smith, Mateusz Paprocki, Ondřej Čertík,
Sergey B. Kirpichev, Matthew Rocklin, AmiT Kumar, et al. 2017.
{``{SymPy}: Symbolic Computing in {Python}.''} \emph{PeerJ Computer
Science} 3 (January): e103. \url{https://doi.org/10.7717/peerj-cs.103}.

\leavevmode\vadjust pre{\hypertarget{ref-shaikhBioSimulatorsCentralRegistry2022}{}}%
Shaikh, Bilal, Lucian P Smith, Dan Vasilescu, Gnaneswara Marupilla,
Michael Wilson, Eran Agmon, Henry Agnew, et al. 2022.
{``{BioSimulators}: A Central Registry of Simulation Engines and
Services for Recommending Specific Tools.''} \emph{Nucleic Acids
Research} 50 (W1): W108--14. \url{https://doi.org/10.1093/nar/gkac331}.

\leavevmode\vadjust pre{\hypertarget{ref-smithAntimonyModularModel2009}{}}%
Smith, Lucian P., Frank T. Bergmann, Deepak Chandran, and Herbert M.
Sauro. 2009. {``Antimony: A Modular Model Definition Language.''}
\emph{Bioinformatics} 25 (18): 2452--54.
\url{https://doi.org/10.1093/bioinformatics/btp401}.

\end{CSLReferences}

\end{document}

### 2024-01-08 article.tex
Remove figure environments

@@
 
-\begin{figure}
-
 \begin{minipage}[c]{0.33\linewidth}
@@
 
-\caption{First-order system of an exponential decay. The variable
+\subcaption{First-order system of an exponential decay. The variable
 \texttt{x} stores the initial condition for \(x\), and the variable
@@
 
-\end{figure}
-
 }
@@
 {\centering
-
-\begin{figure}
 
@@
 
-\caption{Second-order system of an harmonic oscillator, where the
+\subcaption{Second-order system of an harmonic oscillator, where the
 variable \texttt{v} stores the derivative \(\frac{dx}{dt}\) and the rate
@@
 
-\caption{First-order system of two exponential decays by composition of
+\subcaption{First-order system of two exponential decays by composition of
 the \texttt{Decay} system. The subsystem \texttt{linked} has a reference
@@
 which on the corresponding mathematical expression was named \(y\).}
-
-\end{figure}
 
@@
 
-\begin{figure}
-
 \begin{minipage}[c]{0.50\linewidth}
@@
 \[
-\begin{align}
-&\begin{cases}
+\begin{cases}
     \frac{dA}{dt} = -2 \\
@@
 \end{cases}
-\\\\
-&\begin{cases}
+\begin{cases}
     A(0) &= 1 \\
@@
 \end{cases}
-\end{align}
 \]
@@
 \end{minipage}%
-
-\end{figure}
 
@@
 \[
-\begin{align}
-&\begin{cases}
+\begin{cases}
     \frac{dA}{dt} = -2 A^2 \\

### 2024-01-08 article.tex
Remove empty subcaptions
Fix minipage widths
Fix text error
Move figure code to external files

@@
 
-\begin{figure}
-
-\begin{minipage}[t]{\linewidth}
-
-{\centering
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\ImportTok{from}\NormalTok{ poincare }\ImportTok{import}\NormalTok{ Derivative, System, Variable, initial}
-\end{Highlighting}
-\end{Shaded}
-
-}
-
-\end{minipage}%
-\newline
-\begin{minipage}[t]{\linewidth}
-
-{\centering
-
-\begin{minipage}[c]{0.33\linewidth}
-
-{\centering
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{class}\NormalTok{ Decay(System):}
-\NormalTok{    x: Variable }\OperatorTok{=}\NormalTok{ initial(default}\OperatorTok{=}\DecValTok{1}\NormalTok{)}
-\NormalTok{    eq }\OperatorTok{=}\NormalTok{ x.derive() }\OperatorTok{\textless{}\textless{}} \OperatorTok{{-}}\NormalTok{x}
-\end{Highlighting}
-\end{Shaded}
-
-}
-
-\end{minipage}%
-%
-\begin{minipage}[c]{0.33\linewidth}
-
-{\centering
-
-\[
-\frac{dx}{dt} = -x
-\]
-
-}
-
-\end{minipage}%
-%
-\begin{minipage}[c]{0.33\linewidth}
-
-{\centering
-
-\[
-x(0) = 1
-\]
-
-}
-
-\end{minipage}%
-
-\subcaption{First-order system of an exponential decay. The variable
-\texttt{x} stores the initial condition for \(x\), and the variable
-\texttt{eq} stores the rate equation for \(x\).}
-
-}
-
-\subcaption{\label{}}
-\end{minipage}%
-\newline
-\begin{minipage}[t]{\linewidth}
-
-{\centering
-
-\begin{minipage}[c]{0.33\linewidth}
-
-{\centering
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{class}\NormalTok{ Oscillator(System):}
-\NormalTok{    x: Variable }\OperatorTok{=}\NormalTok{ initial(default}\OperatorTok{=}\DecValTok{1}\NormalTok{)}
-\NormalTok{    v: Derivative }\OperatorTok{=}\NormalTok{ x.derive(initial}\OperatorTok{=}\DecValTok{0}\NormalTok{)}
-\NormalTok{    eq }\OperatorTok{=}\NormalTok{ v.derive() }\OperatorTok{\textless{}\textless{}} \OperatorTok{{-}}\NormalTok{x}
-\end{Highlighting}
-\end{Shaded}
-
-}
-
-\end{minipage}%
-%
-\begin{minipage}[c]{0.33\linewidth}
-
-{\centering
-
-\[
-\frac{d^2x}{dt^2} = -x
-\]
-
-}
-
-\end{minipage}%
-%
-\begin{minipage}[c]{0.33\linewidth}
-
-{\centering
-
-\[
-\begin{cases}
-    x(0) &= 1 \\
-    \frac{dx}{dt}(0) &= 0
-\end{cases}
-\]
-
-}
-
-\end{minipage}%
-
-\subcaption{Second-order system of an harmonic oscillator, where the
-variable \texttt{v} stores the derivative \(\frac{dx}{dt}\) and the rate
-equation is specified for the derivative \texttt{v}.}
-
-}
-
-\subcaption{\label{}}
-\end{minipage}%
-\newline
-\begin{minipage}[t]{\linewidth}
-
-{\centering
-
-\begin{minipage}[c]{0.33\linewidth}
-
-{\centering
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{class}\NormalTok{ BigModel(System):}
-\NormalTok{    x: Variable }\OperatorTok{=}\NormalTok{ initial(default}\OperatorTok{=}\DecValTok{1}\NormalTok{)}
-\NormalTok{    linked }\OperatorTok{=}\NormalTok{ Decay(x}\OperatorTok{=}\NormalTok{x)}
-\NormalTok{    independent }\OperatorTok{=}\NormalTok{ Decay(x}\OperatorTok{=}\DecValTok{2}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-}
-
-\end{minipage}%
-%
-\begin{minipage}[c]{0.33\linewidth}
-
-{\centering
-
-\[
-\begin{cases}
-    \frac{dx}{dt} = -x \\
-    \frac{dy}{dt} = -y
-\end{cases}
-\]
-
-}
-
-\end{minipage}%
-%
-\begin{minipage}[c]{0.33\linewidth}
-
-{\centering
-
-\[
-\begin{cases}
-    x(0) &= 1 \\
-    y(0) &= 2
-\end{cases}
-\]
-
-}
-
-\end{minipage}%
-
-\subcaption{First-order system of two exponential decays by composition of
-the \texttt{Decay} system. The subsystem \texttt{linked} has a reference
-to the outer variable \texttt{x}, while the subsystem
-\texttt{independent} defines a new variable with initial condition §2\$,
-which on the corresponding mathematical expression was named \(y\).}
-
-}
-
-\subcaption{\label{}}
-\end{minipage}%
-
-\caption{\label{fig-poincare}Code and corresponding mathematical
-expressions for different systems.}
-
-\end{figure}
+\input{latex/figure1.tex}
 
@@
 
-\begin{figure}
-
-\begin{minipage}[b]{0.50\linewidth}
-
-{\centering
-
-\includegraphics{src/ide/ide1.pdf}
-
-}
-
-\end{minipage}%
-%
-\begin{minipage}[b]{0.50\linewidth}
-
-{\centering
-
-\includegraphics{src/ide/ide2.pdf}
-
-}
-
-\end{minipage}%
-
-\caption{\label{fig-ide}Screenshots of Visual Studio Code showing
-tooltips and highlighted type errors. \texttt{Constant} are assigned
-with \texttt{assign(...,\ constant=True)} and can be used to link
-\texttt{Variable}s initial conditions, while trying to use a
-\texttt{Parameter} instead is flagged as type errors (red underlining).
-The IDE automatically recognizes \texttt{e} as an \texttt{Equation}, and
-provides autocompletion of the variables. A tooltip is shown when
-composing models, which show the expected variables and their default
-values. The IDE highlights wrong names (\texttt{z} is not a name in
-\texttt{Model}) and mismatched types (\texttt{x} is \texttt{Variable}
-and \texttt{a} must be a number or a \texttt{Constant})}
-
-\end{figure}
+\input{latex/figure2.tex}
 
@@
 
-\begin{figure}
-
-\begin{minipage}[t]{\linewidth}
-
-{\centering
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\ImportTok{import}\NormalTok{ numpy }\ImportTok{as}\NormalTok{ np}
-\ImportTok{from}\NormalTok{ poincare }\ImportTok{import}\NormalTok{ Simulator}
-
-\NormalTok{sim }\OperatorTok{=}\NormalTok{ Simulator(Oscillator)}
-\NormalTok{df }\OperatorTok{=}\NormalTok{ sim.solve(save\_at}\OperatorTok{=}\NormalTok{np.linspace(}\DecValTok{0}\NormalTok{, }\DecValTok{5}\NormalTok{, }\DecValTok{101}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-}
-
-\end{minipage}%
-\newline
-\begin{minipage}[t]{0.50\linewidth}
-
-{\centering
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{df.head()}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{longtable}[]{@{}lll@{}}
-\toprule\noalign{}
-& x & v \\
-time & & \\
-\midrule\noalign{}
-\endhead
-\bottomrule\noalign{}
-\endlastfoot
-0.00 & 1.000000 & 0.000000 \\
-0.05 & 0.998750 & -0.049979 \\
-0.10 & 0.995005 & -0.099835 \\
-0.15 & 0.988772 & -0.149440 \\
-0.20 & 0.980069 & -0.198672 \\
-\end{longtable}
-
-}
-
-\end{minipage}%
-%
-\begin{minipage}[t]{0.50\linewidth}
-
-{\centering
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{df.plot()}\OperatorTok{;}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{figure}[H]
-
-{\centering \includegraphics{article_files/figure-latex/cell-9-output-1.pdf}
-
-}
-
-\end{figure}
-
-}
-
-\end{minipage}%
-
-\caption{\label{fig-sim}Simulation of the \texttt{Oscillator} system
-from Figure~\ref{fig-second-order}. The output is a
-\texttt{pandas.DataFrame} with a column for each variable and the time
-as index. It is inspected and plotted with the \texttt{pandas} methods.}
-
-\end{figure}
+\input{latex/figure3.tex}
 
@@
 
-\begin{figure}
-
-\begin{minipage}[t]{\linewidth}
-
-{\centering
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\ImportTok{from}\NormalTok{ simbio }\ImportTok{import}\NormalTok{ Compartment, MassAction, Species, RateLaw, initial}
-\end{Highlighting}
-\end{Shaded}
-
-}
-
-\end{minipage}%
-\newline
-\begin{minipage}[t]{\linewidth}
-
-{\centering
-
-\begin{minipage}[c]{0.50\linewidth}
-
-{\centering
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{class}\NormalTok{ Model(Compartment):}
-    \CommentTok{"""2A {-}\textgreater{} B"""}
-
-\NormalTok{    A: Species }\OperatorTok{=}\NormalTok{ initial(default}\OperatorTok{=}\DecValTok{1}\NormalTok{)}
-\NormalTok{    B: Species }\OperatorTok{=}\NormalTok{ initial(default}\OperatorTok{=}\DecValTok{0}\NormalTok{)}
-\NormalTok{    r }\OperatorTok{=}\NormalTok{ RateLaw(}
-\NormalTok{        reactants}\OperatorTok{=}\NormalTok{[}\DecValTok{2} \OperatorTok{*}\NormalTok{ A],}
-\NormalTok{        products}\OperatorTok{=}\NormalTok{[B],}
-\NormalTok{        rate\_law}\OperatorTok{=}\DecValTok{1}\NormalTok{,}
-\NormalTok{    )}
-\end{Highlighting}
-\end{Shaded}
-
-}
-
-\end{minipage}%
-%
-\begin{minipage}[c]{0.50\linewidth}
-
-{\centering
-
-\[
-\begin{cases}
-    \frac{dA}{dt} = -2 \\
-    \frac{dB}{dt} = +1
-\end{cases}
-\begin{cases}
-    A(0) &= 1 \\
-    B(0) &= 0
-\end{cases}
-\]
-
-}
-
-\end{minipage}%
-
-}
-
-\end{minipage}%
-\newline
-\begin{minipage}[t]{\linewidth}
-
-{\centering
-
-\begin{minipage}[c]{0.50\linewidth}
-
-{\centering
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{class}\NormalTok{ Model(Compartment):}
-    \CommentTok{"""2A {-}\textgreater{} B"""}
-
-\NormalTok{    A: Species }\OperatorTok{=}\NormalTok{ initial(default}\OperatorTok{=}\DecValTok{1}\NormalTok{)}
-\NormalTok{    B: Species }\OperatorTok{=}\NormalTok{ initial(default}\OperatorTok{=}\DecValTok{0}\NormalTok{)}
-\NormalTok{    r }\OperatorTok{=}\NormalTok{ MassAction(}
-\NormalTok{        reactants}\OperatorTok{=}\NormalTok{[}\DecValTok{2} \OperatorTok{*}\NormalTok{ A],}
-\NormalTok{        products}\OperatorTok{=}\NormalTok{[B],}
-\NormalTok{        rate}\OperatorTok{=}\DecValTok{1}\NormalTok{,}
-\NormalTok{    )}
-\end{Highlighting}
-\end{Shaded}
-
-}
-
-\end{minipage}%
-%
-\begin{minipage}[c]{0.50\linewidth}
-
-{\centering
-
-\[
-\begin{cases}
-    \frac{dA}{dt} = -2 A^2 \\
-    \frac{dB}{dt} = +1 A^2
-\end{cases}
-\begin{cases}
-    A(0) &= 1 \\
-    B(0) &= 0
-\end{cases}
-\]
-
-}
-
-\end{minipage}%
-
-}
-
-\end{minipage}%
-
-\caption{\label{fig-simbio}A reaction system for species \(A\) and \(B\)
-with initial conditions \(1\) and \(0\), respectively. A single reaction
-transforming \(2A\) into \(B\) is saved in variable \texttt{r}. The rate
-\(1\) is specified directly for \texttt{RateLaw}, and is proportional to
-the reactants for \texttt{MassAction}.}
-
-\end{figure}
+\input{latex/figure4.tex}
 
@@
 
-\begin{figure}
-
-\begin{minipage}[t]{\linewidth}
-
-{\centering
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\ImportTok{from}\NormalTok{ simbio.io }\ImportTok{import}\NormalTok{ biomodels, sbml}
-
-\NormalTok{model }\OperatorTok{=}\NormalTok{ sbml.load(}\StringTok{"repressilator.sbml"}\NormalTok{)  }\CommentTok{\# from existing SBML file}
-\NormalTok{model }\OperatorTok{=}\NormalTok{ biomodels.load\_model(}\StringTok{"BIOMD12"}\NormalTok{)  }\CommentTok{\# from BioModels}
-\NormalTok{model}
-\end{Highlighting}
-\end{Shaded}
-
-Elowitz2000 - Repressilator
-
-type
-
-total
-
-names
-
-variables
-
-6
-
-PX, PY, PZ, X, Y, Z
-
-parameters
-
-17
-
-cell, beta, alpha0, alpha, eff, n, KM, tau\_mRNA, tau\_prot, ...
-
-equations
-
-12
-
-Reaction1, Reaction2, Reaction3, Reaction4, Reaction5, ...
-
-}
-
-\end{minipage}%
-
-\caption{\label{fig-simbio-io}Creation of a model from a local SBML file
-or one uploaded to BioModels.}
-
-\end{figure}
+\input{latex/figure5.tex}
 
@@
 
-\begin{figure}
-
-{\centering \includegraphics{src/performance/figures/performance.png}
-
-}
-
-\caption{\label{fig-runtime}Performance of different softwares to solve
-models from the curated section of BioModels. (left) Run time for the
-model BIOMD3 as a function of the number of output points. (right) Run
-time for different models for 300 output points, using the geometric
-mean of the different softwares to order them.}
-
-\end{figure}
+\input{latex/figure6.tex}
 

### 2024-01-08 latex/figure2.tex
Move figure code to external files

\begin{figure}[t]

\begin{minipage}[b]{0.50\linewidth}

{\centering

\includegraphics{src/ide/ide1.pdf}

}

\end{minipage}%
%
\begin{minipage}[b]{0.50\linewidth}

{\centering

\includegraphics{src/ide/ide2.pdf}

}

\end{minipage}%

\caption{\label{fig-ide}Screenshots of Visual Studio Code showing
tooltips and highlighted type errors. \texttt{Constant} are assigned
with \texttt{assign(...,\ constant=True)} and can be used to link
\texttt{Variable}s initial conditions, while trying to use a
\texttt{Parameter} instead is flagged as type errors (red underlining).
The IDE automatically recognizes \texttt{e} as an \texttt{Equation}, and
provides autocompletion of the variables. A tooltip is shown when
composing models, which show the expected variables and their default
values. The IDE highlights wrong names (\texttt{z} is not a name in
\texttt{Model}) and mismatched types (\texttt{x} is \texttt{Variable}
and \texttt{a} must be a number or a \texttt{Constant})}

\end{figure}

### 2024-01-08 latex/figure3.tex
Move figure code to external files

\begin{figure}[t]

\begin{minipage}[t]{\linewidth}

{\centering

\begin{Shaded}
\begin{Highlighting}[]
\ImportTok{import}\NormalTok{ numpy }\ImportTok{as}\NormalTok{ np}
\ImportTok{from}\NormalTok{ poincare }\ImportTok{import}\NormalTok{ Simulator}

\NormalTok{sim }\OperatorTok{=}\NormalTok{ Simulator(Oscillator)}
\NormalTok{df }\OperatorTok{=}\NormalTok{ sim.solve(save\_at}\OperatorTok{=}\NormalTok{np.linspace(}\DecValTok{0}\NormalTok{, }\DecValTok{5}\NormalTok{, }\DecValTok{101}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

}

\end{minipage}%
\newline
\begin{minipage}[t]{0.50\linewidth}

{\centering

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{df.head()}
\end{Highlighting}
\end{Shaded}

\begin{longtable}[]{@{}lll@{}}
\toprule\noalign{}
& x & v \\
time & & \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
0.00 & 1.000000 & 0.000000 \\
0.05 & 0.998750 & -0.049979 \\
0.10 & 0.995005 & -0.099835 \\
0.15 & 0.988772 & -0.149440 \\
0.20 & 0.980069 & -0.198672 \\
\end{longtable}

}

\end{minipage}%
%
\begin{minipage}[t]{0.50\linewidth}

{\centering

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{df.plot()}\OperatorTok{;}
\end{Highlighting}
\end{Shaded}

\begin{figure}[H]

{\centering \includegraphics{article_files/figure-latex/cell-9-output-1.pdf}

}

\end{figure}

}

\end{minipage}%

\caption{\label{fig-sim}Simulation of the \texttt{Oscillator} system
from Figure~\ref{fig-second-order}. The output is a
\texttt{pandas.DataFrame} with a column for each variable and the time
as index. It is inspected and plotted with the \texttt{pandas} methods.}

\end{figure}

### 2024-01-08 latex/figure5.tex
Move figure code to external files

\begin{figure}[t]

\begin{minipage}[t]{\linewidth}

{\centering

\begin{Shaded}
\begin{Highlighting}[]
\ImportTok{from}\NormalTok{ simbio.io }\ImportTok{import}\NormalTok{ biomodels, sbml}

\NormalTok{model }\OperatorTok{=}\NormalTok{ sbml.load(}\StringTok{"repressilator.sbml"}\NormalTok{)  }\CommentTok{\# from existing SBML file}
\NormalTok{model }\OperatorTok{=}\NormalTok{ biomodels.load\_model(}\StringTok{"BIOMD12"}\NormalTok{)  }\CommentTok{\# from BioModels}
\NormalTok{model}
\end{Highlighting}
\end{Shaded}

\begin{longtable}{@{}lll@{}}
    \toprule\noalign{}
    Elowitz2000 - Repressilator \\
    type & total & names \\
    \midrule\noalign{}
    \endhead
    \bottomrule\noalign{}
    \endlastfoot
    variables  &  6 & PX, PY, PZ, X, Y, Z \\
    parameters & 17 & cell, beta, alpha0, alpha, eff, n, KM, tau\_mRNA, tau\_prot, ... \\
    equations  & 12 & Reaction1, Reaction2, Reaction3, Reaction4, Reaction5, ... \\
\end{longtable}

}

\end{minipage}%

\caption{\label{fig-simbio-io}Creation of a model from a local SBML file
or one uploaded to BioModels.}

\end{figure}

### 2024-01-08 latex/figure6.tex
Move figure code to external files

\begin{figure}[t]

{\centering \includegraphics{src/performance/figures/performance.png}

}

\caption{\label{fig-runtime}Performance of different softwares to solve
models from the curated section of BioModels. (left) Run time for the
model BIOMD3 as a function of the number of output points. (right) Run
time for different models for 300 output points, using the geometric
mean of the different softwares to order them.}

\end{figure}

### 2024-01-08 article.tex
Change to a4paper
Add affiliations

@@
 \documentclass[
-  letterpaper,
+  a4paper,
   DIV=11,
@@
 
+\usepackage{authblk}
+
 \title{Poincaré and SimBio: a versatile and extensible Python ecosystem
 for modeling systems.}
-\author{Mauro Silberberg \and Rahuman S. Malik-Sheriff \and Henning
-Hermjakob \and Hernán E. Grecco}
+\author[1,2]{Mauro Silberberg}
+\author[3]{Rahuman S. Malik-Sheriff}
+\author[3]{Henning Hermjakob}
+\author[1,2]{Hernán E. Grecco}
+\affil[1]{Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Física. Buenos Aires, Argentina.}
+\affil[2]{CONICET - Universidad de Buenos Aires, Instituto de Física de Buenos Aires (IFIBA). Buenos Aires, Argentina}
+\affil[3]{European Bioinformatics Institute, European Molecular Biology Laboratory (EMBL-EBI), Wellcome Trust Genome Campus, Cambridge, UK}
 \date{}

### 2024-01-08 article.tex
Revert to letterpaper
Move figures around
Small affiliations
Move figures
Change color for code background

@@
 \documentclass[
-  a4paper,
+  letterpaper,
   DIV=11,
@@
 \usepackage{framed}
-\definecolor{shadecolor}{RGB}{241,243,245}
+\definecolor{shadecolor}{RGB}{100,100,100}
 \newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
@@
 \makeatletter
-\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
-\makeatother
-\makeatletter
 \makeatother
@@
 \usepackage{authblk}
+\renewcommand\Affilfont{\small}
 
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, frame hidden, interior hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, enhanced]}{\end{tcolorbox}}\fi
 
@@
 Python solvers, as adding new integrators requires C++ implementation.
+
+\input{latex/figure1.tex}
 
@@
 
+\input{latex/figure2.tex}
+
 Modular code architecture makes code reusable, extensible, and easier to
@@
 
-\input{latex/figure1.tex}
+\input{latex/figure3.tex}
 
@@
 
+\input{latex/figure4.tex}
+
 For this last point, IDEs that support \texttt{dataclass\_transform} (De
@@
 
-\input{latex/figure2.tex}
-
 To simulate a system, we created a \texttt{Simulator} instance
@@
 \texttt{plot} method.
-
-\input{latex/figure3.tex}
 
@@
 SimBio}\label{extensible-definition-of-reaction-networks-using-simbio}}
-
-\input{latex/figure4.tex}
 

### 2024-01-08 latex/figure3.tex
Change color for code background
Update syntax highlighting

@@
 \begin{Highlighting}[]
-\ImportTok{import}\NormalTok{ numpy }\ImportTok{as}\NormalTok{ np}
-\ImportTok{from}\NormalTok{ poincare }\ImportTok{import}\NormalTok{ Simulator}
+\KeywordTok{import} \ClassTok{numpy} \KeywordTok{as} \ClassTok{np}
+\KeywordTok{from} \ClassTok{poincare} \KeywordTok{import} \ClassTok{Simulator}
 
-\NormalTok{sim }\OperatorTok{=}\NormalTok{ Simulator(Oscillator)}
-\NormalTok{df }\OperatorTok{=}\NormalTok{ sim.solve(save\_at}\OperatorTok{=}\NormalTok{np.linspace(}\DecValTok{0}\NormalTok{, }\DecValTok{5}\NormalTok{, }\DecValTok{101}\NormalTok{))}
+\VariableTok{sim} \OperatorTok{=} \ClassTok{Simulator}\KeywordTok{(\ClassTok{Oscillator})}
+\VariableTok{df} \OperatorTok{=} \VariableTok{sim}.\FunctionTok{solve}\KeywordTok{(}\VariableTok{save\_at}\OperatorTok{=}\ClassTok{np}.\FunctionTok{linspace}(\ValueTok{0}, \ValueTok{5}, \ValueTok{101})\KeywordTok{)}
 \end{Highlighting}
@@
 \newline
-\begin{minipage}[t]{0.50\linewidth}
+\begin{minipage}[t]{0.45\linewidth}
 
@@
 \begin{Highlighting}[]
-\NormalTok{df.head()}
+\VariableTok{df}.\FunctionTok{head}\KeywordTok{()}
 \end{Highlighting}
@@
 \end{minipage}%
-%
-\begin{minipage}[t]{0.50\linewidth}
+\hfill
+\begin{minipage}[t]{0.45\linewidth}
 
@@
 \begin{Highlighting}[]
-\NormalTok{df.plot()}\OperatorTok{;}
+\VariableTok{df}.\FunctionTok{plot}\KeywordTok{()}\OperatorTok{;}
 \end{Highlighting}

### 2024-01-08 latex/figure5.tex
Fix table format
Update syntax highlighting

@@
 \begin{Highlighting}[]
-\ImportTok{from}\NormalTok{ simbio.io }\ImportTok{import}\NormalTok{ biomodels, sbml}
+\KeywordTok{from}\ClassTok{ simbio.io }\KeywordTok{import}\ClassTok{ biomodels, sbml}
 
-\NormalTok{model }\OperatorTok{=}\NormalTok{ sbml.load(}\StringTok{"repressilator.sbml"}\NormalTok{)  }\CommentTok{\# from existing SBML file}
-\NormalTok{model }\OperatorTok{=}\NormalTok{ biomodels.load\_model(}\StringTok{"BIOMD12"}\NormalTok{)  }\CommentTok{\# from BioModels}
-\NormalTok{model}
+\VariableTok{model }\OperatorTok{=}\ClassTok{ sbml}.\FunctionTok{load}\KeywordTok{(}\StringTok{"repressilator.sbml"}\KeywordTok{)}  \CommentTok{\# from existing SBML file}
+\VariableTok{model }\OperatorTok{=}\ClassTok{ biomodels}.\FunctionTok{load\_model}\KeywordTok{(}\StringTok{"BIOMD12"}\KeywordTok{)}  \CommentTok{\# from BioModels}
+\VariableTok{model}
 \end{Highlighting}
@@
 \begin{longtable}{@{}lll@{}}
+    \multicolumn{3}{l}{Elowitz2000 - Repressilator} \\
     \toprule\noalign{}
-    Elowitz2000 - Repressilator \\
     type & total & names \\

### 2024-01-10 src/performance/plot.py
Add pdf and svg outputs for performance figure

@@ -64,6 +64,7 @@ if __name__ == "__main__":
 
     match sys.argv[1:]:
         case ["save"]:
-            fig.savefig(path / "figures/performance.png")
+            for format in ["png", "pdf", "svg"]:
+                fig.savefig(path / f"figures/performance.{format}")
         case _:
             plt.show()

### 2024-01-10 latex/figure3.tex
Set code call parentheses as keyword color

@@
 \VariableTok{sim} \OperatorTok{=} \ClassTok{Simulator}\KeywordTok{(\ClassTok{Oscillator})}
-\VariableTok{df} \OperatorTok{=} \VariableTok{sim}.\FunctionTok{solve}\KeywordTok{(}\VariableTok{save\_at}\OperatorTok{=}\ClassTok{np}.\FunctionTok{linspace}(\ValueTok{0}, \ValueTok{5}, \ValueTok{101})\KeywordTok{)}
+\VariableTok{df} \OperatorTok{=} \VariableTok{sim}.\FunctionTok{solve}\KeywordTok{(}\VariableTok{save\_at}\OperatorTok{=}\ClassTok{np}.\FunctionTok{linspace}\KeywordTok{(}\ValueTok{0}, \ValueTok{5}, \ValueTok{101}\KeywordTok{))}
 \end{Highlighting}

### 2024-01-11 article.tex
Update syntax highlighting
Reorder authors
Add Resouces section
Change code comment color
Remove "Trust" from Wellcome Genome Campus name

@@
 \newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.40,0.45,0.13}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\ExtensionTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.28,0.35,0.67}{#1}}
-\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.46,0.62}{#1}}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\RegionMarkerTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.07,0.07,0.07}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.13,1.00}{#1}}
+\newcommand{\ClassTok}[1]{\textcolor[rgb]{0.27,0.56,0.65}{#1}}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.06,0.50}{#1}}
+\newcommand{\ValueTok}[1]{\textcolor[rgb]{0.13,0.57,0.41}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.47,0.37,0.15}{#1}}
+\newcommand{\IDETok}[1]{\textcolor[rgb]{0.58,0.58,0.58}{#1}}
+\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.06,0.50}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.00,0.50,0.00}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.70,0.27,0.27}{#1}}
 
@@
 \author[1,2]{Mauro Silberberg}
+\author[3]{Henning Hermjakob}
 \author[3]{Rahuman S. Malik-Sheriff}
-\author[3]{Henning Hermjakob}
 \author[1,2]{Hernán E. Grecco}
@@
 \affil[2]{CONICET - Universidad de Buenos Aires, Instituto de Física de Buenos Aires (IFIBA). Buenos Aires, Argentina}
-\affil[3]{European Bioinformatics Institute, European Molecular Biology Laboratory (EMBL-EBI), Wellcome Trust Genome Campus, Cambridge, UK}
+\affil[3]{European Bioinformatics Institute, European Molecular Biology Laboratory (EMBL-EBI), Wellcome Genome Campus, Cambridge, UK}
 \date{}
@@
 
+\hypertarget{resources}{%
+\section{Resources}\label{resources}}
+
+Source code repositories for these packages can be found at
+\url{https://github.com/maurosilber/poincare} and
+\url{https://github.com/hgrecco/simbio}.
+
 \hypertarget{references}{%

### 2024-02-27 src/performance/timer.py
Make parent dir before saving results

@@ -124,8 +124,10 @@ if __name__ == "__main__":
     path = Path(__file__).parent
 
     models = pd.read_csv(path / "models.txt")
-    results = models.progress_apply(
+    results: pd.DataFrame = models.progress_apply(
         lambda x: pd.Series(run_model(**x, runner=runner)),
         axis="columns",
     )
-    results.to_csv(path / f"results/{output}.csv", index=False)
+    results_dir = path / "results"
+    results_dir.mkdir(exist_ok=True)
+    results.to_csv(results_dir / f"{output}.csv", index=False)

### 2024-02-27 article.qmd
Use vector file for figure 6

--- 2024-01-02 article.qmd
@@
 (left) Run time for the model BIOMD3 as a function of the number of output points.
-(right) Run time for different models for 300 output points, using the geometric mean of the different softwares to order them.](src/performance/figures/performance.png){#fig-runtime}
+(right) Run time for different models for 300 output points, using the geometric mean of the different softwares to order them.](src/performance/figures/performance.svg){#fig-runtime}
 

### 2024-03-05 article.tex
Address language issues
- Page 1, Line 49: "Python-centric approach" -> "Our Python-centric approach"?
- Page 2, Line 47: "could be a demanding"
- Page 3, Line 39: "Poincare allows defining" -> "Poincare allows one to define..."
- Page 5, Line 48: "Oscilator" -> "Oscillator"
- Page 5, Line 50: "such that allows to easily define of"
- Page 7, Line 24: "on a factor" -> "by a factor"

@@
 standard typed modern Python syntax that provides a seamless integration
-with Integrated Development Environments (IDEs). Python-centric approach
+with Integrated Development Environments (IDEs). Our Python-centric approach
 significantly enhances code analysis, error detection, and refactoring
@@
 SBML. An extension for Visual Studio Code was developed, but its
-maintenance could be demanding task for the systems biology community.
+maintenance could be a demanding task for the systems biology community.
 PySB, using Python's dynamic nature, created a DSL within Python. By
@@
 open-source Python packages for defining and simulating systems.
-Poincaré allows defining differential equation systems, while SimBio
+Poincaré allows one to define differential equation systems, while SimBio
 builds on it for defining reaction networks. They are focused on
@@
   automatically saved in the component for introspection (i.e.,
-  \texttt{Oscilator.x.name\ ==\ "x"}),
+  \texttt{Oscillator.x.name\ ==\ "x"}),
 \item
-  It provides a namespace such that allows to easily define of multiple
+  It provides a namespace that allows to easily define multiple
   independent models in the same script,
@@
 
-Switching backends to \texttt{"numba"} can improve model runtime on a
+Switching backends to \texttt{"numba"} can improve model runtime by a
 factor up to x30 or more for big or long running models, however incurs

### 2024-03-05 article.tex
Page 2, Line 8: "...and representation of complex...", "...series reactions." -> sentence is poorly written, rewrite

@@
 engineering and epidemiology. They are comprised of a set of chemical
-species or biological entities and representation of complex interaction
-and transformation between them through a series reactions. These
+species or biological entities
+and a set of reactions that mediate transformations between them. These
 systems can be modelled and simulated through multiple approaches:

### 2024-03-05 article.tex
Use the acronym package

@@
 
+\usepackage{acronym}
+\acrodef{CRN}[CRN]{chemical reaction network}
+\acrodef{DSL}[DSL]{domain-specific language}
+\acrodef{GPU}[GPU]{graphical processing unit}
+\acrodef{GUI}[GUI]{graphical user interface}
+\acrodef{IDE}[IDE]{integrated development environment}
+\acrodef{JIT}[JIT]{just-in-time}
+\acrodef{ODE}[ODE]{ordinary differential equation}
+\acrodef{PyPI}[PyPI]{Python package index}
+\acrodef{RHS}[RHS]{right-hand side}
+\acrodef{SBML}[SBML]{System Biology Markup Language}
+\acrodef{SDE}[SDE]{stochastic differential equation}
+\acrodef{TPU}[TPU]{tensor processing unit}
+
 \title{Poincaré and SimBio: a versatile and extensible Python ecosystem
@@
 \begin{abstract}
-Chemical Reaction Networks (CRNs) play a pivotal role in diverse fields
+\Acf{CRN} play a pivotal role in diverse fields
 such as systems biology, biochemistry, chemical engineering, and
-epidemiology. High-level modelling of CRNs enables various simulation
+epidemiology. High-level modelling of \acp{CRN} enables various simulation
 approaches, including deterministic and stochastic methods. However,
-existing Python tools for CRN modelling typically wrap external C/C++
+existing Python tools for \ac{CRN} modelling typically wrap external C/C++
 libraries for modelling and simulation, limiting their extensibility and
@@
 Poincaré and SimBio, two novel Python packages for the definition and
-simulation of dynamical systems and CRNs. Poincaré serves as a
+simulation of dynamical systems and \acp{CRN}. Poincaré serves as a
 foundation for dynamical system modelling, while SimBio extends this
-functionality to CRNs, including support for the Systems Biology Markup
-Language (SBML). Poincaré and SimBio are developed as pure Python
+functionality to \acp{CRN}, including support for the 
+SBML. Poincaré and SimBio are developed as pure Python
 packages enabling users to easily extend their simulation capabilities
@@
 standard typed modern Python syntax that provides a seamless integration
-with Integrated Development Environments (IDEs). Our Python-centric approach
+with \acp{IDE}. Our Python-centric approach
 significantly enhances code analysis, error detection, and refactoring
@@
 
+\acresetall % reset acronyms after abstract
+
 \hypertarget{introduction}{%
@@
 
-Chemical Reaction Networks (CRNs) are a fundamental concept of modelling
+\Acp{CRN} are a fundamental concept of modelling
 in numerous fields including systems biology, biochemistry, chemical
@@
 systems can be modelled and simulated through multiple approaches:
-deterministic Ordinary Differential Equations (ODEs) to model
-macroscopic behavior, Stochastic Differential Equations (SDEs) to model
+deterministic \acp{ODE} to model
+macroscopic behavior, \acp{SDE} to model
 microscopic fluctuations, and jump processes (Gillespie-like
@@
 
-Several tools already exist for defining and simulating CRNs.
+Several tools already exist for defining and simulating \acp{CRN}.
 BioSimulators.org (Shaikh et al. 2022), a registry of simulation tools,
@@
 Tellurium (Choi et al. 2018) and PySB (Lopez et al. 2013). COPASI is a
-standalone software with a GUI for defining and simulating models. It's
+standalone software with a \ac{GUI} for defining and simulating models. It's
 widely used for its user-friendly interface and comprehensive features.
@@
 however, it uses a C++ library called libRoadRunner in the backend to
-simulate models. PySB is a Python library that created a DSL using
-standard Python to define models which are then compiled to ODE using a
+simulate models. PySB is a Python library that created a \ac{DSL} using
+standard Python to define models which are then compiled to \acp{ODE} using a
 Perl library called BioNetGen (L. A. Harris et al. 2016).
@@
 
-Another challenge is the way models are written. Many tools use a Domain
-Specific Language (DSL) for coding, and support the System Biology
-Markup Language (SBML) as an exchange format (Hucka et al., n.d.), as
-direct SBML coding is impractical. A DSL allows reuse of code in
+Another challenge is the way models are written. Many tools use a \ac{DSL} for coding, and support the
+\ac{SBML} as an exchange format (Hucka et al., n.d.), as
+direct \ac{SBML} coding is impractical. A \ac{DSL} allows reuse of code in
 different programming environments, but it will not be recognized by
-default in Integrated Development Environments (IDEs). Therefore DSLs
+default in \acp{IDE}. Therefore \acp{DSL}
 cannot provide the development experience such as syntax highlighting,
 code completion, refactoring, and static analysis, unless such support
-is specifically developed. Tellurium uses a Domain Specific Language
-(DSL) called Antimony (Smith et al. 2009), which can be translated to
-SBML. An extension for Visual Studio Code was developed, but its
+is specifically developed. Tellurium uses a
+\ac{DSL} called Antimony (Smith et al. 2009), which can be translated to
+\ac{SBML}. An extension for Visual Studio Code was developed, but its
 maintenance could be a demanding task for the systems biology community.
-PySB, using Python's dynamic nature, created a DSL within Python. By
+PySB, using Python's dynamic nature, created a \ac{DSL} within Python. By
 default, it uses global state to create species and parameters, without
 assigning them to Python variables or adding them explicitly to the
-model, but this approach is not fully compatible with IDEs, affecting
+model, but this approach is not fully compatible with \acp{IDE}, affecting
 the development experience.
@@
 providing an ergonomic experience to end-users by integrating well with
-IDEs and static analysis tools through the use of standard modern Python
+\acp{IDE} and static analysis tools through the use of standard modern Python
 syntax. Since they are coded in Python, every part from model definition
@@
 define reaction systems and interface with systems biology standards
-such as SBML. These are pure Python packages with standard dependencies
+such as \ac{SBML}. These are pure Python packages with standard dependencies
 from the PyData scientific stack such as NumPy (C. R. Harris et al.
-2020) and pandas (McKinney 2010). They are published in PyPI (the Python
-Package Index), where links to the source code and documentation hosted
+2020) and pandas (McKinney 2010). They are published in the \acf{PyPI}
+, where links to the source code and documentation hosted
 in GitHub can be found, and can be easily installed with
@@
 NumPy (C. R. Harris et al. 2020); Numba (Lam, Pitrou, and Seibert 2015),
-a Just-in-Time (JIT) compiler to LLVM; SymPy (Meurer et al. 2017), a
+a \ac{JIT} compiler to LLVM; SymPy (Meurer et al. 2017), a
 library for symbolic mathematics; and JAX (Bradbury et al. 2018), a
-library that support automatic differentiation and compilation to GPUs
-and TPUs. Symbolite is designed to facilitate the easy integration of
+library that support automatic differentiation and compilation to \acp{GPU}
+and \acp{TPU}. Symbolite is designed to facilitate the easy integration of
 new backends.
@@
 \item
-  It allows IDEs to provide autocomplete and refactoring capabilities
+  It allows \acp{IDE} to provide autocomplete and refactoring capabilities
   (\texttt{Oscillator.\textless{}TAB\textgreater{}} shows \texttt{x},
@@
 
-For this last point, IDEs that support \texttt{dataclass\_transform} (De
+For this last point, \acp{IDE} that support \texttt{dataclass\_transform} (De
 Bonte and Traut 2021) can provide a tooltip with the expected signature
@@
 To simulate a system, we created a \texttt{Simulator} instance
-(Figure~\ref{fig-sim}), which compiles a given system and interfaces
-with solvers. By default, it creates a first-order ordinary differential
-equation (ODE) system using \texttt{numpy} as a backend. This can be
+(Figure~\ref{fig-sim}), which compiles the \ac{RHS} of a given system and interfaces
+with \ac{ODE} solvers. By default, it creates a first-order
+\ac{ODE} system using \texttt{numpy} as a backend. This can be
 easily switched to other solvers. The Simulator wraps the output in a
@@
 \texttt{MassAction}. Additionally, SimBio supports importing from and
-exporting to SBML, and downloading them directly from BioModels
+exporting to \ac{SBML}, and downloading them directly from BioModels
 (Malik-Sheriff et al. 2020) (Figure~\ref{fig-simbio-io}).
@@
 
-To evaluate SimBio's reproducibility, we analyzed curated SBML models
+To evaluate SimBio's reproducibility, we analyzed curated \ac{SBML} models
 from BioModels (Malik-Sheriff et al. 2020). Among the first 100 curated
@@
 with them. A user might have to consider the trade-off between
-compilation and run times, as the compilation of the right-hand-side
-(RHS) code might take longer than the runtime itself, and not be worth
+compilation and run times, as the compilation of the
+\ac{RHS} code might take longer than the runtime itself, and not be worth
 it for running the model only once. Another speed-up in the runtime can
@@
 for defining and simulating dynamical systems and reaction networks.
-These packages are deeply integrated with Integrated Development
-Environments (IDEs), enabling code analysis tools to identify errors
+These packages are deeply integrated with
+\acp{IDE}, enabling code analysis tools to identify errors
 prior to execution and assist in refactoring and code completion. We
@@
 Our approach differs from previous tools in that both the model
-definition and its compilation into an Ordinary Differential Equation
-(ODE) function are entirely Python-based. This approach simplifies the
+definition and its compilation into an
+\ac{ODE} function are entirely Python-based. This approach simplifies the
 development of various simulation methods, including performance
@@
 Python-based does not compromise performance compared to C/C++ tools, as
-the ODE functions can be Just-In-Time (JIT) compiled using Numba.
+the \ac{ODE} functions can be \ac{JIT} compiled using Numba.
 
-The inclusion of SBML support facilitates the effortless reuse of models
+The inclusion of \ac{SBML} support facilitates the effortless reuse of models
 created by the systems biology community, along with the vast collection

### 2024-03-05 article.tex
Correct figure reference

@@
 Bonte and Traut 2021) can provide a tooltip with the expected signature
-(Figure~\ref{fig-composition}). This requires the use of type
+(Figure~\ref{fig-ide}). This requires the use of type
 annotations which play a more significant role in static type checking,

### 2024-03-05 article.tex
Replace which <-> that

@@
 with Numba. Our benchmark tests using curated models from the BioModels
-repository, demonstrate that these tools may provide a potentially
-superior performance advantage, compared to other existing tools.
+repository demonstrate that these tools may provide a potentially
+superior performance advantage compared to other existing tools.
 Additionally, to ensure a user-friendly experience, our packages use
@@
 simulations) to account for the discreteness of populations. Instead of
-directly writing the equations for each of these formulations which is
+directly writing the equations for each of these formulations, which is
 error-prone and difficult to reuse, these models can be written in a
@@
 simulate models. PySB is a Python library that created a \ac{DSL} using
-standard Python to define models which are then compiled to \acp{ODE} using a
+standard Python to define models, which are then compiled to \acp{ODE} using a
 Perl library called BioNetGen (L. A. Harris et al. 2016).
@@
 \item
-  It allows creation of instances which can be composed into a bigger
+  It allows creation of instances that can be composed into a bigger
   model (Figure~\ref{fig-composition}).
@@
 (Figure~\ref{fig-ide}). This requires the use of type
-annotations which play a more significant role in static type checking,
+annotations, which play a more significant role in static type checking
 as they can help to identify errors before running the code. For
@@
 To simulate a system, we created a \texttt{Simulator} instance
-(Figure~\ref{fig-sim}), which compiles the \ac{RHS} of a given system and interfaces
+(Figure~\ref{fig-sim}) that compiles the \ac{RHS} of a given system and interfaces
 with \ac{ODE} solvers. By default, it creates a first-order
@@
 factor up to x30 or more for big or long running models, however incurs
-in a compile time penalty for the first run which must be taken into
+in a compile time penalty for the first run that must be taken into
 account.
@@
 (Figure~\ref{fig-simbio}). Additionally, SimBio features
-\texttt{MassAction}, a subclass of \texttt{RateLaw}, which intuitively
+\texttt{MassAction}, a subclass of \texttt{RateLaw}, that intuitively
 incorporates reactants into the rate law (Figure~\ref{fig-simbio}).
@@
 from BioModels (Malik-Sheriff et al. 2020). Among the first 100 curated
-models from BioModels we selected 60 which did not contains events, as
+models from BioModels we selected 60 that did not contains events, as
 SimBio doesn't support events yet. We simulated the selected models with
@@
 be had by switching the LSODA \texttt{scipy} solver for a more efficient
-\texttt{numbalsoda} implementation, which avoids the Python interpreter
+\texttt{numbalsoda} implementation that avoids the Python interpreter
 between each of the integration steps. This last combination beats all

### 2024-03-06 article.tex
Use bibtex instead of what quarto generated

@@
 Several tools already exist for defining and simulating \acp{CRN}.
-BioSimulators.org (Shaikh et al. 2022), a registry of simulation tools,
-list at least 15 Python software including COPASI (Hoops et al. 2006),
-Tellurium (Choi et al. 2018) and PySB (Lopez et al. 2013). COPASI is a
+BioSimulators.org \cite{shaikhBioSimulatorsCentralRegistry2022}, a registry of simulation tools,
+list at least 15 Python software including COPASI \cite{hoopsCOPASICOmplexPAthway2006},
+Tellurium \cite{choiTelluriumExtensiblePythonbased2018} and PySB \cite{lopezProgrammingBiologicalModels2013}. COPASI is a
 standalone software with a \ac{GUI} for defining and simulating models. It's
@@
 standard Python to define models, which are then compiled to \acp{ODE} using a
-Perl library called BioNetGen (L. A. Harris et al. 2016).
+Perl library called BioNetGen \cite{harrisBioNetGenAdvancesRulebased2016}.
 
@@
 Another challenge is the way models are written. Many tools use a \ac{DSL} for coding, and support the
-\ac{SBML} as an exchange format (Hucka et al., n.d.), as
+\ac{SBML} as an exchange format \cite{huckaSBMLL3V2}, as
 direct \ac{SBML} coding is impractical. A \ac{DSL} allows reuse of code in
@@
 is specifically developed. Tellurium uses a
-\ac{DSL} called Antimony (Smith et al. 2009), which can be translated to
+\ac{DSL} called Antimony \cite{smithAntimonyModularModel2009}, which can be translated to
 \ac{SBML}. An extension for Visual Studio Code was developed, but its
@@
 such as \ac{SBML}. These are pure Python packages with standard dependencies
-from the PyData scientific stack such as NumPy (C. R. Harris et al.
-2020) and pandas (McKinney 2010). They are published in the \acf{PyPI}
+from the PyData scientific stack such as NumPy \cite{harrisArrayProgrammingNumPy2020}
+and pandas \cite{mckinneyDataStructuresStatistical2010}. They are published in the \acf{PyPI}
 , where links to the source code and documentation hosted
@@
 compiled to various backends. Currently, we have implementations for
-NumPy (C. R. Harris et al. 2020); Numba (Lam, Pitrou, and Seibert 2015),
-a \ac{JIT} compiler to LLVM; SymPy (Meurer et al. 2017), a
-library for symbolic mathematics; and JAX (Bradbury et al. 2018), a
+NumPy \cite{harrisArrayProgrammingNumPy2020}; Numba \cite{lamNumbaLLVMbasedPython2015},
+a \ac{JIT} compiler to LLVM; SymPy \cite{meurerSymPySymbolicComputing2017}, a
+library for symbolic mathematics; and JAX \cite{jax2018github}, a
 library that support automatic differentiation and compilation to \acp{GPU}
@@
 
-For this last point, \acp{IDE} that support \texttt{dataclass\_transform} (De
-Bonte and Traut 2021) can provide a tooltip with the expected signature
+For this last point, \acp{IDE} that support \texttt{dataclass\_transform} \cite{debontePEP681Data2021}
+can provide a tooltip with the expected signature
 (Figure~\ref{fig-ide}). This requires the use of type
@@
 exporting to \ac{SBML}, and downloading them directly from BioModels
-(Malik-Sheriff et al. 2020) (Figure~\ref{fig-simbio-io}).
+\cite{malik-sheriffBioModels15Years2020} (Figure~\ref{fig-simbio-io}).
 
@@
 To evaluate SimBio's reproducibility, we analyzed curated \ac{SBML} models
-from BioModels (Malik-Sheriff et al. 2020). Among the first 100 curated
+from BioModels \cite{malik-sheriffBioModels15Years2020}. Among the first 100 curated
 models from BioModels we selected 60 that did not contains events, as
@@
 
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-jax2018github}{}}%
-Bradbury, James, Roy Frostig, Peter Hawkins, Matthew James Johnson,
-Chris Leary, Dougal Maclaurin, George Necula, et al. 2018. {``{JAX}:
-Composable Transformations of {Python}+{NumPy} Programs.''}
-
-\leavevmode\vadjust pre{\hypertarget{ref-choiTelluriumExtensiblePythonbased2018}{}}%
-Choi, Kiri, J Kyle Medley, Matthias König, Kaylene Stocking, Lucian
-Smith, Stanley Gu, and Herbert M Sauro. 2018. {``Tellurium: {An}
-Extensible Python-Based Modeling Environment for Systems and Synthetic
-Biology.''} \emph{Bio Systems} 171 (September): 74--79.
-\url{https://doi.org/10.1016/j.biosystems.2018.07.006}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-debontePEP681Data2021}{}}%
-De Bonte, Erik, and Eric Traut. 2021. {``{PEP} 681 {\textendash} {Data
-Class Transforms}.''} \emph{Python Enhancement Proposals}.
-https://peps.python.org/pep-0681/.
-
-\leavevmode\vadjust pre{\hypertarget{ref-harrisArrayProgrammingNumPy2020}{}}%
-Harris, Charles R., K. Jarrod Millman, Stéfan J. van der Walt, Ralf
-Gommers, Pauli Virtanen, David Cournapeau, Eric Wieser, et al. 2020.
-{``Array Programming with {NumPy}.''} \emph{Nature} 585 (7825): 357--62.
-\url{https://doi.org/10.1038/s41586-020-2649-2}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-harrisBioNetGenAdvancesRulebased2016}{}}%
-Harris, Leonard A, Justin S Hogg, José-Juan Tapia, John A P Sekar,
-Sanjana Gupta, Ilya Korsunsky, Arshi Arora, Dipak Barua, Robert P
-Sheehan, and James R Faeder. 2016. {``{BioNetGen} 2.2: Advances in
-Rule-Based Modeling.''} \emph{Bioinformatics (Oxford, England)} 32 (21):
-3366--68. \url{https://doi.org/10.1093/bioinformatics/btw469}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hoopsCOPASICOmplexPAthway2006}{}}%
-Hoops, Stefan, Sven Sahle, Ralph Gauges, Christine Lee, Jürgen Pahle,
-Natalia Simus, Mudita Singhal, Liang Xu, Pedro Mendes, and Ursula
-Kummer. 2006. {``{COPASI}{\textemdash}a {COmplex PAthway SImulator}.''}
-\emph{Bioinformatics} 22 (24): 3067--74.
-\url{https://doi.org/10.1093/bioinformatics/btl485}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-huckaSBMLL3V2}{}}%
-Hucka, Michael, Frank Bergmann, Stefan Hoops, Sarah M Keating, Nicolas
-Le Novère, Chris J Myers, Brett G Olivier, et al. n.d. {``{SBML
-L3V2}.''}
-
-\leavevmode\vadjust pre{\hypertarget{ref-lamNumbaLLVMbasedPython2015}{}}%
-Lam, Siu Kwan, Antoine Pitrou, and Stanley Seibert. 2015. {``Numba: A
-{LLVM-based Python JIT} Compiler.''} In \emph{Proceedings of the {Second
-Workshop} on the {LLVM Compiler Infrastructure} in {HPC}}, 1--6. {LLVM}
-'15. {New York, NY, USA}: {Association for Computing Machinery}.
-\url{https://doi.org/10.1145/2833157.2833162}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lopezProgrammingBiologicalModels2013}{}}%
-Lopez, Carlos F, Jeremy L Muhlich, John A Bachman, and Peter K Sorger.
-2013. {``Programming Biological Models in {Python} Using {PySB}.''}
-\emph{Molecular Systems Biology} 9 (January): 646.
-\url{https://doi.org/10.1038/msb.2013.1}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-malik-sheriffBioModels15Years2020}{}}%
-Malik-Sheriff, Rahuman S, Mihai Glont, Tung V N Nguyen, Krishna Tiwari,
-Matthew G Roberts, Ashley Xavier, Manh T Vu, et al. 2020.
-{``{BioModels}{\textemdash}15 Years of Sharing Computational Models in
-Life Science.''} \emph{Nucleic Acids Research} 48 (D1): D407--15.
-\url{https://doi.org/10.1093/nar/gkz1055}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-mckinneyDataStructuresStatistical2010}{}}%
-McKinney, Wes. 2010. {``Data {Structures} for {Statistical Computing} in
-{Python}.''} In \emph{Python in {Science Conference}}, 56--61. {Austin,
-Texas}. \url{https://doi.org/10.25080/Majora-92bf1922-00a}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-meurerSymPySymbolicComputing2017}{}}%
-Meurer, Aaron, Christopher P. Smith, Mateusz Paprocki, Ondřej Čertík,
-Sergey B. Kirpichev, Matthew Rocklin, AmiT Kumar, et al. 2017.
-{``{SymPy}: Symbolic Computing in {Python}.''} \emph{PeerJ Computer
-Science} 3 (January): e103. \url{https://doi.org/10.7717/peerj-cs.103}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-shaikhBioSimulatorsCentralRegistry2022}{}}%
-Shaikh, Bilal, Lucian P Smith, Dan Vasilescu, Gnaneswara Marupilla,
-Michael Wilson, Eran Agmon, Henry Agnew, et al. 2022.
-{``{BioSimulators}: A Central Registry of Simulation Engines and
-Services for Recommending Specific Tools.''} \emph{Nucleic Acids
-Research} 50 (W1): W108--14. \url{https://doi.org/10.1093/nar/gkac331}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-smithAntimonyModularModel2009}{}}%
-Smith, Lucian P., Frank T. Bergmann, Deepak Chandran, and Herbert M.
-Sauro. 2009. {``Antimony: A Modular Model Definition Language.''}
-\emph{Bioinformatics} 25 (18): 2452--54.
-\url{https://doi.org/10.1093/bioinformatics/btp401}.
-
-\end{CSLReferences}
+\bibliographystyle{plain}
+\bibliography{bibliography.bib}
 

### 2024-03-06 article.tex
Remove latex packages

@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
-%
-\documentclass[
-  letterpaper,
-  DIV=11,
-  numbers=noendperiod]{scrartcl}
+\documentclass{article}
+\usepackage[a4paper, margin=2.5cm]{geometry}
 
-\usepackage{amsmath,amssymb}
-\usepackage{iftex}
-\ifPDFTeX
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-\usepackage{lmodern}
-\ifPDFTeX\else  
-    % xetex/luatex font selection
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-% Make \paragraph and \subparagraph free-standing
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
+\usepackage{amsmath}
+\usepackage[svgnames]{xcolor}
+\usepackage{hyperref}
+\usepackage{graphicx}
+\usepackage{float}
+\usepackage{caption}
+\usepackage{subcaption}
+\usepackage[skins,breakable]{tcolorbox}
 
-\usepackage{color}
 \usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
 \DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
 \definecolor{shadecolor}{RGB}{100,100,100}
@@
 \providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs}
 % Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
-\usepackage{graphicx}
+
 \makeatletter
@@
 \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
-  \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
-\KOMAoption{captions}{tableheading}
-\makeatletter
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
-\@ifpackageloaded{caption}{}{\usepackage{caption}}
-\AtBeginDocument{%
-\ifdefined\contentsname
-  \renewcommand*\contentsname{Table of contents}
-\else
-  \newcommand\contentsname{Table of contents}
-\fi
-\ifdefined\listfigurename
-  \renewcommand*\listfigurename{List of Figures}
-\else
-  \newcommand\listfigurename{List of Figures}
-\fi
-\ifdefined\listtablename
-  \renewcommand*\listtablename{List of Tables}
-\else
-  \newcommand\listtablename{List of Tables}
-\fi
-\ifdefined\figurename
-  \renewcommand*\figurename{Figure}
-\else
-  \newcommand\figurename{Figure}
-\fi
-\ifdefined\tablename
-  \renewcommand*\tablename{Table}
-\else
-  \newcommand\tablename{Table}
-\fi
-}
-\@ifpackageloaded{float}{}{\usepackage{float}}
-\floatstyle{ruled}
-\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
-\floatname{codelisting}{Listing}
-\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
-\makeatother
-\makeatletter
-\@ifpackageloaded{caption}{}{\usepackage{caption}}
-\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
-\makeatother
-\makeatletter
-\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
-\makeatother
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
 \hypersetup{
@@
   urlcolor={Blue},
-  pdfcreator={LaTeX via pandoc}}
-
-\usepackage{authblk}
-\renewcommand\Affilfont{\small}
+}
 
@@
 
-\title{Poincaré and SimBio: a versatile and extensible Python ecosystem
-for modeling systems.}
+\usepackage{authblk}
+\title{Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems.}
 \author[1,2]{Mauro Silberberg}
@@
 \affil[3]{European Bioinformatics Institute, European Molecular Biology Laboratory (EMBL-EBI), Wellcome Genome Campus, Cambridge, UK}
-\date{}
 

### 2024-03-08 article.tex
Simplify code boxes

@@
 \documentclass{article}
-\usepackage[a4paper, margin=2.5cm]{geometry}
+\usepackage[a4paper, margin=3cm, bottom=4cm]{geometry}
 
@@
 \usepackage{subcaption}
-\usepackage[skins,breakable]{tcolorbox}
+\usepackage{tcolorbox}
 
 \usepackage{fancyvrb}
+\newenvironment{Shaded}
+{\begin{tcolorbox}[title=input,boxrule=0pt]}
+{\end{tcolorbox}}
+\newenvironment{Shaded2}
+{\begin{tcolorbox}[title=output,boxrule=0pt]}
+{\end{tcolorbox}}
 \DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-\definecolor{shadecolor}{RGB}{100,100,100}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
 \newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.13,1.00}{#1}}
@@
 \begin{document}
+
 \maketitle
+
 \begin{abstract}
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, enhanced]}{\end{tcolorbox}}\fi
 

### 2024-03-08 article.tex
Rename shaded code environments

@@
 \usepackage{fancyvrb}
-\newenvironment{Shaded}
+\newenvironment{CodeInput}
 {\begin{tcolorbox}[title=input,boxrule=0pt]}
 {\end{tcolorbox}}
-\newenvironment{Shaded2}
+\newenvironment{CodeOutput}
 {\begin{tcolorbox}[title=output,boxrule=0pt]}

### 2024-03-09 src/performance/runners/simbio.py
Update SimBio runner to ignore SBML namespaces

@@ -1,7 +1,7 @@
 from __future__ import annotations
 
 from dataclasses import dataclass
-from typing import Literal
+from typing import Literal, Sequence
 
 import numpy as np
 import scipy.integrate
@@ -19,9 +19,13 @@ class SimBio(Runner):
     rtol: float
     solver: type[LSODA] | type[CVODE]
     backend: Literal["numpy", "numba"]
+    ignore_namespaces: Sequence[str]
 
     def load_model(self, sbml: str):
-        self.runner = Simulator(loads(sbml), backend=self.backend)
+        self.runner = Simulator(
+            loads(sbml, ignore_namespaces=self.ignore_namespaces),
+            backend=self.backend,
+        )
         return self
 
     def run_model(self, *, stop: float, n_points: int):

### 2024-03-20 src/performance/timer.py
ruff format

@@ -15,14 +15,9 @@ class Runner(Protocol):
     atol: float
     rtol: float
 
-    def __init__(self, *, atol: float, rtol: float):
-        ...
-
-    def load_model(self, sbml: str):
-        ...
-
-    def run_model(self, *, stop: float, n_points: int):
-        ...
+    def __init__(self, *, atol: float, rtol: float): ...
+    def load_model(self, sbml: str): ...
+    def run_model(self, *, stop: float, n_points: int): ...
 
 
 @dataclass

### 2024-03-20 src/performance/timer.py
Run performance benchmark multiple times separately
To evaluate standard deviation of times,
and to allow incrementally runnning new models.

@@ -4,11 +4,7 @@ from dataclasses import dataclass
 from time import perf_counter
 from typing import Protocol
 
-import pandas as pd
 from biomodels import get_omex
-from tqdm import tqdm
-
-tqdm.pandas()
 
 
 class Runner(Protocol):
@@ -31,7 +27,6 @@ def time(
     runner: Runner,
     model: Model,
     n_points: list[int],
-    n_runs: int,
 ) -> dict[str, float]:
     time_start = perf_counter()
     runner.load_model(model.sbml)
@@ -46,10 +41,9 @@ def time(
 
     for n in n_points:
         start = perf_counter()
-        for _ in range(n_runs):
-            runner.run_model(stop=model.stop, n_points=n)
+        runner.run_model(stop=model.stop, n_points=n)
         stop = perf_counter()
-        output[str(n)] = (stop - start) / n_runs
+        output[str(n)] = stop - start
     return output
 
 
@@ -64,7 +58,6 @@ def run_model(model_id: str, t_end: float, *, runner: type[Runner]):
             runner=runner(atol=1e-9, rtol=1e-6),
             model=model,
             n_points=[10, 30, 100, 300, 1000, 3000],
-            n_runs=100,
         )
 
     except Exception:
@@ -81,6 +74,9 @@ if __name__ == "__main__":
     import sys
     from functools import partial
     from pathlib import Path
+
+    import pandas as pd
+    from tqdm import tqdm
 
     match sys.argv[1:]:
         case ["copasi"]:
@@ -114,15 +110,16 @@ if __name__ == "__main__":
         case _:
             raise ValueError(sys.argv[1:])
 
-    output = "-".join(sys.argv[1:])
-
     path = Path(__file__).parent
-
-    models = pd.read_csv(path / "models.txt")
-    results: pd.DataFrame = models.progress_apply(
-        lambda x: pd.Series(run_model(**x, runner=runner)),
-        axis="columns",
-    )
-    results_dir = path / "results"
-    results_dir.mkdir(exist_ok=True)
-    results.to_csv(results_dir / f"{output}.csv", index=False)
+    output_path = path / "results" / "-".join(sys.argv[1:])
+    output_path.mkdir(exist_ok=True, parents=True)
+    models = pd.read_csv(path / "times.txt", header=None)
+    for _, (model_num, log_time) in tqdm(models.iterrows(), total=len(models)):
+        model_id = f"BIOMD{model_num}"
[…]

### 2024-03-20 src/performance/timer.py
Add header to models file

@@ -113,7 +113,7 @@ if __name__ == "__main__":
     path = Path(__file__).parent
     output_path = path / "results" / "-".join(sys.argv[1:])
     output_path.mkdir(exist_ok=True, parents=True)
-    models = pd.read_csv(path / "times.txt", header=None)
+    models = pd.read_csv(path / "times.txt")
     for _, (model_num, log_time) in tqdm(models.iterrows(), total=len(models)):
         model_id = f"BIOMD{model_num}"
         path = (output_path / model_id).with_suffix(".csv")

### 2024-03-20 src/performance/timer.py
Fix CSV save

@@ -116,10 +116,10 @@ if __name__ == "__main__":
     models = pd.read_csv(path / "times.txt")
     for _, (model_num, log_time) in tqdm(models.iterrows(), total=len(models)):
         model_id = f"BIOMD{model_num}"
-        path = (output_path / model_id).with_suffix(".csv")
-        if path.exists():
+        model_path = (output_path / model_id).with_suffix(".csv")
+        if model_path.exists():
             continue
         result = pd.DataFrame(
             [run_model(model_id, 10.0**log_time, runner=runner) for _ in range(10)]
         )
-        result.to_csv(output_path)
+        result.to_csv(model_path)

### 2024-03-20 src/performance/runners/simbio.py
Add default ignored namespaces

@@ -19,7 +19,16 @@ class SimBio(Runner):
     rtol: float
     solver: type[LSODA] | type[CVODE]
     backend: Literal["numpy", "numba"]
-    ignore_namespaces: Sequence[str]
+    ignore_namespaces: Sequence[str] = (
+        "jd",
+        "jd2",
+        "celldesigner",
+        "html",
+        "math",
+        "sbml",
+        "rdf",
+        "jigcell",
+    )
 
     def load_model(self, sbml: str):
         self.runner = Simulator(

### 2024-03-20 src/performance/timer.py
Change result names

@@ -83,19 +83,19 @@ if __name__ == "__main__":
             from .runners.copasi import COPASI as runner
         case ["roadrunner"]:
             from .runners.roadrunner import Tellurium as runner
-        case ["simbio", "numpy", "lsoda"]:
+        case ["simbio-numpy-lsoda"]:
             from .runners.simbio import LSODA, SimBio
 
             runner = partial(SimBio, backend="numpy", solver=LSODA)
-        case ["simbio", "numpy", "cvode"]:
+        case ["simbio-numpy-cvode"]:
             from .runners.simbio import CVODE, SimBio
 
             runner = partial(SimBio, backend="numpy", solver=CVODE)
-        case ["simbio", "numba", "lsoda"]:
+        case ["simbio-numba-lsoda"]:
             from .runners.simbio import LSODA, SimBio
 
             runner = partial(SimBio, backend="numba", solver=LSODA)
-        case ["simbio", "numba", "numbalsoda"]:
+        case ["simbio-numba-numbalsoda"]:
             from .runners.simbio import LSODA, SimBio
 
             runner = partial(
@@ -103,7 +103,7 @@ if __name__ == "__main__":
                 backend="numba",
                 solver=partial(LSODA, implementation="numbalsoda"),
             )
-        case ["simbio", "numba", "cvode"]:
+        case ["simbio-numba-cvode"]:
             from .runners.simbio import CVODE, SimBio
 
             runner = partial(SimBio, backend="numba", solver=CVODE)
@@ -111,7 +111,7 @@ if __name__ == "__main__":
             raise ValueError(sys.argv[1:])
 
     path = Path(__file__).parent
-    output_path = path / "results" / "-".join(sys.argv[1:])
+    output_path = path / "results" / sys.argv[1]
     output_path.mkdir(exist_ok=True, parents=True)
     models = pd.read_csv(path / "times.txt")
     for _, (model_num, log_time) in tqdm(models.iterrows(), total=len(models)):

### 2024-03-21 src/performance/timer.py
Change to 20 runs

@@ -120,6 +120,6 @@ if __name__ == "__main__":
         if model_path.exists():
             continue
         result = pd.DataFrame(
-            [run_model(model_id, 10.0**log_time, runner=runner) for _ in range(10)]
+            [run_model(model_id, 10.0**log_time, runner=runner) for _ in range(20)]
         )
         result.to_csv(model_path)

### 2024-03-21 src/performance/timer.py
Add try-except for run_model

@@ -1,5 +1,6 @@
 from __future__ import annotations
 
+import math
 from dataclasses import dataclass
 from time import perf_counter
 from typing import Protocol
@@ -31,8 +32,12 @@ def time(
     time_start = perf_counter()
     runner.load_model(model.sbml)
     time_loaded = perf_counter()
-    runner.run_model(stop=model.stop, n_points=n_points[0])
-    time_first_run = perf_counter()
+    try:
+        runner.run_model(stop=model.stop, n_points=n_points[0])
+    except Exception:
+        time_first_run = math.nan
+    else:
+        time_first_run = perf_counter()
 
     output = {
         "load": time_loaded - time_start,
@@ -41,8 +46,12 @@ def time(
 
     for n in n_points:
         start = perf_counter()
-        runner.run_model(stop=model.stop, n_points=n)
-        stop = perf_counter()
+        try:
+            runner.run_model(stop=model.stop, n_points=n)
+        except Exception:
+            stop = math.nan
+        else:
+            stop = perf_counter()
         output[str(n)] = stop - start
     return output
 

### 2024-03-21 article.tex
Rewrite introduction to clarify definition, translation and solving

@@
 
-\Acp{CRN} are a fundamental concept of modelling
-in numerous fields including systems biology, biochemistry, chemical
-engineering and epidemiology. They are comprised of a set of chemical
-species or biological entities
-and a set of reactions that mediate transformations between them. These
-systems can be modelled and simulated through multiple approaches:
-deterministic \acp{ODE} to model
-macroscopic behavior, \acp{SDE} to model
-microscopic fluctuations, and jump processes (Gillespie-like
-simulations) to account for the discreteness of populations. Instead of
-directly writing the equations for each of these formulations, which is
-error-prone and difficult to reuse, these models can be written in a
-higher-level description that can be compiled for these different types
-of simulations.
+\Acp{CRN} are a fundamental concept of modelling in numerous fields
+including systems biology, biochemistry, chemical engineering and epidemiology.
+They are comprised of a set of chemical species or biological entities
+and a set of reactions that mediate transformations between them.
+These systems can be simulated through multiple approaches:
+deterministic \acp{ODE} to model macroscopic behavior,
+\acp{SDE} to model microscopic fluctuations,
+and jump processes (Gillespie-like simulations) to account for the discreteness of populations.
+Instead of directly writing the equations for each of these formulations,
+which is error-prone and difficult to reuse,
+these models can be \emph{defined} in a higher-level description
+that can be \emph{translated} into equations for the different types of simulations
+and, then, \emph{solved} numerically.
 
-Several tools already exist for defining and simulating \acp{CRN}.
-BioSimulators.org \cite{shaikhBioSimulatorsCentralRegistry2022}, a registry of simulation tools,
-list at least 15 Python software including COPASI \cite{hoopsCOPASICOmplexPAthway2006},
-Tellurium \cite{choiTelluriumExtensiblePythonbased2018} and PySB \cite{lopezProgrammingBiologicalModels2013}. COPASI is a
-standalone software with a \ac{GUI} for defining and simulating models. It's
-widely used for its user-friendly interface and comprehensive features.
-There are also python bindings and interfaces for COPASI to allow
-advanced scripting. Tellurium is a Python-based modeling environment,
-however, it uses a C++ library called libRoadRunner in the backend to
-simulate models. PySB is a Python library that created a \ac{DSL} using
-standard Python to define models, which are then compiled to \acp{ODE} using a
-Perl library called BioNetGen \cite{harrisBioNetGenAdvancesRulebased2016}.
+Several tools already exist to \emph{define}, \emph{translate} and \emph{solve} \acp{CRN}.
+BioSimulators.org \cite{shaikhBioSimulatorsCentralRegistry2022},
+a registry of simulation tools,
+lists at least 15 Python software including
+COPASI \cite{hoopsCOPASICOmplexPAthway2006},
+Tellurium \cite{choiTelluriumExtensiblePythonbased2018} and
+PySB \cite{lopezProgrammingBiologicalModels2013}.
+COPASI is a standalone software with a \ac{GUI}
+that is widely used for its user-friendly interface and comprehensive features.
+It also includes Python bindings \cite{bergmannBASICOSimplifiedPython2023b} that allow advanced scripting.
+Tellurium is a Python-based modeling environment
+that uses a C++ library called libRoadRunner in the backend to translate and solve models.
+PySB is a Python library that created a \ac{DSL} using standard Python to define models,
+which are then translated to \acp{ODE} using a Perl library called BioNetGen \cite{harrisBioNetGenAdvancesRulebased2016}.
 
-One limitation of these tools is that they are not extensible from
-Python, as they are not fully Python packages but wrap libraries in
-other languages to do the simulation. While they enable model definition
-and simulation control via Python scripts, they don't fully leverage
-Python's extensive package ecosystem. For example, in COPASI, users are
-restricted to predefined distributions for Random Parameter Scans and
-cannot utilize the diverse distributions available in
-\texttt{scipy.stats}. Similarly, Tellurium doesn't allow the use of
-Python solvers, as adding new integrators requires C++ implementation.
+One limitation of these tools is their extensibility from Python.
+As they wrap libraries in other languages for defining, translating and/or solving models,
+these steps cannot be altered or easily inspected from Python.
+While they enable model definition and running simulations via Python scripts,
+they cannot fully leverage Python's extensive package ecosystem.
+For example,
+COPASI and Tellurium do not allow the use of solvers defined in other Python packages,
+and adding new integrators requires working with C++.
+In particular, the step that translates into equations is not exposed by any of these tools.
+As such, it is not possible to apply custom optimizations to the equations
+or use automatic differentation packages such as JAX \cite{jax2018github} to compute the model's jacobian.
 
@@
 
-Another challenge is the way models are written. Many tools use a \ac{DSL} for coding, and support the
-\ac{SBML} as an exchange format \cite{huckaSBMLL3V2}, as
-direct \ac{SBML} coding is impractical. A \ac{DSL} allows reuse of code in
-different programming environments, but it will not be recognized by
-default in \acp{IDE}. Therefore \acp{DSL}
-cannot provide the development experience such as syntax highlighting,
-code completion, refactoring, and static analysis, unless such support
-is specifically developed. Tellurium uses a
-\ac{DSL} called Antimony \cite{smithAntimonyModularModel2009}, which can be translated to
-\ac{SBML}. An extension for Visual Studio Code was developed, but its
-maintenance could be a demanding task for the systems biology community.
-PySB, using Python's dynamic nature, created a \ac{DSL} within Python. By
-default, it uses global state to create species and parameters, without
-assigning them to Python variables or adding them explicitly to the
-model, but this approach is not fully compatible with \acp{IDE}, affecting
-the development experience.
+Another challenge is the way models are defined.
+Many tools support the \ac{SBML} \cite{huckaSBMLL3V2} as an exchange format,
+a \emph{de facto} standard for \acp{CRN} that defines species, parameters and reactions between species.
+As writing \ac{SBML} directly is impractical,
+Tellurium uses a \ac{DSL} called Antimony \cite{smithAntimonyModularModel2009} for defining models.
+\Acp{DSL} allows to reuse the same code in different programming environments,
+but are not recognized by default in \acp{IDE}
+and, therefore, they cannot provide syntax highlighting, code completion, refactoring, and static analysis.
+For Antimony, an extension providing this capabilities was developed for Visual Studio Code,
+but its maintenance could be a demanding task for the systems biology community.
+In the case of PySB,
+using Python's dynamic nature,
+it developers designed a \ac{DSL} within Python.
+By default, it uses the global scope to create species and parameters,
+without explicitly assigning them to Python variables or to the model,
+but this approach is not fully compatible with \acp{IDE},
+affecting the development experience.
 
-To overcome these limitations, we developed poincaré and SimBio,
-open-source Python packages for defining and simulating systems.
-Poincaré allows one to define differential equation systems, while SimBio
-builds on it for defining reaction networks. They are focused on
-providing an ergonomic experience to end-users by integrating well with
-\acp{IDE} and static analysis tools through the use of standard modern Python
-syntax. Since they are coded in Python, every part from model definition
-to simulation can be extended from Python syntax. Being the first-ever
-pure Python packages for systems modelling, they offer extensive
-extensibility, from simple tasks like reusing integrators defined in
-other packages, to complex ones like altering the compilation process to
-leverage some structure in the equations. For example, using a for-loop
-in the compiled equations could improve the runtime performance if there
-is some repetitive structure in the system, as happens in spatial
-modelling. The models built using these packages can be introspected to
-create other representations, such as graphs connecting species and/or
-reactions, or tables with parameters or equations. Furthermore, they
-have a modular architecture with a clear separation of concerns, making
-it easier to maintain or to contribute new code, which is beneficial for
-developers and maintainers. We showcased the reliability of these tools
-by benchmarking them against the simulation results from other tools. We
-also highlighted the substantial performance improvements our tools
-offer, as this is crucial for construction and simulation of models of
-whole cells and organisms, which necessitate the simulation of
-significantly large-scale models.
+To overcome these limitations,
+we developed poincaré and SimBio,
+open-source Python packages for defining, translating and solving systems.
+Poincaré allows one to define differential equation systems
+using variables, parameters and constants,
+and assiging rate equations to variables.
+For defining \acp{CRN},
+SimBio builds on top of poincaré providing species and reactions that keep track of stoichiometries.
+Both are focused on providing an ergonomic experience to end-users
+by integrating well with \acp{IDE} and static analysis tools
+through the use of standard modern Python syntax.
+Moreover, since they are coded in pure Python,
+each step from model definition, translation to equations or solving
+can be extended or debugged from Python.
+Being the first-ever pure Python packages for systems modelling,
+they offer extensive extensibility,
+from simple tasks like reusing integrators defined in other packages,
+to complex ones like altering the compilation process to leverage some structure in the equations.
+For example, using a for-loop in the compiled equations could improve the runtime performance
+if there is some repetitive structure in the system,
+as happens in spatial modelling.
+The models built using these packages can be introspected to create other representations,
+such as graphs connecting species and/or reactions, or tables with parameters or equations.
+Furthermore, they have a modular architecture with a clear separation of concerns,
+making it easier to maintain or to contribute new code,
+which is beneficial for developers and maintainers.
+We showcased the reliability of these tools by benchmarking them against the simulation results from other tools.
+We also highlighted the substantial performance improvements our tools offer,
+as this is crucial for construction and simulation of models of whole cells and organisms,
+which necessitate the simulation of significantly large-scale models.
 

### 2024-03-21 article.tex
symbolite and sympy

@@
 
-Symbolite is a lightweight symbolics package to create algebraic
-mathematical expressions. Symbolite expressions can be inspected and
-compiled to various backends. Currently, we have implementations for
-NumPy \cite{harrisArrayProgrammingNumPy2020}; Numba \cite{lamNumbaLLVMbasedPython2015},
-a \ac{JIT} compiler to LLVM; SymPy \cite{meurerSymPySymbolicComputing2017}, a
-library for symbolic mathematics; and JAX \cite{jax2018github}, a
-library that support automatic differentiation and compilation to \acp{GPU}
-and \acp{TPU}. Symbolite is designed to facilitate the easy integration of
-new backends.
+Symbolite is a lightweight symbolics package to create algebraic mathematical expressions.
+Unlike SymPy \cite{meurerSymPySymbolicComputing2017},
+a widely-used Python library for symbolic mathematics,
+it only provides the building of an expression tree
+which can be inspected and compiled to various backends.
+Symbolite is designed to facilitate the integration of new backends.
+Currently, we have implementations for
+NumPy \cite{harrisArrayProgrammingNumPy2020};
+Numba \cite{lamNumbaLLVMbasedPython2015},
+a \ac{JIT} compiler to LLVM;
+SymPy \cite{meurerSymPySymbolicComputing2017};
+and JAX \cite{jax2018github}, a
+library that support automatic differentiation and compilation to \acp{GPU} and \acp{TPU}.
 

### 2024-03-21 article.tex
pip install simbio

@@
 
-Modular code architecture makes code reusable, extensible, and easier to
-maintain. Therefore, we split the code to define and simulate reaction
-systems into three Python packages: symbolite, to create symbolic
-expressions; poincaré, to define dynamical systems; and simbio, to
-define reaction systems and interface with systems biology standards
-such as \ac{SBML}. These are pure Python packages with standard dependencies
-from the PyData scientific stack such as NumPy \cite{harrisArrayProgrammingNumPy2020}
-and pandas \cite{mckinneyDataStructuresStatistical2010}. They are published in the \acf{PyPI}
-, where links to the source code and documentation hosted
-in GitHub can be found, and can be easily installed with
-\texttt{pip\ install\ \textless{}package\_name\textgreater{}}.
+Modular code architecture makes code reusable, extensible, and easier to maintain.
+Therefore, we split the code into three Python packages:
+symbolite, to create symbolic expressions;
+poincaré, to define dynamical systems;
+and simbio, to define \acp{CRN} and interface with systems biology standards such as \ac{SBML}.
+These are pure Python packages with standard dependencies from the PyData scientific stack
+such as NumPy \cite{harrisArrayProgrammingNumPy2020} and pandas \cite{mckinneyDataStructuresStatistical2010}.
+They are published in the \acf{PyPI},
+where links to the source code and documentation hosted in GitHub can be found,
+and can be easily installed with \texttt{pip\ install\ simbio},
+which installs symbolite and poincare as dependencies.
 

### 2024-03-21 article.tex
Results

@@
 it developers designed a \ac{DSL} within Python.
-By default, it uses the global scope to create species and parameters,
+To save keystrokes,
+it uses the global scope to create species and parameters,
 without explicitly assigning them to Python variables or to the model,
@@
 
-Poincare is a package to define and simulate dynamical systems. It
-provides a \texttt{System} class, where one can define
-\texttt{Constant}s, \texttt{Parameter}s, \texttt{Variable}s, and create
-equations linking a variable's derivative with an expression
-(Figure~\ref{fig-first-order}). It also allows to define higher-order
-systems by assigning an initial condition to a \texttt{Derivative}
-(Figure~\ref{fig-second-order}).
+Poincare is a package to define and simulate dynamical systems.
+It provides a \texttt{System} class,
+where one can define
+\texttt{Constant}s; which can be numbers or refer to other constants;
+\texttt{Parameter}s, which can be numbers or time-dependent expressions;
+\texttt{Variable}s, which represent the state of the system and must be provided with an initial condition;
+and create equations linking a variable's derivative with an expression (Figure~\ref{fig-first-order}).
+It also allows to define higher-order systems
+by assigning an initial condition to a \texttt{Derivative} (Figure~\ref{fig-second-order}).
 
-Utilizing classes for system definition offers several advantages:
+Within the constraints imposed by Python's current typing and static analyzers,
+we define models utilizing Python classes
+such that we can benefit from \acp{IDE}' autocomplete and refactoring capabilities.
+This design offer several advantages:
 
@@
 \item
-  The variable name to which a component is assigned can be
-  automatically saved in the component for introspection (i.e.,
-  \texttt{Oscillator.x.name\ ==\ "x"}),
+  The variable name to which a component is assigned can be automatically saved in the component for introspection
+  (i.e., \texttt{Oscillator.x.name\ ==\ "x"}),
 \item
-  It provides a namespace that allows to easily define multiple
-  independent models in the same script,
+  It provides a namespace that allows to easily define multiple independent models in the same script,
 \item
   It allows \acp{IDE} to provide autocomplete and refactoring capabilities
-  (\texttt{Oscillator.\textless{}TAB\textgreater{}} shows \texttt{x},
-  \texttt{v} and \texttt{eq}),
+  (\texttt{Oscillator.\textless{}TAB\textgreater{}} shows \texttt{x}, \texttt{v} and \texttt{eq}),
 \item
-  It allows creation of instances that can be composed into a bigger
-  model (Figure~\ref{fig-composition}).
+  It allows creation of instances that can be composed into a bigger model (Figure~\ref{fig-composition}).
 \end{enumerate}
@@
 
-For this last point, \acp{IDE} that support \texttt{dataclass\_transform} \cite{debontePEP681Data2021}
-can provide a tooltip with the expected signature
-(Figure~\ref{fig-ide}). This requires the use of type
-annotations, which play a more significant role in static type checking
-as they can help to identify errors before running the code. For
-instance, to parameterize the initial conditions of variables we have to
-use a \texttt{Constant}. If we try to use a \texttt{Parameter}, which
-could be a time-dependent expression, it is flagged as a type error
-(Figure~\ref{fig-ide}).
+For this last point,
+\acp{IDE} that support \texttt{dataclass\_transform} \cite{debontePEP681Data2021}
+can provide a tooltip with the expected signature (Figure~\ref{fig-ide}).
+This requires the use of type annotations,
+which play a more significant role in static type checking
+as they can help to identify errors before running the code.
+For instance, to parameterize the initial conditions of variables
+we have to use a \texttt{Constant}.
+If we try to use a \texttt{Parameter},
+which could be a time-dependent expression,
+it is flagged as a type error (Figure~\ref{fig-ide}).
 
-To simulate a system, we created a \texttt{Simulator} instance
-(Figure~\ref{fig-sim}) that compiles the \ac{RHS} of a given system and interfaces
-with \ac{ODE} solvers. By default, it creates a first-order
-\ac{ODE} system using \texttt{numpy} as a backend. This can be
-easily switched to other solvers. The Simulator wraps the output in a
-\texttt{pandas.DataFrame}, which can be easily plotted with the standard
-\texttt{plot} method.
+To simulate a system,
+we created a \texttt{Simulator} instance (Figure~\ref{fig-sim})
+that translates the model into the \ac{RHS} equations
+and interfaces with \ac{ODE} solvers.
+By default,
+it creates a first-order \ac{ODE} system using \texttt{numpy} as a backend,
+and uses the LSODA solver from \texttt{scipy}.
+This can be easily switched to other solvers.
+The Simulator wraps the output in a \texttt{pandas.DataFrame},
+which can be easily plotted with the standard \texttt{plot} method.
 
-Switching backends to \texttt{"numba"} can improve model runtime by a
-factor up to x30 or more for big or long running models, however incurs
-in a compile time penalty for the first run that must be taken into
-account.
+Switching backends to \texttt{numba} can improve model runtime by a factor up to x30 or more for big or long running models,
+however incurs in a compile time penalty for the first run that must be taken into account.
 

### 2024-03-21 article.tex
semantic line-break in discussion

@@
 
-In this article, we introduced a suite of Python packages we developed
-for defining and simulating dynamical systems and reaction networks.
-These packages are deeply integrated with
-\acp{IDE}, enabling code analysis tools to identify errors
-prior to execution and assist in refactoring and code completion. We
-adopted standard modern Python syntax to ensure seamless IDE
-integration, supported by the extensive Python community.
+In this article,
+we introduced a suite of Python packages we developed
+for defining and simulating dynamical systems and \acp{CRN}.
+These packages are deeply integrated with \acp{IDE},
+enabling code analysis tools to identify errors prior to execution
+and assist in refactoring and code completion.
+We adopted standard modern Python syntax to ensure seamless IDE integration,
+supported by the extensive Python community.
 
 Our approach differs from previous tools in that both the model
-definition and its compilation into an
-\ac{ODE} function are entirely Python-based. This approach simplifies the
-development of various simulation methods, including performance
-enhancements that exploit specific model structures. Importantly, being
-Python-based does not compromise performance compared to C/C++ tools, as
-the \ac{ODE} functions can be \ac{JIT} compiled using Numba.
+definition and its compilation into an \ac{ODE} function are entirely Python-based.
+This approach simplifies the development of various simulation methods,
+including performance enhancements that exploit specific model structures.
+Importantly, being Python-based does not compromise performance compared to C/C++ tools,
+as the \ac{RHS} functions can be \ac{JIT} compiled using Numba.
 
-The inclusion of \ac{SBML} support facilitates the effortless reuse of models
-created by the systems biology community, along with the vast collection
-of public models hosted in the BioModels repository. The modular
-architecture of these packages facilitates their reuse, enhancement, and
-extension by the wider Python community. For instance, an individual
-from outside the systems biology field could contribute a stochastic
-integrator to poincaré, which would then be available in SimBio. This
-clear separation of concerns also makes the packages more
-comprehensible, lowering the barrier for contributing improvements or
-new features. Such an architecture ensures their maintainability and
-ongoing development well into the future.
+The inclusion of \ac{SBML} support facilitates the effortless reuse of models created by the systems biology community,
+along with the vast collection of public models hosted in the BioModels repository.
+The modular architecture of these packages facilitates their reuse, enhancement, and extension by the wider Python community.
+For instance, an individual from outside the systems biology field could contribute a stochastic integrator to poincaré,
+which would then be available in SimBio.
+This clear separation of concerns also makes the packages more comprehensible,
+lowering the barrier for contributing improvements or new features.
+Such an architecture ensures their maintainability and ongoing development well into the future.
 

### 2024-03-21 article.tex
Remove mention of stoichiometry in Species

@@
 
-For the reaction networks, our focus is on first-order differential
-equations that describe the rate of change of species. SimBio simplifies
-the definition of these network models by introducing \texttt{Species},
-which consists of a \texttt{poincare.Variable} and a stoichiometric
-number, and \texttt{RateLaw}, a construct that converts reactants into
-products taking into account the stoichiometry
-(Figure~\ref{fig-simbio}). Additionally, SimBio features
-\texttt{MassAction}, a subclass of \texttt{RateLaw}, that intuitively
-incorporates reactants into the rate law (Figure~\ref{fig-simbio}).
+For the \acp{CRN},
+our focus is on first-order differential equations that describe the rate of change of species.
+SimBio simplifies the definition of these network models by introducing \texttt{Species},
+and \texttt{RateLaw}, a construct that converts reactant species into product species taking into account the stoichiometry (Figure~\ref{fig-simbio}).
+Additionally, SimBio features \texttt{MassAction},
+a subclass of \texttt{RateLaw},
+that intuitively incorporates reactants and their stoichiometry into the rate law (Figure~\ref{fig-simbio}).
 

### 2024-03-21 article.tex
Remove mention of SBML export which is not yet implemented

@@
 
-Several commonly used reactions are predefined as \texttt{MassAction}
-subclasses, such as \texttt{MichaelisMenten}
-(\(S + E \leftrightarrow ES \rightarrow P + E\)) and its approximate
-form without the intermediate species \(ES\), and it is also simple to
-implement your own as subclasses of \texttt{RateLaw} or
-\texttt{MassAction}. Additionally, SimBio supports importing from and
-exporting to \ac{SBML}, and downloading them directly from BioModels
-\cite{malik-sheriffBioModels15Years2020} (Figure~\ref{fig-simbio-io}).
+Several commonly used reactions are predefined as \texttt{MassAction} subclasses,
+such as \texttt{MichaelisMenten} (\(S + E \leftrightarrow ES \rightarrow P + E\))
+and its approximate form without the intermediate species \(ES\),
+and it is also simple to implement used-defined ones
+as subclasses of \texttt{RateLaw} or \texttt{MassAction}.
+Additionally, SimBio supports importing models from \ac{SBML},
+and downloading them directly from BioModels \cite{malik-sheriffBioModels15Years2020} (Figure~\ref{fig-simbio-io}).
 

### 2024-03-21 article.tex
Reproducibility: mention SBML test suite

@@
 
-To evaluate SimBio's reproducibility, we analyzed curated \ac{SBML} models
-from BioModels \cite{malik-sheriffBioModels15Years2020}. Among the first 100 curated
-models from BioModels we selected 60 that did not contains events, as
-SimBio doesn't support events yet. We simulated the selected models with
-COPASI and used the simulated results as ground truth, and demonstrated
-that SimBio reproduces the results with minor numerical differences
-depending on the solver tolerance.
+To evaluate SimBio's reproducibility,
+we used the \ac{SBML} test suite \cite{SBMLTestSuite},
+which provides a set of \ac{SBML} models
+and the expected result of a simulation.
+Excluding models that use not yet supported \ac{SBML} feature,
+every simulation returned correct results within the solver tolerances.
 
-For performance testing, we ran simulations using COPASI,
-Tellurium/RoadRunner, and SimBio. Within SimBio, both NumPy and Numba
-backends were considered. The LSODA solver was used for COPASI and
-SimBio, while for Tellurium the comparable CVODE solver was used. In all
-cases, we used relative and absolute tolerances of \(10^{-6}\). We
-measured three simulation stages: loading, the initial (cold) run, and
-subsequent (warm) runs for each model.
+To evaluate SimBio's performance,
+we used curated \ac{SBML} models from BioModels \cite{malik-sheriffBioModels15Years2020}.
+Among the first 250 curated models,
+we considered the 117 that used supported \ac{SBML} features.
+We ran simulations using COPASI, Tellurium/RoadRunner, and SimBio.
+Within SimBio, both NumPy and Numba
+backends were considered.
+The LSODA solver was used for COPASI and SimBio,
+while for Tellurium the comparable CVODE solver was used.
+In all cases, we used relative and absolute tolerances of \(10^{-6}\).
+We measured three simulation stages:
+loading, the initial (cold) run, and subsequent (warm) runs for each model.
 
-For COPASI and Tellurium/RoadRunner, we noted that its runtime depended
-on the number of evaluation points, something that does not seem to
-happen with SimBio (Figure~\ref{fig-runtime}, left). While SimBio's
-NumPy backend is slower than both COPASI and RoadRunner, we obtained an
-order of magnitude speed-up using the numba backend putting it on par
-with them. A user might have to consider the trade-off between
-compilation and run times, as the compilation of the
-\ac{RHS} code might take longer than the runtime itself, and not be worth
-it for running the model only once. Another speed-up in the runtime can
-be had by switching the LSODA \texttt{scipy} solver for a more efficient
-\texttt{numbalsoda} implementation that avoids the Python interpreter
-between each of the integration steps. This last combination beats all
-other methods, which is also true for the other models we tested
-(Figure~\ref{fig-runtime}, right).
+For COPASI and Tellurium/RoadRunner,
+we noted that its runtime depended on the number of evaluation points,
+something that does not seem to happen with SimBio (Figure~\ref{fig-runtime}, left).
+While SimBio's NumPy backend is slower than both COPASI and RoadRunner,
+we obtained an order of magnitude speed-up using the numba backend
+putting it on par with them.
+A user might have to consider the trade-off between compilation and run times,
+as the compilation of the \ac{RHS} code might take longer than the runtime itself,
+and not be worth it for running the model only once.
+Another speed-up in the runtime can be had by switching the LSODA \texttt{scipy} solver
+for a more efficient \texttt{numbalsoda} implementation that avoids the Python interpreter
+between each of the integration steps.
+This last combination beats all other methods,
+which is also true for the other models we tested (Figure~\ref{fig-runtime}, right).
 

### 2024-03-22 article.tex
Update explanation of performance evaluation

@@
 and the expected result of a simulation.
-Excluding models that use not yet supported \ac{SBML} feature,
+Excluding models that use \ac{SBML} features not yet supported by SimBio,
 every simulation returned correct results within the solver tolerances.
@@
 To evaluate SimBio's performance,
-we used curated \ac{SBML} models from BioModels \cite{malik-sheriffBioModels15Years2020}.
-Among the first 250 curated models,
+we selected \ac{SBML} models from the curated section of BioModels \cite{malik-sheriffBioModels15Years2020}.
+Among the first 250 models,
 we considered the 117 that used supported \ac{SBML} features.
-We ran simulations using COPASI, Tellurium/RoadRunner, and SimBio.
-Within SimBio, both NumPy and Numba
-backends were considered.
-The LSODA solver was used for COPASI and SimBio,
-while for Tellurium the comparable CVODE solver was used.
-In all cases, we used relative and absolute tolerances of \(10^{-6}\).
-We measured three simulation stages:
-loading, the initial (cold) run, and subsequent (warm) runs for each model.
+We ran simulations using
+COPASI (with BasiCO) with the LSODA solver,
+RoadRunner (with Tellurium) with the comparable CVODE solver, and
+SimBio with the LSODA solver.
+For SimBio, we considered three variants:
+NumPy backend and scipy's LSODA solver,
+Numba backend and scipy's LSODA solver,
+Numba backend and numbalsoda's LSODA solver.
+In all cases, we used a relative tolerance of \(10^{-9}\) and absolute tolerances of \(10^{-9}\).
+We measured two simulation stages:
+an initial cold run that includes the reading of the \ac{SBML} model
+and subsequent warm runs.
 
-For COPASI and Tellurium/RoadRunner,
+For COPASI and Tellurium,
 we noted that its runtime depended on the number of evaluation points,

### 2024-04-03 article.tex
Remove performance mention before performance section
Move figures to the end
Add software and hardware (yoda)

@@
 \documentclass{article}
-\usepackage[a4paper, margin=3cm, bottom=4cm]{geometry}
+\usepackage[a4paper, margin=3cm]{geometry}
 
@@
 
-\input{latex/figure1.tex}
-
 Another challenge is the way models are defined.
@@
 
-\input{latex/figure2.tex}
-
 Modular code architecture makes code reusable, extensible, and easier to maintain.
@@
 Poincaré}\label{versatile-modelling-and-simulation-of-dynamical-systems-with-poincaruxe9}}
-
-\input{latex/figure3.tex}
 
@@
 
-\input{latex/figure4.tex}
-
 For this last point,
@@
 
-Switching backends to \texttt{numba} can improve model runtime by a factor up to x30 or more for big or long running models,
-however incurs in a compile time penalty for the first run that must be taken into account.
-
 \hypertarget{extensible-definition-of-reaction-networks-using-simbio}{%
@@
 
-\input{latex/figure5.tex}
-
 \hypertarget{reproducibility-and-performance}{%
@@
 performance}\label{reproducibility-and-performance}}
-
-\input{latex/figure6.tex}
 
@@
 we considered the 117 that used supported \ac{SBML} features.
-We ran simulations using
-COPASI (with BasiCO) with the LSODA solver,
-RoadRunner (with Tellurium) with the comparable CVODE solver, and
-SimBio with the LSODA solver.
+We ran simulations
+on a Intel Xeon CPU E5-2620 v4
+using
+COPASI v4.42.284
+(with BasiCO v0.64) with the LSODA solver,
+RoadRunner v2.5.0 (with Tellurium v2.2.10) with the comparable CVODE solver, and
+SimBio v0.3.2 with the LSODA solver.
 For SimBio, we considered three variants:
-NumPy backend and scipy's LSODA solver,
-Numba backend and scipy's LSODA solver,
+NumPy (v1.26.4) backend and scipy (v1.12)'s LSODA solver,
+Numba (v0.59.0) backend and scipy's LSODA solver,
 Numba backend and numbalsoda's LSODA solver.
-In all cases, we used a relative tolerance of \(10^{-9}\) and absolute tolerances of \(10^{-9}\).
+In all cases, we used a relative tolerance of \(10^{-6}\) and absolute tolerances of \(10^{-9}\).
 We measured two simulation stages:
@@
 For COPASI and Tellurium,
-we noted that its runtime depended on the number of evaluation points,
-something that does not seem to happen with SimBio (Figure~\ref{fig-runtime}, left).
-While SimBio's NumPy backend is slower than both COPASI and RoadRunner,
-we obtained an order of magnitude speed-up using the numba backend
-putting it on par with them.
+we noted that its runtime depended on the number of intermediate evaluation points (Figure~\ref{fig-runtime}, left).
+For SimBio,
+its NumPy backend can be orders of magnitude slower than both COPASI and RoadRunner  (Figure~\ref{fig-runtime}, right).
+Nevertheless, switching to the numba backend,
+which \ac{JIT} compiles the \ac{RHS} equations,
+puts it on par with them.
+Another speed-up in the runtime can be had by switching the LSODA \texttt{scipy} solver
+for a more efficient \texttt{numbalsoda} implementation that avoids calling into the Python interpreter
+between each of the integration steps.
 A user might have to consider the trade-off between compilation and run times,
@@
 and not be worth it for running the model only once.
-Another speed-up in the runtime can be had by switching the LSODA \texttt{scipy} solver
-for a more efficient \texttt{numbalsoda} implementation that avoids the Python interpreter
-between each of the integration steps.
-This last combination beats all other methods,
-which is also true for the other models we tested (Figure~\ref{fig-runtime}, right).
 
@@
 
+\newpage
+
+\input{latex/figure1.tex}
+\input{latex/figure2.tex}
+\input{latex/figure3.tex}
+\input{latex/figure4.tex}
+\input{latex/figure5.tex}
+\input{latex/figure6.tex}
+
 \end{document}

### 2024-04-03 latex/figure6.tex
Use vector file for figure 6
Add software and hardware (yoda)

@@
 
-{\centering \includegraphics{src/performance/figures/performance.png}
+{\centering \includegraphics{src/performance/figures/performance}
 
@@
 
-\caption{\label{fig-runtime}Performance of different softwares to solve
-models from the curated section of BioModels. (left) Run time for the
-model BIOMD3 as a function of the number of output points. (right) Run
-time for different models for 300 output points, using the geometric
-mean of the different softwares to order them.}
+\caption{
+    \label{fig-runtime}
+    Performance of different softwares to solve models from the curated section of BioModels.
+    (left) Run time for the model BIOMD3 as a function of the number of output points.
+    (right) Run time for different models for 300 output points,
+    using the geometric mean of the different softwares to order them.
+    Each point corresponds to the median of 20 runs,
+    with a neglibible errorbar given by the interquartile range.
+}
 

### 2024-04-03 src/performance/plot.py
Update plot

@@ -1,39 +1,95 @@
+from dataclasses import dataclass
 from pathlib import Path
 
+import matplotlib.pyplot as plt
 import pandas as pd
 from matplotlib.figure import Figure
+from scipy.stats import iqr
 
 
-def load(file):
-    return pd.read_csv(file).drop(columns="t_end").set_index("model_id")
+@dataclass
+class Metadata:
+    name: str
+    color: str
 
 
-def plot_number_of_points(fig: Figure, df: dict[str, pd.DataFrame], model_id: str):
+def load_and_summarize(p: Path):
+    def load_one_and_melt(p: Path):
+        return (
+            pd.read_csv(p, index_col=0)
+            .drop(columns="t_end")
+            .assign(cold=lambda x: x.load + x.first_run)
+            .melt(["model_id"])
+            .assign(method=p.parent.stem)
+        )
+
+    dfs = pd.concat(map(load_one_and_melt, p.rglob("*.csv")))
+    return dfs.groupby(["variable", "method", "model_id"])["value"].agg(["median", iqr])
+
+
+def plot_number_of_points(
+    fig: Figure,
+    df: pd.DataFrame,
+    model_id: str,
+    metadata: dict[str, Metadata],
+):
+    df_model = (
+        df.reset_index()
+        .set_index("model_id")
+        .loc[model_id]
+        .set_index("variable")
+        .drop(index=["cold", "first_run", "load"])
+        .reset_index()
+        .astype({"variable": float})
+        .sort_values("variable")
+    )
+
     ax = fig.add_subplot()
     ax.set_xlabel("Number of evaluated time points")
     ax.set_ylabel("Run time [s]")
-    ax.set_ylim(1e-4, 2e-2)
+    ax.set_xlim(5e0, 5e3)
+    ax.set_ylim(5e-4, 5e-2)
+    ax.set_xscale("log")
+    ax.set_yscale("log")
 
-    times = (
-        pd.DataFrame(v.loc[model_id].rename(k) for k, v in df.items())
-        .drop(columns=["load", "first_run"])
-        .T
-    )
-    times.index = times.index.astype(float)
-    times.plot(ax=ax, logx=True, logy=True, marker="o", legend=False)
+    for method, data in df_model.groupby("method"):
+        ax.errorbar(
+            x=data["variable"],
+            y=data["median"],
+            yerr=data["iqr"],
+            fmt="--o",
[…]

### 2024-04-03 article.tex
Add compilation time
Change hardware to Mac M2 pro

@@
 We ran simulations
-on a Intel Xeon CPU E5-2620 v4
+on a MacBook Pro with M2 Pro CPU
 using
@@
 and not be worth it for running the model only once.
+For the models considered,
+amortizing the compilation time required from 2 upto 200 runs.
 

### 2024-04-03 latex/figure5.tex
Simplify code boxes
Rename shaded code environments
Rename load_model to load in biomodels

@@
 
-\begin{Shaded}
+\begin{CodeInput}
 \begin{Highlighting}[]
@@
 \VariableTok{model }\OperatorTok{=}\ClassTok{ sbml}.\FunctionTok{load}\KeywordTok{(}\StringTok{"repressilator.sbml"}\KeywordTok{)}  \CommentTok{\# from existing SBML file}
-\VariableTok{model }\OperatorTok{=}\ClassTok{ biomodels}.\FunctionTok{load\_model}\KeywordTok{(}\StringTok{"BIOMD12"}\KeywordTok{)}  \CommentTok{\# from BioModels}
+\VariableTok{model }\OperatorTok{=}\ClassTok{ biomodels}.\FunctionTok{load}\KeywordTok{(}\StringTok{"BIOMD12"}\KeywordTok{)}  \CommentTok{\# from BioModels}
 \VariableTok{model}
 \end{Highlighting}
-\end{Shaded}
+\end{CodeInput}
 
+\begin{CodeOutput}
     \begin{longtable}{@{}lll@{}}
@@
     \end{longtable}
+\end{CodeOutput}
 

### 2024-04-03 article.tex
Fix abstract acronyms

@@
 \acrodef{RHS}[RHS]{right-hand side}
-\acrodef{SBML}[SBML]{System Biology Markup Language}
+\acrodef{SBML}[SBML]{Systems Biology Markup Language}
 \acrodef{SDE}[SDE]{stochastic differential equation}
@@
 \begin{abstract}
-\Acf{CRN} play a pivotal role in diverse fields
-such as systems biology, biochemistry, chemical engineering, and
-epidemiology. High-level modelling of \acp{CRN} enables various simulation
-approaches, including deterministic and stochastic methods. However,
-existing Python tools for \ac{CRN} modelling typically wrap external C/C++
-libraries for modelling and simulation, limiting their extensibility and
-integration with the broader Python ecosystem. In response, we developed
-Poincaré and SimBio, two novel Python packages for the definition and
-simulation of dynamical systems and \acp{CRN}. Poincaré serves as a
-foundation for dynamical system modelling, while SimBio extends this
-functionality to \acp{CRN}, including support for the 
-SBML. Poincaré and SimBio are developed as pure Python
-packages enabling users to easily extend their simulation capabilities
-by writing new or leveraging other Python packages. Moreover, this does
-not compromise the performance, as code can be Just-In-Time compiled
-with Numba. Our benchmark tests using curated models from the BioModels
-repository demonstrate that these tools may provide a potentially
+\Acfp{CRN} play a pivotal role in diverse fields
+such as systems biology, biochemistry, chemical engineering, and epidemiology.
+High-level modelling of \acp{CRN} enables various simulation approaches,
+including deterministic and stochastic methods.
+However,
+existing Python tools for \ac{CRN} modelling typically wrap external C/C++ libraries for modelling and simulation,
+limiting their extensibility and integration with the broader Python ecosystem.
+In response, we developed Poincaré and SimBio,
+two novel Python packages for the definition and simulation of dynamical systems and \acp{CRN}.
+Poincaré serves as a foundation for dynamical systems modelling,
+while SimBio extends this functionality to \acp{CRN},
+including support for the \ac{SBML}.
+Poincaré and SimBio are developed as pure Python packages
+enabling users to easily extend their simulation capabilities
+by writing new or leveraging other Python packages.
+Moreover, this does not compromise the performance,
+as code can be \ac{JIT} compiled with Numba.
+Our benchmark tests using curated models from the BioModels repository
+demonstrate that these tools may provide a potentially
 superior performance advantage compared to other existing tools.
-Additionally, to ensure a user-friendly experience, our packages use
-standard typed modern Python syntax that provides a seamless integration
-with \acp{IDE}. Our Python-centric approach
-significantly enhances code analysis, error detection, and refactoring
-capabilities, positioning Poincaré and SimBio as valuable tools for the
-modelling community.
+Additionally, to ensure a user-friendly experience,
+our packages use standard typed modern Python syntax that provides a seamless integration with \acp{IDE}.
+Our Python-centric approach significantly enhances code analysis, error detection, and refactoring capabilities,
+positioning Poincaré and SimBio as valuable tools for the modelling community.
 \end{abstract}

### 2024-04-03 article.tex
Rewrite biosimulators and PYthon

@@
 a registry of simulation tools,
-lists at least 15 Python software including
+lists at least 15 softwares categorized under Python including
 COPASI \cite{hoopsCOPASICOmplexPAthway2006},

### 2024-04-03 article.tex
Add name and cite to software

@@
 that is widely used for its user-friendly interface and comprehensive features.
-It also includes Python bindings \cite{bergmannBASICOSimplifiedPython2023b} that allow advanced scripting.
+It also includes Python bindings (BASICO \cite{bergmannBASICOSimplifiedPython2023b}) that allow advanced scripting.
 Tellurium is a Python-based modeling environment
-that uses a C++ library called libRoadRunner in the backend to translate and solve models.
+that uses a C++ library called libRoadRunner \cite{welshLibRoadRunnerHighPerformance2023} in the backend to translate and solve models.
 PySB is a Python library that created a \ac{DSL} using standard Python to define models,

### 2024-04-04 latex/figure2.tex
Update IDE figures

@@
 \begin{minipage}[b]{0.50\linewidth}
-
-{\centering
-
+{
+    \centering 
     \includegraphics{src/ide/ide1.pdf}
-
 }
-
 \end{minipage}%
@@
 \begin{minipage}[b]{0.50\linewidth}
+{
+    \centering
+    \includegraphics{src/ide/ide2.pdf}
+}
+\end{minipage}%
 
-{\centering
-
-\includegraphics{src/ide/ide2.pdf}
-
+\caption{
+    \label{fig-ide}
+    Screenshots of Visual Studio Code showing
+    tooltips (green arrows) and
+    highlighted type errors (red arrows).
+    On the left,
+    we show that \texttt{a},
+    a \texttt{Constant} assigned with \texttt{assign(...,\ constant=True)},
+    can be used for \texttt{Variable b}'s initial condition.
+    Instead,
+    it is flagged as a type error (red underlining)
+    when using \texttt{c}, a \texttt{Parameter},
+    for \texttt{Variable d}'s initial condition,
+    The IDE automatically recognizes \texttt{e} as an \texttt{Equation},
+    and provides autocompletion of the \texttt{Model}'s components.
+    A tooltip is shown when composing models (green arrow, right),
+    which show the expected variables and their default values.
+    The IDE highlights wrong names (\texttt{z} is not a name in \texttt{Model})
+    and mismatched types (\texttt{x} is \texttt{Variable} and \texttt{a} must be a number or a \texttt{Constant})
 }
 
-\end{minipage}%
-
-\caption{\label{fig-ide}Screenshots of Visual Studio Code showing
-tooltips and highlighted type errors. \texttt{Constant} are assigned
-with \texttt{assign(...,\ constant=True)} and can be used to link
-\texttt{Variable}s initial conditions, while trying to use a
-\texttt{Parameter} instead is flagged as type errors (red underlining).
-The IDE automatically recognizes \texttt{e} as an \texttt{Equation}, and
-provides autocompletion of the variables. A tooltip is shown when
-composing models, which show the expected variables and their default
-values. The IDE highlights wrong names (\texttt{z} is not a name in
-\texttt{Model}) and mismatched types (\texttt{x} is \texttt{Variable}
-and \texttt{a} must be a number or a \texttt{Constant})}
-
 \end{figure}

### 2024-04-04 article.tex
Mention dataclass

@@
 Poincare is a package to define and simulate dynamical systems.
-It provides a \texttt{System} class,
+Based on Python inmutable dataclasses,
+it provides a \texttt{System} class,
 where one can define
@@
 Within the constraints imposed by Python's current typing and static analyzers,
-we define models utilizing Python classes
+we define models utilizing Python (data)classes
 such that we can benefit from \acp{IDE}' autocomplete and refactoring capabilities.

### 2024-04-04 article.tex
Poincaré currently limited to ODE

@@
 \acrodef{SDE}[SDE]{stochastic differential equation}
+\acrodef{DDE}[DDE]{delay differential equation}
 \acrodef{TPU}[TPU]{tensor processing unit}
@@
 that translates the model into the \ac{RHS} equations
-and interfaces with \ac{ODE} solvers.
+and interfaces with solvers
+wrapping the output in a \texttt{pandas.DataFrame},
+which can be easily plotted with the standard \texttt{plot} method.
+Currently,
+it only supports translating into first-order \acp{ODE},
+but it would be possible to add support for \acp{SDE} or \acp{DDE}.
 By default,
-it creates a first-order \ac{ODE} system using \texttt{numpy} as a backend,
+it uses \texttt{numpy} as a backend,
 and uses the LSODA solver from \texttt{scipy}.
-This can be easily switched to other solvers.
-The Simulator wraps the output in a \texttt{pandas.DataFrame},
-which can be easily plotted with the standard \texttt{plot} method.
+This can be easily switched to other solvers or backends, such as \texttt{numba}.
 

### 2024-04-04 article.tex
Add unsupported SBML features
Mention independent variable

@@
 and create equations linking a variable's derivative with an expression (Figure~\ref{fig-first-order}).
-It also allows to define higher-order systems
-by assigning an initial condition to a \texttt{Derivative} (Figure~\ref{fig-second-order}).
+It also allows to
+create an \texttt{Independent} variable to define non-autonomous systems,
+and define higher-order systems by assigning an initial condition to a \texttt{Derivative} (Figure~\ref{fig-second-order}).
 
@@
 and downloading them directly from BioModels \cite{malik-sheriffBioModels15Years2020} (Figure~\ref{fig-simbio-io}).
+Work is in progress to support exporting to \ac{SBML} and add more \ac{SBML} features.
+Currently \ac{SBML} unsupported features include
+algebraic rules, constraints, events, reactions with the fast attribute or with math stoichiometry,
+units and compartments with size different from 1.
 

### 2024-04-04 article.tex
Rerun on Macbook Air M2

@@
 We ran simulations
-on a MacBook Pro with M2 Pro CPU
-using
+on a MacBook Air with M2 CPU
+using Python 3.11.8,
 COPASI v4.42.284
-(with BasiCO v0.64) with the LSODA solver,
+(with BasiCO v0.58) with the LSODA solver,
 RoadRunner v2.5.0 (with Tellurium v2.2.10) with the comparable CVODE solver, and
@@
 For SimBio, we considered three variants:
-NumPy (v1.26.4) backend and scipy (v1.12)'s LSODA solver,
+NumPy (v1.26.4) backend and scipy (v1.12.0)'s LSODA solver,
 Numba (v0.59.0) backend and scipy's LSODA solver,
-Numba backend and numbalsoda's LSODA solver.
+Numba backend and numbalsoda's (v0.3.5) LSODA solver.
 In all cases, we used a relative tolerance of \(10^{-6}\) and absolute tolerances of \(10^{-9}\).

### 2024-04-04 src/performance/plot.py

@@ -48,7 +48,7 @@ def plot_number_of_points(
     ax.set_xlabel("Number of evaluated time points")
     ax.set_ylabel("Run time [s]")
     ax.set_xlim(5e0, 5e3)
-    ax.set_ylim(5e-4, 5e-2)
+    ax.set_ylim(1e-4, 2e-2)
     ax.set_xscale("log")
     ax.set_yscale("log")
 
@@ -87,6 +87,7 @@ def plot_runtime(
             y=data["median"],
             yerr=data["iqr"],
             capsize=1,
+            capthick=3,
             linestyle="",
             color=metadata[method].color,
         )

### 2024-04-04 src/performance/timer.py
Rename file

@@ -122,7 +122,7 @@ if __name__ == "__main__":
     path = Path(__file__).parent
     output_path = path / "results" / sys.argv[1]
     output_path.mkdir(exist_ok=True, parents=True)
-    models = pd.read_csv(path / "times.txt")
+    models = pd.read_csv(path / "models.csv")
     for _, (model_num, log_time) in tqdm(models.iterrows(), total=len(models)):
         model_id = f"BIOMD{model_num}"
         model_path = (output_path / model_id).with_suffix(".csv")

### 2024-04-04 article.tex
Improve abstract

@@
 such as systems biology, biochemistry, chemical engineering, and epidemiology.
-High-level modelling of \acp{CRN} enables various simulation approaches,
-including deterministic and stochastic methods.
+High-level definitions of \acp{CRN} enables to use various simulation approaches,
+including deterministic and stochastic methods,
+from the same model.
 However,
-existing Python tools for \ac{CRN} modelling typically wrap external C/C++ libraries for modelling and simulation,
+existing Python tools for simulation of \ac{CRN} typically wrap external C/C++ libraries for
+model definition, translation into equations and/or numerically solving them,
 limiting their extensibility and integration with the broader Python ecosystem.
 In response, we developed Poincaré and SimBio,
-two novel Python packages for the definition and simulation of dynamical systems and \acp{CRN}.
+two novel Python packages for simulation of dynamical systems and \acp{CRN}.
 Poincaré serves as a foundation for dynamical systems modelling,

### 2024-04-04 article.tex
Format Latex

@@
 
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+\hypertarget{introduction}{
+  \section{Introduction}
+  \label{introduction}
+}
 
@@
 
-\hypertarget{results}{%
-\section{Results}\label{results}}
+\hypertarget{results}{
+  \section{Results}
+  \label{results}
+}
 
@@
 
-\hypertarget{versatile-modelling-and-simulation-of-dynamical-systems-with-poincaruxe9}{%
-\subsection{Versatile modelling and simulation of dynamical systems with
-Poincaré}\label{versatile-modelling-and-simulation-of-dynamical-systems-with-poincaruxe9}}
+\hypertarget{versatile-modelling-and-simulation-of-dynamical-systems-with-poincaruxe9}{
+  \subsection{Versatile modelling and simulation of dynamical systems with Poincaré}
+  \label{versatile-modelling-and-simulation-of-dynamical-systems-with-poincaruxe9}
+}
 
@@
 
-\hypertarget{extensible-definition-of-reaction-networks-using-simbio}{%
-\subsection{Extensible definition of reaction networks using
-SimBio}\label{extensible-definition-of-reaction-networks-using-simbio}}
+\hypertarget{extensible-definition-of-reaction-networks-using-simbio}{
+  \subsection{Extensible definition of reaction networks usingSimBio}
+  \label{extensible-definition-of-reaction-networks-using-simbio}
+}
 
@@
 
-\hypertarget{reproducibility-and-performance}{%
-\subsection{Reproducibility and
-performance}\label{reproducibility-and-performance}}
+\hypertarget{reproducibility-and-performance}{
+  \subsection{Reproducibility andperformance}
+  \label{reproducibility-and-performance}
+}
 
@@
 
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
+\hypertarget{discussion}{
+  \section{Discussion}
+  \label{discussion}
+}
 
@@
 
-\hypertarget{resources}{%
-\section{Resources}\label{resources}}
+\hypertarget{resources}{
+  \section{Resources}
+  \label{resources}
+}
 

### 2024-04-04 article.tex
Fix typo

@@
 and, therefore, they cannot provide syntax highlighting, code completion, refactoring, and static analysis.
-For Antimony, an extension providing this capabilities was developed for Visual Studio Code,
+For Antimony, an extension providing these capabilities was developed for Visual Studio Code,
 but its maintenance could be a demanding task for the systems biology community.

### 2024-04-08 article.tex
Hernan's changes

@@
 For COPASI and Tellurium,
-we noted that its runtime depended on the number of intermediate evaluation points (Figure~\ref{fig-runtime}, left).
+we noted that its runtime depended on the number of intermediate evaluation points returned to the user (Figure~\ref{fig-runtime}, left).
+We speculate that this is due to memory allocation and data transfer in the Python bindings, 
+as the number of total function evaluations (i.e. including those performed by the integrator steper)
+is around 3000 and therefore much larger than most part of the x-axis.
 For SimBio,
@@
 The modular architecture of these packages facilitates their reuse, enhancement, and extension by the wider Python community.
+Therefore, it should be also easy to integrate with existing infrastructure such a BioSimulators.org  \cite{shaikhBioSimulatorsCentralRegistry2022},
+or combine with other packages like SimService \cite{sego_simservice_2024} to build more complex simulations.
 For instance, an individual from outside the systems biology field could contribute a stochastic integrator to poincaré,
@@
 
-Source code repositories for these packages can be found at
+Source code repositories, example and documentation for these packages can be found at
 \url{https://github.com/maurosilber/poincare} and

### 2024-06-14 oup-authoring-template.tex
Into OUP template

--- 2024-04-08 article.tex
@@
-\documentclass{article}
-\usepackage[a4paper, margin=3cm]{geometry}
+%%
+%% Copyright 2022 OXFORD UNIVERSITY PRESS
+%%
+%% This file is part of the 'oup-authoring-template Bundle'.
+%% ---------------------------------------------
+%%
+%% It may be distributed under the conditions of the LaTeX Project Public
+%% License, either version 1.2 of this license or (at your option) any
+%% later version.  The latest version of this license is in
+%%    http://www.latex-project.org/lppl.txt
+%% and version 1.2 or later is part of all distributions of LaTeX
+%% version 1999/12/01 or later.
+%%
+%% The list of all files belonging to the 'oup-authoring-template Bundle' is
+%% given in the file `manifest.txt'.
+%%
+%% Template article for OXFORD UNIVERSITY PRESS's document class `oup-authoring-template'
+%% with bibliographic references
+%%
 
-\usepackage{amsmath}
-\usepackage[svgnames]{xcolor}
-\usepackage{hyperref}
-\usepackage{graphicx}
-\usepackage{float}
-\usepackage{caption}
-\usepackage{subcaption}
-\usepackage{tcolorbox}
+\documentclass[namedate,numsec,webpdf,modern,large]{oup-authoring-template}
 
-\usepackage{fancyvrb}
-\newenvironment{CodeInput}
-{\begin{tcolorbox}[title=input,boxrule=0pt]}
-{\end{tcolorbox}}
-\newenvironment{CodeOutput}
-{\begin{tcolorbox}[title=output,boxrule=0pt]}
-{\end{tcolorbox}}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.13,1.00}{#1}}
-\newcommand{\ClassTok}[1]{\textcolor[rgb]{0.27,0.56,0.65}{#1}}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.06,0.50}{#1}}
-\newcommand{\ValueTok}[1]{\textcolor[rgb]{0.13,0.57,0.41}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.47,0.37,0.15}{#1}}
-\newcommand{\IDETok}[1]{\textcolor[rgb]{0.58,0.58,0.58}{#1}}
-\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.06,0.50}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.00,0.50,0.00}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.70,0.27,0.27}{#1}}
+%\onecolumn % for one column layouts
 
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs}
-% Correct order of tables after \paragraph or \subparagraph
+%\usepackage{showframe}
 
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\graphicspath{{Fig/}}
 
-\hypersetup{
-  pdftitle={Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems.},
-  colorlinks=true,
-  linkcolor={blue},
-  filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={Blue},
-}
+% line numbers
+%\usepackage[mathlines, switch]{lineno}
+%\usepackage[right]{lineno}
+
+\theoremstyle{thmstyleone}%
+\newtheorem{theorem}{Theorem}%  meant for continuous numbers
+%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
+%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
+\newtheorem{proposition}[theorem]{Proposition}%
+%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
+\theoremstyle{thmstyletwo}%
+\newtheorem{example}{Example}%
+\newtheorem{remark}{Remark}%
+\theoremstyle{thmstylethree}%
+\newtheorem{definition}{Definition}
 
@@
 
-\usepackage{authblk}
-\title{Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems.}
-\author[1,2]{Mauro Silberberg}
-\author[3]{Henning Hermjakob}
-\author[3]{Rahuman S. Malik-Sheriff}
-\author[1,2]{Hernán E. Grecco}
-\affil[1]{Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Física. Buenos Aires, Argentina.}
-\affil[2]{CONICET - Universidad de Buenos Aires, Instituto de Física de Buenos Aires (IFIBA). Buenos Aires, Argentina}
-\affil[3]{European Bioinformatics Institute, European Molecular Biology Laboratory (EMBL-EBI), Wellcome Genome Campus, Cambridge, UK}
+\usepackage{tcolorbox}
+\newenvironment{CodeInput}{\begin{tcolorbox}[title=input,boxrule=0pt]}{\end{tcolorbox}}
+\newenvironment{CodeOutput}{\begin{tcolorbox}[title=output,boxrule=0pt]}{\end{tcolorbox}}
+
+\usepackage{fancyvrb}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.13,1.00}{#1}}
+\newcommand{\ClassTok}[1]{\textcolor[rgb]{0.27,0.56,0.65}{#1}}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.06,0.50}{#1}}
+\newcommand{\ValueTok}[1]{\textcolor[rgb]{0.13,0.57,0.41}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.47,0.37,0.15}{#1}}
+\newcommand{\IDETok}[1]{\textcolor[rgb]{0.58,0.58,0.58}{#1}}
+\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.06,0.50}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.00,0.50,0.00}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.70,0.27,0.27}{#1}}
 
@@
 
-\maketitle
+\journaltitle{Journal Title Here}
+\DOI{DOI HERE}
+\copyrightyear{2022}
+\pubyear{2019}
+\access{Advance Access Publication Date: Day Month Year}
+\appnotes{Paper}
 
-\begin{abstract}
+\firstpage{1}
+
+%\subtitle{Subject Section}
+
+\title[Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems]{Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems}
+
+\author[1,2]{Mauro Silberberg\ORCID{0000-0000-0000-0000}}
+\author[3]{Henning Hermjakob\ORCID{0000-0000-0000-0000}}
+\author[3]{Rahuman S. Malik-Sheriff\ORCID{0000-0000-0000-0000}}
+\author[1,2,$\ast$]{Hernán E. Grecco\ORCID{0000-0000-0000-0000}}
+
+\authormark{Author Name et al.}
+
+\address[1]{\orgname{Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Física}, \orgaddress{\state{Buenos Aires}, \country{Argentina}}}
+\address[2]{\orgname{CONICET - Universidad de Buenos Aires, Instituto de Física de Buenos Aires (IFIBA)}, \orgaddress{\state{Buenos Aires}, \country{Argentina}}}
+\address[3]{\orgdiv{European Bioinformatics Institute}, \orgname{European Molecular Biology Laboratory (EMBL-EBI)}, \orgaddress{\street{Wellcome Genome Campus}, \state{Cambridge}, \country{UK}}}
+
+\corresp[$\ast$]{Corresponding author. \href{email:hgrecco@df.uba.ar}{hgrecco@df.uba.ar}}
+
+\received{Date}{0}{Year}
+\revised{Date}{0}{Year}
+\accepted{Date}{0}{Year}
+
+%\editor{Associate Editor: Name}
+
+%\abstract{
+%\textbf{Motivation:} .\\
+%\textbf{Results:} .\\
+%\textbf{Availability:} .\\
+%\textbf{Contact:} \href{name@email.com}{name@email.com}\\
+%\textbf{Supplementary information:} Supplementary data are available at \textit{Journal Name}
+%online.}
+
+\abstract{
   \Acfp{CRN} play a pivotal role in diverse fields
@@
   positioning Poincaré and SimBio as valuable tools for the modelling community.
-\end{abstract}
-
+}
+\keywords{poincare, SimBio, Python, ODE}
 \acresetall % reset acronyms after abstract
 
-\hypertarget{introduction}{
+% \boxedtext{
+% \begin{itemize}
+% \item Key boxed text here.
+% \item Key boxed text here.
+% \item Key boxed text here.
+% \end{itemize}}
+
+\maketitle
+
 \section{Introduction}
-  \label{introduction}
-}
 
@@
 
-\hypertarget{results}{
-  \section{Results}
-  \label{results}
-}
+\section{Results}\label{results}
 
@@
 
-\hypertarget{versatile-modelling-and-simulation-of-dynamical-systems-with-poincaruxe9}{
 \subsection{Versatile modelling and simulation of dynamical systems with Poincaré}
 \label{versatile-modelling-and-simulation-of-dynamical-systems-with-poincaruxe9}
-}
 
@@
 \begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
 \item
@@
 
-\hypertarget{extensible-definition-of-reaction-networks-using-simbio}{
 \subsection{Extensible definition of reaction networks usingSimBio}
 \label{extensible-definition-of-reaction-networks-using-simbio}
-}
 
@@
 
-\hypertarget{reproducibility-and-performance}{
-  \subsection{Reproducibility andperformance}
+\subsection{Reproducibility and performance}
 \label{reproducibility-and-performance}
-}
 
@@
 
-\hypertarget{discussion}{
-  \section{Discussion}
-  \label{discussion}
-}
+\section{Discussion}\label{discussion}
 
@@
 
-\hypertarget{resources}{
-  \section{Resources}
-  \label{resources}
-}
+\section{Resources}\label{resources}
 
@@
 
-\bibliographystyle{plain}
-\bibliography{bibliography.bib}
+%%%%%%%%%%%%%%
 
-\newpage
+\section{Competing interests}
+No competing interest is declared.
 
-\input{latex/figure1.tex}
-\input{latex/figure2.tex}
-\input{latex/figure3.tex}
-\input{latex/figure4.tex}
-\input{latex/figure5.tex}
-\input{latex/figure6.tex}
+\section{Author contributions statement}
+
+Must include all authors, identified by initials, for example:
+S.R. and D.A. conceived the experiment(s),  S.R. conducted the experiment(s), S.R. and D.A. analysed the results.  S.R. and D.A. wrote and reviewed the manuscript.
+
+\section{Acknowledgments}
+The authors thank the anonymous reviewers for their valuable suggestions. This work is supported in part by funds from the National Science Foundation (NSF: \# 1636933 and \# 1920920).
+
+% bibliography
+\bibliographystyle{abbrvnat}
+\bibliography{bibliography}
+
+%% sample for biography with author's image
+% \begin{biography}{{\color{black!20}\rule{77pt}{77pt}}}{\author{Author Name.} This is sample author biography text.
+The values provided in the optional argument are meant for sample purposes.
+There is no need to include the width and height of an image in the optional argument for live articles.
+This is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text.}
+% \end{biography}
+
+%% sample for biography without author's image
+% \begin{biography}{}{\author{Author Name.} This is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text.}
+% \end{biography}
+
+% FIGURE 1
+\begin{figure*}[t]
+
+  \begin{minipage}[t]{\linewidth}
+
+  {\centering
+
+  \begin{CodeInput}
+  \begin{Highlighting}[]
+  \KeywordTok{from} \ClassTok{poincare} \KeywordTok{import} \ClassTok{Derivative}, \ClassTok{System}, \ClassTok{Variable}, \FunctionTok{initial}
+  \end{Highlighting}
+  \end{CodeInput}
+
+  }
+
+  \end{minipage}%
+  \newline
+  \begin{minipage}[t]{\linewidth}
+
+  {\centering
+
+  \begin{minipage}[c]{0.6\linewidth}
+
+  {\centering
+
+  \begin{CodeInput}
+  \begin{Highlighting}[]
+  \KeywordTok{class} \ClassTok{Decay}\KeywordTok{(}\ClassTok{System}\KeywordTok{)}\OperatorTok{:}
+      \VariableTok{x}\OperatorTok{:} \ClassTok{Variable} \OperatorTok{=} \FunctionTok{initial}\KeywordTok{(\VariableTok{default}\OperatorTok{=}\ValueTok{1})}
+      \VariableTok{eq} \OperatorTok{=} \VariableTok{x}.\FunctionTok{derive}\KeywordTok{()} \OperatorTok{\textless{}\textless{}} \OperatorTok{-}\VariableTok{x}
+  \end{Highlighting}
+  \end{CodeInput}
+
+  }
+
+  \end{minipage}%
+  %
+  \begin{minipage}[c]{0.2\linewidth}
+
+  {\centering
+
+  \[
+  \frac{dx}{dt} = -x
+  \]
+
+  }
+
+  \end{minipage}%
+  %
+  \begin{minipage}[c]{0.2\linewidth}
+
+  {\centering
+
+  \[
+  x(0) = 1
+  \]
+
+  }
+
+  \end{minipage}%
+
+  % \subcaption{\label{fig-first-order}First-order system of an exponential decay. The variable
+  % \texttt{x} stores the initial condition for \(x\), and the variable
+  % \texttt{eq} stores the rate equation for \(x\).}
+
+  }
+
+  \end{minipage}%
+  \newline
+  \begin{minipage}[t]{\linewidth}
+
+  {\centering
+
+  \begin{minipage}[c]{0.6\linewidth}
+
+  {\centering
+
+  \begin{CodeInput}
+  \begin{Highlighting}[]
+  \KeywordTok{class} \ClassTok{Oscillator}\KeywordTok{(}\ClassTok{System}\KeywordTok{)}\OperatorTok{:}
+      \VariableTok{x}: \ClassTok{Variable} \OperatorTok{=} \FunctionTok{initial}\KeywordTok{(\VariableTok{default}\OperatorTok{=}\ValueTok{1})}
+      \VariableTok{v}: \ClassTok{Derivative} \OperatorTok{=} \VariableTok{x}.\FunctionTok{derive}\KeywordTok{(\VariableTok{initial}\OperatorTok{=}\ValueTok{0})}
+      \VariableTok{eq} \OperatorTok{=} \VariableTok{v}.\FunctionTok{derive}\KeywordTok{()} \OperatorTok{\textless{}\textless{}} \OperatorTok{-}\VariableTok{x}
+  \end{Highlighting}
+  \end{CodeInput}
+
+  }
+
+  \end{minipage}%
+  %
+  \begin{minipage}[c]{0.2\linewidth}
+
+  {\centering
+
+  \[
+  \frac{d^2x}{dt^2} = -x
+  \]
+
+  }
+
+  \end{minipage}%
+  %
+  \begin{minipage}[c]{0.2\linewidth}
+
+  {\centering
+
+  \[
+  \begin{cases}
+      x(0) &= 1 \\
+      \frac{dx}{dt}(0) &= 0
+  \end{cases}
+  \]
+
+  }
+
+  \end{minipage}%
+
+  % \subcaption{\label{fig-second-order}Second-order system of an harmonic oscillator, where the
+  % variable \texttt{v} stores the derivative \(\frac{dx}{dt}\) and the rate
+  % equation is specified for the derivative \texttt{v}.}
+
+  }
+
+  \end{minipage}%
+  \newline
+  \begin{minipage}[t]{\linewidth}
+
+  {\centering
+
+  \begin{minipage}[c]{0.6\linewidth}
+
+  {\centering
+
+  \begin{CodeInput}
+  \begin{Highlighting}[]
+  \KeywordTok{class} \ClassTok{BigModel}\KeywordTok{(}\ClassTok{System}\KeywordTok{)}:
+      \VariableTok{x}: \ClassTok{Variable} \OperatorTok{=} \FunctionTok{initial}\KeywordTok{(\VariableTok{default}\OperatorTok{=}\ValueTok{1})}
+      \VariableTok{linked} \OperatorTok{=} \ClassTok{Decay}\KeywordTok{(\VariableTok{x}\OperatorTok{=}\VariableTok{x})}
+      \VariableTok{independent} \OperatorTok{=} \ClassTok{Decay}\KeywordTok{(\VariableTok{x}\OperatorTok{=}\ValueTok{2})}
+  \end{Highlighting}
+  \end{CodeInput}
+
+  }
+
+  \end{minipage}%
+  %
+  \begin{minipage}[c]{0.2\linewidth}
+
+  {\centering
+
+  \[
+  \begin{cases}
+      \frac{dx}{dt} = -x \\
+      \frac{dx_{ind}}{dt} = -x_{ind}
+  \end{cases}
+  \]
+
+  }
+
+  \end{minipage}%
+  %
+  \begin{minipage}[c]{0.2\linewidth}
+
+  {\centering
+
+  \[
+  \begin{cases}
+      x(0) &= 1 \\
+      x_{ind}(0) &= 2
+  \end{cases}
+  \]
+
+  }
+
+  \end{minipage}%
+
+  % \subcaption{\label{fig-composition}First-order system of two exponential decays by composition of
+  % the \texttt{Decay} system. The subsystem \texttt{linked} has a reference
+  % to the outer variable \texttt{x}, while the subsystem
+  % \texttt{independent} defines a new variable with initial condition $2$,
+  % which on the corresponding mathematical expression was named \(x_{ind}\).}
+
+  }
+
+  \end{minipage}%
+
+  \caption{\label{fig-poincare}Code and corresponding mathematical
+  expressions for different systems.}
+
+  \end{figure*}
+
+% FIGURE 2
+\begin{figure*}[t]
+
+  \begin{minipage}[b]{0.50\linewidth}
+  {
+      \centering 
+      \includegraphics[width=\textwidth]{src/ide/ide1.pdf}
+  }
+  \end{minipage}%
+  %
+  \begin{minipage}[b]{0.50\linewidth}
+  {
+      \centering
+      \includegraphics[width=\textwidth]{src/ide/ide2.pdf}
+  }
+  \end{minipage}%
+
+  \caption{
+      \label{fig-ide}
+      Screenshots of Visual Studio Code showing
+      tooltips (green arrows) and
+      highlighted type errors (red arrows).
+      On the left,
+      we show that \texttt{a},
+      a \texttt{Constant} assigned with \texttt{assign(...,\ constant=True)},
+      can be used for \texttt{Variable b}'s initial condition.
+      Instead,
+      it is flagged as a type error (red underlining)
+      when using \texttt{c}, a \texttt{Parameter},
+      for \texttt{Variable d}'s initial condition,
+      The IDE automatically recognizes \texttt{e} as an \texttt{Equation},
+      and provides autocompletion of the \texttt{Model}'s components.
+      A tooltip is shown when composing models (green arrow, right),
+      which show the expected variables and their default values.
+      The IDE highlights wrong names (\texttt{z} is not a name in \texttt{Model})
+      and mismatched types (\texttt{x} is \texttt{Variable} and \texttt{a} must be a number or a \texttt{Constant})
+  }
+
+  \end{figure*}
+
+% FIGURE 3
+\begin{figure*}[t]
+
+  \begin{minipage}[t]{\linewidth}
+
+  {\centering
+
+  \begin{CodeInput}
+  \begin{Highlighting}[]
+  \KeywordTok{import} \ClassTok{numpy} \KeywordTok{as} \ClassTok{np}
+  \KeywordTok{from} \ClassTok{poincare} \KeywordTok{import} \ClassTok{Simulator}
+
+  \VariableTok{sim} \OperatorTok{=} \ClassTok{Simulator}\KeywordTok{(\ClassTok{Oscillator})}
+  \VariableTok{df} \OperatorTok{=} \VariableTok{sim}.\FunctionTok{solve}\KeywordTok{(}\VariableTok{save\_at}\OperatorTok{=}\ClassTok{np}.\FunctionTok{linspace}\KeywordTok{(}\ValueTok{0}, \ValueTok{5}, \ValueTok{101}\KeywordTok{))}
+  \end{Highlighting}
+  \end{CodeInput}
+
+  }
+
+  \end{minipage}%
+  \newline
+  \begin{minipage}[t]{0.45\linewidth}
+
+  {\centering
+
+  \begin{CodeInput}
+  \begin{Highlighting}[]
+  \VariableTok{df}.\FunctionTok{head}\KeywordTok{()}
+  \end{Highlighting}
+  \end{CodeInput}
+
+  \begin{CodeOutput}
+      \begin{tabular}[]{@{}lll@{}}
+          \toprule
+          & x & v \\
+          time & & \\
+          \midrule
+          0.00 & 1.000000 & 0.000000 \\
+          0.05 & 0.998750 & -0.049979 \\
+          0.10 & 0.995005 & -0.099835 \\
+          0.15 & 0.988772 & -0.149440 \\
+          0.20 & 0.980069 & -0.198672 \\
+          \botrule
+      \end{tabular}
+  \end{CodeOutput}
+
+  }
+
+  \end{minipage}%
+  \hfill
+  \begin{minipage}[t]{0.45\linewidth}
+
+  {\centering
+
+  \begin{CodeInput}
+  \begin{Highlighting}[]
+  \VariableTok{df}.\FunctionTok{plot}\KeywordTok{()}\OperatorTok{;}
+  \end{Highlighting}
+  \end{CodeInput}
+
+  \begin{CodeOutput}
+    \includegraphics[width=\textwidth]{article_files/cell-9-output-1}
+  \end{CodeOutput}
+
+  }
+
+  \end{minipage}%
+
+  \caption{\label{fig-sim}Simulation of the \texttt{Oscillator} system
+  from Figure~\ref{fig-second-order}. The output is a
+  \texttt{pandas.DataFrame} with a column for each variable and the time
+  as index. It is inspected and plotted with the \texttt{pandas} methods.}
+
+  \end{figure*}
+
+% FIGURE 4
+\begin{figure*}[t]
+
+  \begin{minipage}[t]{\linewidth}
+    \centering 
+    \begin{CodeInput}
+    \begin{Highlighting}[]
+    \KeywordTok{from}\ClassTok{ simbio }\KeywordTok{import}\ClassTok{ Compartment, MassAction, Species, RateLaw, \FunctionTok{initial}}
+    \end{Highlighting}
+    \end{CodeInput}
+  \end{minipage}%
+  \newline
+  \begin{minipage}[t]{\linewidth}
+    \centering 
+    \begin{minipage}[c]{0.60\linewidth}
+      \centering 
+      \begin{CodeInput}
+      \begin{Highlighting}[]
+        \KeywordTok{class}\ClassTok{ Model}\KeywordTok{(}\ClassTok{Compartment}\KeywordTok{)}:
+            \CommentTok{"""2A {-}\textgreater{} B"""}
+
+        \VariableTok{    A}: \ClassTok{Species }\OperatorTok{=}\FunctionTok{ initial}\KeywordTok{(}\VariableTok{default}\OperatorTok{=}\ValueTok{1}\KeywordTok{)}
+        \VariableTok{    B}: \ClassTok{Species }\OperatorTok{=}\FunctionTok{ initial}\KeywordTok{(}\VariableTok{default}\OperatorTok{=}\ValueTok{0}\KeywordTok{)}
+        \VariableTok{    r }\OperatorTok{=}\ClassTok{ RateLaw}\KeywordTok{(}
+        \VariableTok{        reactants}\OperatorTok{=}\KeywordTok{[}\ValueTok{2} \OperatorTok{*}\VariableTok{ A}\KeywordTok{]},
+        \VariableTok{        products}\OperatorTok{=}\NormalTok{\KeywordTok{[}B\KeywordTok{]},}
+        \VariableTok{        rate}\OperatorTok{=}\ValueTok{1}\NormalTok{,}
+        \KeywordTok{    )}
+      \end{Highlighting}
+      \end{CodeInput}
+    \end{minipage}%
+    %
+    \begin{minipage}[c]{0.40\linewidth}
+      \centering 
+      \[
+      \begin{cases}
+          \frac{dA}{dt} = -2 \\
+          \frac{dB}{dt} = +1
+      \end{cases}
+      \begin{cases}
+          A(0) &= 1 \\
+          B(0) &= 0
+      \end{cases}
+      \]
+    \end{minipage}%
+  \end{minipage}%
+  \newline
+  \begin{minipage}[t]{\linewidth}
+      \centering 
+      \begin{minipage}[c]{0.60\linewidth}
+          \centering 
+          \begin{CodeInput}
+          \begin{Highlighting}[]
+          \KeywordTok{class}\ClassTok{ Model}\KeywordTok{(}\ClassTok{Compartment}\KeywordTok{)}:
+              \CommentTok{"""2A {-}\textgreater{} B"""}
+
+          \VariableTok{    A}: \ClassTok{Species }\OperatorTok{=}\FunctionTok{ initial}\KeywordTok{(}\VariableTok{default}\OperatorTok{=}\ValueTok{1}\KeywordTok{)}
+          \VariableTok{    B}: \ClassTok{Species }\OperatorTok{=}\FunctionTok{ initial}\KeywordTok{(}\VariableTok{default}\OperatorTok{=}\ValueTok{0}\KeywordTok{)}
+          \VariableTok{    r }\OperatorTok{=}\ClassTok{ MassAction}\KeywordTok{(}
+          \VariableTok{        reactants}\OperatorTok{=}\KeywordTok{[}\ValueTok{2} \OperatorTok{*}\VariableTok{ A}\KeywordTok{]},
+          \VariableTok{        products}\OperatorTok{=}\NormalTok{\KeywordTok{[}B\KeywordTok{]},}
+          \VariableTok{        rate}\OperatorTok{=}\ValueTok{1}\NormalTok{,}
+          \KeywordTok{    )}
+          \end{Highlighting}
+          \end{CodeInput}
+      \end{minipage}%
+      %
+      \begin{minipage}[c]{0.40\linewidth}
+          \centering 
+          \[
+          \begin{cases}
+              \frac{dA}{dt} = -2 A^2 \\
+              \frac{dB}{dt} = +1 A^2
+          \end{cases}
+          \begin{cases}
+              A(0) &= 1 \\
+              B(0) &= 0
+          \end{cases}
+          \]
+      \end{minipage}%
+  \end{minipage}%
+
+  \caption{
+    A reaction system for species \(A\) and \(B\)
+    with initial conditions \(1\) and \(0\), respectively. A single reaction
+    transforming \(2A\) into \(B\) is saved in variable \texttt{r}. The rate
+    \(1\) is specified directly for \texttt{RateLaw}, and is proportional to
+    the reactants for \texttt{MassAction}.
+  }
+  \label{fig-simbio}
+\end{figure*}
+
+% FIGURE 5
+\begin{figure*}[t]
+  \centering
+
+  \begin{CodeInput}
+  \begin{Highlighting}[]
+  \KeywordTok{from}\ClassTok{ simbio.io }\KeywordTok{import}\ClassTok{ biomodels, sbml}
+
+  \VariableTok{model }\OperatorTok{=}\ClassTok{ sbml}.\FunctionTok{load}\KeywordTok{(}\StringTok{"repressilator.sbml"}\KeywordTok{)}  \CommentTok{\# from existing SBML file}
+  \VariableTok{model }\OperatorTok{=}\ClassTok{ biomodels}.\FunctionTok{load}\KeywordTok{(}\StringTok{"BIOMD12"}\KeywordTok{)}  \CommentTok{\# from BioModels}
+  \VariableTok{model}
+  \end{Highlighting}
+  \end{CodeInput}
+
+  \begin{CodeOutput}
+      \begin{tabular}{@{}lll@{}}
+          \multicolumn{3}{l}{Elowitz2000 - Repressilator} \\
+          \toprule
+          type & total & names \\
+          \midrule
+          variables  &  6 & PX, PY, PZ, X, Y, Z \\
+          parameters & 17 & cell, beta, alpha0, alpha, eff, n, KM, tau\_mRNA, tau\_prot, ... \\
+          equations  & 12 & Reaction1, Reaction2, Reaction3, Reaction4, Reaction5, ... \\
+          \botrule
+      \end{tabular}
+  \end{CodeOutput}
+
+  \caption{Creation of a model from a local SBML file or one uploaded to BioModels.}
+  \label{fig-simbio-io}
+\end{figure*}
+
+% FIGURE 6
+\begin{figure*}[t]
+  \centering
+  \includegraphics[width=\textwidth]{src/performance/figures/performance}
+  \caption{
+    Performance of different softwares to solve models from the curated section of BioModels.
+    (left) Run time for the model BIOMD3 as a function of the number of output points.
+    (right) Run time for different models for 300 output points,
+    using the geometric mean of the different softwares to order them.
+    Each point corresponds to the median of 20 runs,
+    with a neglibible errorbar given by the interquartile range.
+  }
+  \label{fig-runtime}
+\end{figure*}
 

### 2024-06-14 oup-authoring-template.tex
Add subcaption package

@@
 \newtheorem{definition}{Definition}
+
+\usepackage{subcaption}
 
@@
 \begin{figure*}[t]
-
   \begin{minipage}[t]{\linewidth}
@@
 
-  % \subcaption{\label{fig-first-order}First-order system of an exponential decay. The variable
-  % \texttt{x} stores the initial condition for \(x\), and the variable
-  % \texttt{eq} stores the rate equation for \(x\).}
+  \subcaption{\label{fig-first-order}First-order system of an exponential decay. The variable
+  \texttt{x} stores the initial condition for \(x\), and the variable
+  \texttt{eq} stores the rate equation for \(x\).}
 
@@
 
-  % \subcaption{\label{fig-second-order}Second-order system of an harmonic oscillator, where the
-  % variable \texttt{v} stores the derivative \(\frac{dx}{dt}\) and the rate
-  % equation is specified for the derivative \texttt{v}.}
+  \subcaption{\label{fig-second-order}Second-order system of an harmonic oscillator, where the
+  variable \texttt{v} stores the derivative \(\frac{dx}{dt}\) and the rate
+  equation is specified for the derivative \texttt{v}.}
 
@@
 
-  % \subcaption{\label{fig-composition}First-order system of two exponential decays by composition of
-  % the \texttt{Decay} system. The subsystem \texttt{linked} has a reference
-  % to the outer variable \texttt{x}, while the subsystem
-  % \texttt{independent} defines a new variable with initial condition $2$,
-  % which on the corresponding mathematical expression was named \(x_{ind}\).}
+  \subcaption{\label{fig-composition}First-order system of two exponential decays by composition of
+  the \texttt{Decay} system. The subsystem \texttt{linked} has a reference
+  to the outer variable \texttt{x}, while the subsystem
+  \texttt{independent} defines a new variable with initial condition $2$,
+  which on the corresponding mathematical expression was named \(x_{ind}\).}
 

### 2024-06-14 final/oup-authoring-template.tex
Move to final dir
Change cite to citep

--- 2024-06-14 oup-authoring-template.tex
@@
 Several tools already exist to \emph{define}, \emph{translate} and \emph{solve} \acp{CRN}.
-BioSimulators.org \cite{shaikhBioSimulatorsCentralRegistry2022},
+BioSimulators.org \citep{shaikhBioSimulatorsCentralRegistry2022},
 a registry of simulation tools,
 lists at least 15 softwares categorized under Python including
-COPASI \cite{hoopsCOPASICOmplexPAthway2006},
-Tellurium \cite{choiTelluriumExtensiblePythonbased2018} and
-PySB \cite{lopezProgrammingBiologicalModels2013}.
+COPASI \citep{hoopsCOPASICOmplexPAthway2006},
+Tellurium \citep{choiTelluriumExtensiblePythonbased2018} and
+PySB \citep{lopezProgrammingBiologicalModels2013}.
 COPASI is a standalone software with a \ac{GUI}
 that is widely used for its user-friendly interface and comprehensive features.
-It also includes Python bindings (BASICO \cite{bergmannBASICOSimplifiedPython2023b}) that allow advanced scripting.
+It also includes Python bindings, BASICO \citep{bergmannBASICOSimplifiedPython2023b}, that allow advanced scripting.
 Tellurium is a Python-based modeling environment
-that uses a C++ library called libRoadRunner \cite{welshLibRoadRunnerHighPerformance2023} in the backend to translate and solve models.
+that uses a C++ library called libRoadRunner \citep{welshLibRoadRunnerHighPerformance2023} in the backend to translate and solve models.
 PySB is a Python library that created a \ac{DSL} using standard Python to define models,
-which are then translated to \acp{ODE} using a Perl library called BioNetGen \cite{harrisBioNetGenAdvancesRulebased2016}.
+which are then translated to \acp{ODE} using a Perl library called BioNetGen \citep{harrisBioNetGenAdvancesRulebased2016}.
 
@@
 As such, it is not possible to apply custom optimizations to the equations
-or use automatic differentation packages such as JAX \cite{jax2018github} to compute the model's jacobian.
+or use automatic differentation packages such as JAX \citep{jax2018github} to compute the model's jacobian.
 
 Another challenge is the way models are defined.
-Many tools support the \ac{SBML} \cite{huckaSBMLL3V2} as an exchange format,
+Many tools support the \ac{SBML} \citep{huckaSBMLL3V2} as an exchange format,
 a \emph{de facto} standard for \acp{CRN} that defines species, parameters and reactions between species.
 As writing \ac{SBML} directly is impractical,
-Tellurium uses a \ac{DSL} called Antimony \cite{smithAntimonyModularModel2009} for defining models.
+Tellurium uses a \ac{DSL} called Antimony \citep{smithAntimonyModularModel2009} for defining models.
 \Acp{DSL} allows to reuse the same code in different programming environments,
@@
 These are pure Python packages with standard dependencies from the PyData scientific stack
-such as NumPy \cite{harrisArrayProgrammingNumPy2020} and pandas \cite{mckinneyDataStructuresStatistical2010}.
+such as NumPy \citep{harrisArrayProgrammingNumPy2020} and pandas \citep{mckinneyDataStructuresStatistical2010}.
 They are published in the \acf{PyPI},
@@
 Symbolite is a lightweight symbolics package to create algebraic mathematical expressions.
-Unlike SymPy \cite{meurerSymPySymbolicComputing2017},
+Unlike SymPy \citep{meurerSymPySymbolicComputing2017},
 a widely-used Python library for symbolic mathematics,
@@
 Currently, we have implementations for
-NumPy \cite{harrisArrayProgrammingNumPy2020};
-Numba \cite{lamNumbaLLVMbasedPython2015},
+NumPy \citep{harrisArrayProgrammingNumPy2020};
+Numba \citep{lamNumbaLLVMbasedPython2015},
 a \ac{JIT} compiler to LLVM;
-SymPy \cite{meurerSymPySymbolicComputing2017};
-and JAX \cite{jax2018github}, a
+SymPy \citep{meurerSymPySymbolicComputing2017};
+and JAX \citep{jax2018github}, a
 library that support automatic differentiation and compilation to \acp{GPU} and \acp{TPU}.
@@
 For this last point,
-\acp{IDE} that support \texttt{dataclass\_transform} \cite{debontePEP681Data2021}
+\acp{IDE} that support \texttt{dataclass\_transform} \citep{debontePEP681Data2021}
 can provide a tooltip with the expected signature (Figure~\ref{fig-ide}).
@@
 Additionally, SimBio supports importing models from \ac{SBML},
-and downloading them directly from BioModels \cite{malik-sheriffBioModels15Years2020} (Figure~\ref{fig-simbio-io}).
+and downloading them directly from BioModels \citep{malik-sheriffBioModels15Years2020} (Figure~\ref{fig-simbio-io}).
 Work is in progress to support exporting to \ac{SBML} and add more \ac{SBML} features.
@@
 To evaluate SimBio's reproducibility,
-we used the \ac{SBML} test suite \cite{SBMLTestSuite},
+we used the \ac{SBML} test suite \citep{SBMLTestSuite},
 which provides a set of \ac{SBML} models
@@
 To evaluate SimBio's performance,
-we selected \ac{SBML} models from the curated section of BioModels \cite{malik-sheriffBioModels15Years2020}.
+we selected \ac{SBML} models from the curated section of BioModels \citep{malik-sheriffBioModels15Years2020}.
 Among the first 250 models,
@@
 The modular architecture of these packages facilitates their reuse, enhancement, and extension by the wider Python community.
-Therefore, it should be also easy to integrate with existing infrastructure such a BioSimulators.org  \cite{shaikhBioSimulatorsCentralRegistry2022},
-or combine with other packages like SimService \cite{sego_simservice_2024} to build more complex simulations.
+Therefore, it should be also easy to integrate with existing infrastructure such a BioSimulators.org \citep{shaikhBioSimulatorsCentralRegistry2022},
+or combine with other packages like SimService \citep{sego_simservice_2024} to build more complex simulations.
 For instance, an individual from outside the systems biology field could contribute a stochastic integrator to poincaré,
@@
       \centering 
-      \includegraphics[width=\textwidth]{src/ide/ide1.pdf}
+      \includegraphics[width=\textwidth]{ide1.pdf}
   }
@@
       \centering
-      \includegraphics[width=\textwidth]{src/ide/ide2.pdf}
+      \includegraphics[width=\textwidth]{ide2.pdf}
   }
@@
   \begin{CodeOutput}
-    \includegraphics[width=\textwidth]{article_files/cell-9-output-1}
+    \includegraphics[width=\textwidth]{cell-output}
   \end{CodeOutput}
@@
   \centering
-  \includegraphics[width=\textwidth]{src/performance/figures/performance}
+  \includegraphics[width=\textwidth]{performance}
   \caption{

### 2024-06-15 final/thin-figures.tex
Add thin figures variant

--- 2024-06-14 final/oup-authoring-template.tex
@@
-%%
-%% Copyright 2022 OXFORD UNIVERSITY PRESS
-%%
-%% This file is part of the 'oup-authoring-template Bundle'.
-%% ---------------------------------------------
-%%
-%% It may be distributed under the conditions of the LaTeX Project Public
-%% License, either version 1.2 of this license or (at your option) any
-%% later version.  The latest version of this license is in
-%%    http://www.latex-project.org/lppl.txt
-%% and version 1.2 or later is part of all distributions of LaTeX
-%% version 1999/12/01 or later.
-%%
-%% The list of all files belonging to the 'oup-authoring-template Bundle' is
-%% given in the file `manifest.txt'.
-%%
-%% Template article for OXFORD UNIVERSITY PRESS's document class `oup-authoring-template'
-%% with bibliographic references
-%%
-
-\documentclass[namedate,numsec,webpdf,modern,large]{oup-authoring-template}
-
-%\onecolumn % for one column layouts
-
-%\usepackage{showframe}
-
-\graphicspath{{Fig/}}
-
-% line numbers
-%\usepackage[mathlines, switch]{lineno}
-%\usepackage[right]{lineno}
-
-\theoremstyle{thmstyleone}%
-\newtheorem{theorem}{Theorem}%  meant for continuous numbers
-%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
-%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
-\newtheorem{proposition}[theorem]{Proposition}%
-%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
-\theoremstyle{thmstyletwo}%
-\newtheorem{example}{Example}%
-\newtheorem{remark}{Remark}%
-\theoremstyle{thmstylethree}%
-\newtheorem{definition}{Definition}
-
-\usepackage{subcaption}
-
-\usepackage{acronym}
-\acrodef{CRN}[CRN]{chemical reaction network}
-\acrodef{DSL}[DSL]{domain-specific language}
-\acrodef{GPU}[GPU]{graphical processing unit}
-\acrodef{GUI}[GUI]{graphical user interface}
-\acrodef{IDE}[IDE]{integrated development environment}
-\acrodef{JIT}[JIT]{just-in-time}
-\acrodef{ODE}[ODE]{ordinary differential equation}
-\acrodef{PyPI}[PyPI]{Python package index}
-\acrodef{RHS}[RHS]{right-hand side}
-\acrodef{SBML}[SBML]{Systems Biology Markup Language}
-\acrodef{SDE}[SDE]{stochastic differential equation}
-\acrodef{DDE}[DDE]{delay differential equation}
-\acrodef{TPU}[TPU]{tensor processing unit}
-
-\usepackage{tcolorbox}
-\newenvironment{CodeInput}{\begin{tcolorbox}[title=input,boxrule=0pt]}{\end{tcolorbox}}
-\newenvironment{CodeOutput}{\begin{tcolorbox}[title=output,boxrule=0pt]}{\end{tcolorbox}}
-
-\usepackage{fancyvrb}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.13,1.00}{#1}}
-\newcommand{\ClassTok}[1]{\textcolor[rgb]{0.27,0.56,0.65}{#1}}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.06,0.50}{#1}}
-\newcommand{\ValueTok}[1]{\textcolor[rgb]{0.13,0.57,0.41}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.47,0.37,0.15}{#1}}
-\newcommand{\IDETok}[1]{\textcolor[rgb]{0.58,0.58,0.58}{#1}}
-\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.06,0.50}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.00,0.50,0.00}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.70,0.27,0.27}{#1}}
-
-\begin{document}
-
-\journaltitle{Journal Title Here}
-\DOI{DOI HERE}
-\copyrightyear{2022}
-\pubyear{2019}
-\access{Advance Access Publication Date: Day Month Year}
-\appnotes{Paper}
-
-\firstpage{1}
-
-%\subtitle{Subject Section}
-
-\title[Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems]{Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems}
-
-\author[1,2]{Mauro Silberberg\ORCID{0000-0000-0000-0000}}
-\author[3]{Henning Hermjakob\ORCID{0000-0000-0000-0000}}
-\author[3]{Rahuman S. Malik-Sheriff\ORCID{0000-0000-0000-0000}}
-\author[1,2,$\ast$]{Hernán E. Grecco\ORCID{0000-0000-0000-0000}}
-
-\authormark{Author Name et al.}
-
-\address[1]{\orgname{Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Física}, \orgaddress{\state{Buenos Aires}, \country{Argentina}}}
-\address[2]{\orgname{CONICET - Universidad de Buenos Aires, Instituto de Física de Buenos Aires (IFIBA)}, \orgaddress{\state{Buenos Aires}, \country{Argentina}}}
-\address[3]{\orgdiv{European Bioinformatics Institute}, \orgname{European Molecular Biology Laboratory (EMBL-EBI)}, \orgaddress{\street{Wellcome Genome Campus}, \state{Cambridge}, \country{UK}}}
-
-\corresp[$\ast$]{Corresponding author. \href{email:hgrecco@df.uba.ar}{hgrecco@df.uba.ar}}
-
-\received{Date}{0}{Year}
-\revised{Date}{0}{Year}
-\accepted{Date}{0}{Year}
-
-%\editor{Associate Editor: Name}
-
-%\abstract{
-%\textbf{Motivation:} .\\
-%\textbf{Results:} .\\
-%\textbf{Availability:} .\\
-%\textbf{Contact:} \href{name@email.com}{name@email.com}\\
-%\textbf{Supplementary information:} Supplementary data are available at \textit{Journal Name}
-%online.}
-
-\abstract{
-  \Acfp{CRN} play a pivotal role in diverse fields
-  such as systems biology, biochemistry, chemical engineering, and epidemiology.
-  High-level definitions of \acp{CRN} enables to use various simulation approaches,
-  including deterministic and stochastic methods,
-  from the same model.
-  However,
-  existing Python tools for simulation of \ac{CRN} typically wrap external C/C++ libraries for
-  model definition, translation into equations and/or numerically solving them,
-  limiting their extensibility and integration with the broader Python ecosystem.
-  In response, we developed Poincaré and SimBio,
-  two novel Python packages for simulation of dynamical systems and \acp{CRN}.
-  Poincaré serves as a foundation for dynamical systems modelling,
-  while SimBio extends this functionality to \acp{CRN},
-  including support for the \ac{SBML}.
-  Poincaré and SimBio are developed as pure Python packages
-  enabling users to easily extend their simulation capabilities
-  by writing new or leveraging other Python packages.
-  Moreover, this does not compromise the performance,
-  as code can be \ac{JIT} compiled with Numba.
-  Our benchmark tests using curated models from the BioModels repository
-  demonstrate that these tools may provide a potentially
-  superior performance advantage compared to other existing tools.
-  Additionally, to ensure a user-friendly experience,
-  our packages use standard typed modern Python syntax that provides a seamless integration with \acp{IDE}.
-  Our Python-centric approach significantly enhances code analysis, error detection, and refactoring capabilities,
-  positioning Poincaré and SimBio as valuable tools for the modelling community.
-}
-\keywords{poincare, SimBio, Python, ODE}
-\acresetall % reset acronyms after abstract
-
-% \boxedtext{
-% \begin{itemize}
-% \item Key boxed text here.
-% \item Key boxed text here.
-% \item Key boxed text here.
-% \end{itemize}}
-
-\maketitle
-
-\section{Introduction}
-
-\Acp{CRN} are a fundamental concept of modelling in numerous fields
-including systems biology, biochemistry, chemical engineering and epidemiology.
-They are comprised of a set of chemical species or biological entities
-and a set of reactions that mediate transformations between them.
-These systems can be simulated through multiple approaches:
-deterministic \acp{ODE} to model macroscopic behavior,
-\acp{SDE} to model microscopic fluctuations,
-and jump processes (Gillespie-like simulations) to account for the discreteness of populations.
-Instead of directly writing the equations for each of these formulations,
-which is error-prone and difficult to reuse,
-these models can be \emph{defined} in a higher-level description
-that can be \emph{translated} into equations for the different types of simulations
-and, then, \emph{solved} numerically.
-
-Several tools already exist to \emph{define}, \emph{translate} and \emph{solve} \acp{CRN}.
-BioSimulators.org \citep{shaikhBioSimulatorsCentralRegistry2022},
-a registry of simulation tools,
-lists at least 15 softwares categorized under Python including
-COPASI \citep{hoopsCOPASICOmplexPAthway2006},
-Tellurium \citep{choiTelluriumExtensiblePythonbased2018} and
-PySB \citep{lopezProgrammingBiologicalModels2013}.
-COPASI is a standalone software with a \ac{GUI}
-that is widely used for its user-friendly interface and comprehensive features.
-It also includes Python bindings, BASICO \citep{bergmannBASICOSimplifiedPython2023b}, that allow advanced scripting.
-Tellurium is a Python-based modeling environment
-that uses a C++ library called libRoadRunner \citep{welshLibRoadRunnerHighPerformance2023} in the backend to translate and solve models.
-PySB is a Python library that created a \ac{DSL} using standard Python to define models,
-which are then translated to \acp{ODE} using a Perl library called BioNetGen \citep{harrisBioNetGenAdvancesRulebased2016}.
-
-One limitation of these tools is their extensibility from Python.
-As they wrap libraries in other languages for defining, translating and/or solving models,
-these steps cannot be altered or easily inspected from Python.
-While they enable model definition and running simulations via Python scripts,
-they cannot fully leverage Python's extensive package ecosystem.
-For example,
-COPASI and Tellurium do not allow the use of solvers defined in other Python packages,
-and adding new integrators requires working with C++.
-In particular, the step that translates into equations is not exposed by any of these tools.
-As such, it is not possible to apply custom optimizations to the equations
-or use automatic differentation packages such as JAX \citep{jax2018github} to compute the model's jacobian.
-
-Another challenge is the way models are defined.
-Many tools support the \ac{SBML} \citep{huckaSBMLL3V2} as an exchange format,
-a \emph{de facto} standard for \acp{CRN} that defines species, parameters and reactions between species.
-As writing \ac{SBML} directly is impractical,
-Tellurium uses a \ac{DSL} called Antimony \citep{smithAntimonyModularModel2009} for defining models.
-\Acp{DSL} allows to reuse the same code in different programming environments,
-but are not recognized by default in \acp{IDE}
-and, therefore, they cannot provide syntax highlighting, code completion, refactoring, and static analysis.
-For Antimony, an extension providing these capabilities was developed for Visual Studio Code,
-but its maintenance could be a demanding task for the systems biology community.
-In the case of PySB,
-using Python's dynamic nature,
-it developers designed a \ac{DSL} within Python.
-To save keystrokes,
-it uses the global scope to create species and parameters,
-without explicitly assigning them to Python variables or to the model,
-but this approach is not fully compatible with \acp{IDE},
-affecting the development experience.
-
-To overcome these limitations,
-we developed poincaré and SimBio,
-open-source Python packages for defining, translating and solving systems.
-Poincaré allows one to define differential equation systems
-using variables, parameters and constants,
-and assiging rate equations to variables.
-For defining \acp{CRN},
-SimBio builds on top of poincaré providing species and reactions that keep track of stoichiometries.
-Both are focused on providing an ergonomic experience to end-users
-by integrating well with \acp{IDE} and static analysis tools
-through the use of standard modern Python syntax.
-Moreover, since they are coded in pure Python,
-each step from model definition, translation to equations or solving
-can be extended or debugged from Python.
-Being the first-ever pure Python packages for systems modelling,
-they offer extensive extensibility,
-from simple tasks like reusing integrators defined in other packages,
-to complex ones like altering the compilation process to leverage some structure in the equations.
-For example, using a for-loop in the compiled equations could improve the runtime performance
-if there is some repetitive structure in the system,
-as happens in spatial modelling.
-The models built using these packages can be introspected to create other representations,
-such as graphs connecting species and/or reactions, or tables with parameters or equations.
-Furthermore, they have a modular architecture with a clear separation of concerns,
-making it easier to maintain or to contribute new code,
-which is beneficial for developers and maintainers.
-We showcased the reliability of these tools by benchmarking them against the simulation results from other tools.
-We also highlighted the substantial performance improvements our tools offer,
-as this is crucial for construction and simulation of models of whole cells and organisms,
-which necessitate the simulation of significantly large-scale models.
-
-\section{Results}\label{results}
-
-Modular code architecture makes code reusable, extensible, and easier to maintain.
-Therefore, we split the code into three Python packages:
-symbolite, to create symbolic expressions;
-poincaré, to define dynamical systems;
-and simbio, to define \acp{CRN} and interface with systems biology standards such as \ac{SBML}.
-These are pure Python packages with standard dependencies from the PyData scientific stack
-such as NumPy \citep{harrisArrayProgrammingNumPy2020} and pandas \citep{mckinneyDataStructuresStatistical2010}.
-They are published in the \acf{PyPI},
-where links to the source code and documentation hosted in GitHub can be found,
-and can be easily installed with \texttt{pip\ install\ simbio},
-which installs symbolite and poincare as dependencies.
-
-Symbolite is a lightweight symbolics package to create algebraic mathematical expressions.
-Unlike SymPy \citep{meurerSymPySymbolicComputing2017},
-a widely-used Python library for symbolic mathematics,
-it only provides the building of an expression tree
-which can be inspected and compiled to various backends.
-Symbolite is designed to facilitate the integration of new backends.
-Currently, we have implementations for
-NumPy \citep{harrisArrayProgrammingNumPy2020};
-Numba \citep{lamNumbaLLVMbasedPython2015},
-a \ac{JIT} compiler to LLVM;
-SymPy \citep{meurerSymPySymbolicComputing2017};
-and JAX \citep{jax2018github}, a
-library that support automatic differentiation and compilation to \acp{GPU} and \acp{TPU}.
-
-\subsection{Versatile modelling and simulation of dynamical systems with Poincaré}
-\label{versatile-modelling-and-simulation-of-dynamical-systems-with-poincaruxe9}
-
-Poincare is a package to define and simulate dynamical systems.
-Based on Python inmutable dataclasses,
-it provides a \texttt{System} class,
-where one can define
-\texttt{Constant}s; which can be numbers or refer to other constants;
-\texttt{Parameter}s, which can be numbers or time-dependent expressions;
-\texttt{Variable}s, which represent the state of the system and must be provided with an initial condition;
-and create equations linking a variable's derivative with an expression (Figure~\ref{fig-first-order}).
-It also allows to
-create an \texttt{Independent} variable to define non-autonomous systems,
-and define higher-order systems by assigning an initial condition to a \texttt{Derivative} (Figure~\ref{fig-second-order}).
-
-Within the constraints imposed by Python's current typing and static analyzers,
-we define models utilizing Python (data)classes
-such that we can benefit from \acp{IDE}' autocomplete and refactoring capabilities.
-This design offer several advantages:
-
-\begin{enumerate}
-\item
-  The variable name to which a component is assigned can be automatically saved in the component for introspection
-  (i.e., \texttt{Oscillator.x.name\ ==\ "x"}),
-\item
-  It provides a namespace that allows to easily define multiple independent models in the same script,
-\item
-  It allows \acp{IDE} to provide autocomplete and refactoring capabilities
-  (\texttt{Oscillator.\textless{}TAB\textgreater{}} shows \texttt{x}, \texttt{v} and \texttt{eq}),
-\item
-  It allows creation of instances that can be composed into a bigger model (Figure~\ref{fig-composition}).
-\end{enumerate}
-
-For this last point,
-\acp{IDE} that support \texttt{dataclass\_transform} \citep{debontePEP681Data2021}
-can provide a tooltip with the expected signature (Figure~\ref{fig-ide}).
-This requires the use of type annotations,
-which play a more significant role in static type checking
-as they can help to identify errors before running the code.
-For instance, to parameterize the initial conditions of variables
-we have to use a \texttt{Constant}.
-If we try to use a \texttt{Parameter},
-which could be a time-dependent expression,
-it is flagged as a type error (Figure~\ref{fig-ide}).
-
-To simulate a system,
-we created a \texttt{Simulator} instance (Figure~\ref{fig-sim})
-that translates the model into the \ac{RHS} equations
-and interfaces with solvers
-wrapping the output in a \texttt{pandas.DataFrame},
-which can be easily plotted with the standard \texttt{plot} method.
-Currently,
-it only supports translating into first-order \acp{ODE},
-but it would be possible to add support for \acp{SDE} or \acp{DDE}.
-By default,
-it uses \texttt{numpy} as a backend,
-and uses the LSODA solver from \texttt{scipy}.
-This can be easily switched to other solvers or backends, such as \texttt{numba}.
-
-\subsection{Extensible definition of reaction networks usingSimBio}
-\label{extensible-definition-of-reaction-networks-using-simbio}
-
-For the \acp{CRN},
-our focus is on first-order differential equations that describe the rate of change of species.
-SimBio simplifies the definition of these network models by introducing \texttt{Species},
-and \texttt{RateLaw}, a construct that converts reactant species into product species taking into account the stoichiometry (Figure~\ref{fig-simbio}).
-Additionally, SimBio features \texttt{MassAction},
-a subclass of \texttt{RateLaw},
-that intuitively incorporates reactants and their stoichiometry into the rate law (Figure~\ref{fig-simbio}).
-
-Several commonly used reactions are predefined as \texttt{MassAction} subclasses,
-such as \texttt{MichaelisMenten} (\(S + E \leftrightarrow ES \rightarrow P + E\))
-and its approximate form without the intermediate species \(ES\),
-and it is also simple to implement used-defined ones
-as subclasses of \texttt{RateLaw} or \texttt{MassAction}.
-Additionally, SimBio supports importing models from \ac{SBML},
-and downloading them directly from BioModels \citep{malik-sheriffBioModels15Years2020} (Figure~\ref{fig-simbio-io}).
-Work is in progress to support exporting to \ac{SBML} and add more \ac{SBML} features.
-Currently \ac{SBML} unsupported features include
-algebraic rules, constraints, events, reactions with the fast attribute or with math stoichiometry,
-units and compartments with size different from 1.
-
-\subsection{Reproducibility and performance}
-\label{reproducibility-and-performance}
-
-To evaluate SimBio's reproducibility,
-we used the \ac{SBML} test suite \citep{SBMLTestSuite},
-which provides a set of \ac{SBML} models
-and the expected result of a simulation.
-Excluding models that use \ac{SBML} features not yet supported by SimBio,
-every simulation returned correct results within the solver tolerances.
-
-To evaluate SimBio's performance,
-we selected \ac{SBML} models from the curated section of BioModels \citep{malik-sheriffBioModels15Years2020}.
-Among the first 250 models,
-we considered the 117 that used supported \ac{SBML} features.
-We ran simulations
-on a MacBook Air with M2 CPU
-using Python 3.11.8,
-COPASI v4.42.284
-(with BasiCO v0.58) with the LSODA solver,
-RoadRunner v2.5.0 (with Tellurium v2.2.10) with the comparable CVODE solver, and
-SimBio v0.3.2 with the LSODA solver.
-For SimBio, we considered three variants:
-NumPy (v1.26.4) backend and scipy (v1.12.0)'s LSODA solver,
-Numba (v0.59.0) backend and scipy's LSODA solver,
-Numba backend and numbalsoda's (v0.3.5) LSODA solver.
-In all cases, we used a relative tolerance of \(10^{-6}\) and absolute tolerances of \(10^{-9}\).
-We measured two simulation stages:
-an initial cold run that includes the reading of the \ac{SBML} model
-and subsequent warm runs.
-
-For COPASI and Tellurium,
-we noted that its runtime depended on the number of intermediate evaluation points returned to the user (Figure~\ref{fig-runtime}, left).
-We speculate that this is due to memory allocation and data transfer in the Python bindings, 
-as the number of total function evaluations (i.e. including those performed by the integrator steper)
-is around 3000 and therefore much larger than most part of the x-axis.
-For SimBio,
-its NumPy backend can be orders of magnitude slower than both COPASI and RoadRunner  (Figure~\ref{fig-runtime}, right).
-Nevertheless, switching to the numba backend,
-which \ac{JIT} compiles the \ac{RHS} equations,
-puts it on par with them.
-Another speed-up in the runtime can be had by switching the LSODA \texttt{scipy} solver
-for a more efficient \texttt{numbalsoda} implementation that avoids calling into the Python interpreter
-between each of the integration steps.
-A user might have to consider the trade-off between compilation and run times,
-as the compilation of the \ac{RHS} code might take longer than the runtime itself,
-and not be worth it for running the model only once.
-For the models considered,
-amortizing the compilation time required from 2 upto 200 runs.
-
-\section{Discussion}\label{discussion}
-
-In this article,
-we introduced a suite of Python packages we developed
-for defining and simulating dynamical systems and \acp{CRN}.
-These packages are deeply integrated with \acp{IDE},
-enabling code analysis tools to identify errors prior to execution
-and assist in refactoring and code completion.
-We adopted standard modern Python syntax to ensure seamless IDE integration,
-supported by the extensive Python community.
-
-Our approach differs from previous tools in that both the model
-definition and its compilation into an \ac{ODE} function are entirely Python-based.
-This approach simplifies the development of various simulation methods,
-including performance enhancements that exploit specific model structures.
-Importantly, being Python-based does not compromise performance compared to C/C++ tools,
-as the \ac{RHS} functions can be \ac{JIT} compiled using Numba.
-
-The inclusion of \ac{SBML} support facilitates the effortless reuse of models created by the systems biology community,
-along with the vast collection of public models hosted in the BioModels repository.
-The modular architecture of these packages facilitates their reuse, enhancement, and extension by the wider Python community.
-Therefore, it should be also easy to integrate with existing infrastructure such a BioSimulators.org \citep{shaikhBioSimulatorsCentralRegistry2022},
-or combine with other packages like SimService \citep{sego_simservice_2024} to build more complex simulations.
-For instance, an individual from outside the systems biology field could contribute a stochastic integrator to poincaré,
-which would then be available in SimBio.
-This clear separation of concerns also makes the packages more comprehensible,
-lowering the barrier for contributing improvements or new features.
-Such an architecture ensures their maintainability and ongoing development well into the future.
-
-\section{Resources}\label{resources}
-
-Source code repositories, example and documentation for these packages can be found at
-\url{https://github.com/maurosilber/poincare} and
-\url{https://github.com/hgrecco/simbio}.
-
-%%%%%%%%%%%%%%
-
-\section{Competing interests}
-No competing interest is declared.
-
-\section{Author contributions statement}
-
-Must include all authors, identified by initials, for example:
-S.R. and D.A. conceived the experiment(s),  S.R. conducted the experiment(s), S.R. and D.A. analysed the results.  S.R. and D.A. wrote and reviewed the manuscript.
-
-\section{Acknowledgments}
-The authors thank the anonymous reviewers for their valuable suggestions. This work is supported in part by funds from the National Science Foundation (NSF: \# 1636933 and \# 1920920).
-
-% bibliography
-\bibliographystyle{abbrvnat}
-\bibliography{bibliography}
-
-%% sample for biography with author's image
-% \begin{biography}{{\color{black!20}\rule{77pt}{77pt}}}{\author{Author Name.} This is sample author biography text.
-The values provided in the optional argument are meant for sample purposes.
-There is no need to include the width and height of an image in the optional argument for live articles.
-This is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text.}
-% \end{biography}
-
-%% sample for biography without author's image
-% \begin{biography}{}{\author{Author Name.} This is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text.}
-% \end{biography}
-
 % FIGURE 1
@@
   \begin{figure*}[t]
-
     \begin{minipage}[t]{\linewidth}
-
-  {\centering
-
       \begin{CodeInput}
@@
       \end{CodeInput}
-
-  }
-
     \end{minipage}%
@@
     \begin{minipage}[t]{0.45\linewidth}
-
-  {\centering
-
       \begin{CodeInput}
@@
       \end{CodeInput}
-
+      %
       \begin{CodeOutput}
@@
       \end{CodeOutput}
-
-  }
-
     \end{minipage}%
@@
     as index. It is inspected and plotted with the \texttt{pandas} methods.}
-
   \end{figure*}
@@
   \end{figure*}
-
-\end{document}

### 2024-06-15 final/thin-figures.tex
Reorder figure 2
Two column figure 3
Thin Figure 5

@@
 % FIGURE 2
-  \begin{figure*}[t]
-
-    \begin{minipage}[b]{0.50\linewidth}
-    {
-        \centering 
-        \includegraphics[width=\textwidth]{ide1.pdf}
-    }
-    \end{minipage}%
-    %
-    \begin{minipage}[b]{0.50\linewidth}
-    {
-        \centering
-        \includegraphics[width=\textwidth]{ide2.pdf}
-    }
-    \end{minipage}%
-
+\begin{figure}[t]
+  \includegraphics[width=\columnwidth]{ide1.pdf}
+  \includegraphics[width=\columnwidth]{ide2.pdf}
   \caption{
@@
   }  
-
-    \end{figure*}
+\end{figure}
 
 % FIGURE 3
-  \begin{figure*}[t]
-    \begin{minipage}[t]{\linewidth}
+\begin{figure}[t]
+  \begin{minipage}[t]{\columnwidth}
     \begin{CodeInput}
@@
     \VariableTok{df} \OperatorTok{=} \VariableTok{sim}.\FunctionTok{solve}\KeywordTok{(}\VariableTok{save\_at}\OperatorTok{=}\ClassTok{np}.\FunctionTok{linspace}\KeywordTok{(}\ValueTok{0}, \ValueTok{5}, \ValueTok{101}\KeywordTok{))}
-      \end{Highlighting}
-      \end{CodeInput}
-    \end{minipage}%
-    \newline
-    \begin{minipage}[t]{0.45\linewidth}
-      \begin{CodeInput}
-      \begin{Highlighting}[]
     \VariableTok{df}.\FunctionTok{head}\KeywordTok{()}
@@
     \end{CodeInput}
-      %
     \begin{CodeOutput}
@@
   \hfill
-    \begin{minipage}[t]{0.45\linewidth}
+  \begin{minipage}[t]{\columnwidth}
     \begin{CodeInput}
     \begin{Highlighting}[]
-      \VariableTok{df}.\FunctionTok{plot}\KeywordTok{()}\OperatorTok{;}
+    \VariableTok{df}.\FunctionTok{plot}\KeywordTok{()}
     \end{Highlighting}
@@
   as index. It is inspected and plotted with the \texttt{pandas} methods.}
-  \end{figure*}
+\end{figure}
 
@@
 % FIGURE 5
-  \begin{figure*}[t]
+\begin{figure}[t]
   \centering
@@
 
-    \VariableTok{model }\OperatorTok{=}\ClassTok{ sbml}.\FunctionTok{load}\KeywordTok{(}\StringTok{"repressilator.sbml"}\KeywordTok{)}  \CommentTok{\# from existing SBML file}
-    \VariableTok{model }\OperatorTok{=}\ClassTok{ biomodels}.\FunctionTok{load}\KeywordTok{(}\StringTok{"BIOMD12"}\KeywordTok{)}  \CommentTok{\# from BioModels}
+  \CommentTok{\# from existing SBML file}
+  \VariableTok{model }\OperatorTok{=}\ClassTok{ sbml}.\FunctionTok{load}\KeywordTok{(}\StringTok{"repressilator.sbml"}\KeywordTok{)}
+  \CommentTok{\# from BioModels}
+  \VariableTok{model }\OperatorTok{=}\ClassTok{ biomodels}.\FunctionTok{load}\KeywordTok{(}\StringTok{"BIOMD12"}\KeywordTok{)}
   \VariableTok{model}
@@
           variables  &  6 & PX, PY, PZ, X, Y, Z \\
-            parameters & 17 & cell, beta, alpha0, alpha, eff, n, KM, tau\_mRNA, tau\_prot, ... \\
-            equations  & 12 & Reaction1, Reaction2, Reaction3, Reaction4, Reaction5, ... \\
+          parameters & 17 & cell, beta, alpha0, alpha, eff, n, ... \\
+          equations  & 12 & Reaction1, Reaction2, Reaction3, ... \\
           \botrule
@@
   \label{fig-simbio-io}
-  \end{figure*}
+\end{figure}
 

### 2024-06-15 final/oup-authoring-template.tex
Remove centering
Added ORCID
Change to thin figures

@@
 
-\author[1,2]{Mauro Silberberg\ORCID{0000-0000-0000-0000}}
-\author[3]{Henning Hermjakob\ORCID{0000-0000-0000-0000}}
-\author[3]{Rahuman S. Malik-Sheriff\ORCID{0000-0000-0000-0000}}
-\author[1,2,$\ast$]{Hernán E. Grecco\ORCID{0000-0000-0000-0000}}
+\author[1,2]{Mauro Silberberg\ORCID{0000-0002-2402-1100}}
+\author[3]{Henning Hermjakob\ORCID{0000-0001-8479-0262}}
+\author[3]{Rahuman S. Malik-Sheriff\ORCID{0000-0003-0705-9809}}
+\author[1,2,$\ast$]{Hernán E. Grecco\ORCID{0000-0002-1165-4320}}
 
@@
 
+% FIGURE 1
+\begin{figure}[t]
+  \begin{minipage}[t]{\columnwidth}    
+  \begin{CodeInput}
+\begin{Highlighting}[]
+\KeywordTok{from} \ClassTok{poincare} \KeywordTok{import} (\ClassTok{Derivative}, \ClassTok{System},
+                    \ClassTok{Variable}, \FunctionTok{initial})
+\end{Highlighting}
+  \end{CodeInput}
+  \end{minipage}%
+  \newline
+  \begin{minipage}[t]{\columnwidth}
+  \begin{minipage}[c]{\columnwidth}
+  \begin{CodeInput}
+\begin{Highlighting}[]
+\KeywordTok{class} \ClassTok{Decay}\KeywordTok{(}\ClassTok{System}\KeywordTok{)}\OperatorTok{:}
+  \VariableTok{x}\OperatorTok{:} \ClassTok{Variable} \OperatorTok{=} \FunctionTok{initial}\KeywordTok{(}\VariableTok{default}\OperatorTok{=}\ValueTok{1}\KeywordTok{)}
+  \VariableTok{eq} \OperatorTok{=} \VariableTok{x}.\FunctionTok{derive}\KeywordTok{()} \OperatorTok{\textless{}\textless{}} \OperatorTok{-}\VariableTok{x}
+\end{Highlighting}
+  \end{CodeInput}
+  \end{minipage}%
+  %
+  \newline
+  \begin{minipage}[c]{0.5\columnwidth}
+  \[ \frac{dx}{dt} = -x \]
+  \end{minipage}%
+  \begin{minipage}[c]{0.5\columnwidth}
+  \[ x(0) = 1 \]
+  \end{minipage}%
+
+  \subcaption{\label{fig-first-order}First-order system of an exponential decay. The variable
+  \texttt{x} stores the initial condition for \(x\), and the variable
+  \texttt{eq} stores the rate equation for \(x\).}
+  \end{minipage}%
+  \newline
+  \begin{minipage}[t]{\columnwidth}
+
+  \begin{minipage}[c]{\columnwidth}    
+  \begin{CodeInput}
+\begin{Highlighting}[]
+\KeywordTok{class} \ClassTok{Oscillator}\KeywordTok{(}\ClassTok{System}\KeywordTok{)}\OperatorTok{:}
+  \VariableTok{x}: \ClassTok{Variable} \OperatorTok{=} \FunctionTok{initial}\KeywordTok{(\VariableTok{default}\OperatorTok{=}\ValueTok{1})}
+  \VariableTok{v}: \ClassTok{Derivative} \OperatorTok{=} \VariableTok{x}.\FunctionTok{derive}\KeywordTok{(\VariableTok{initial}\OperatorTok{=}\ValueTok{0})}
+  \VariableTok{eq} \OperatorTok{=} \VariableTok{v}.\FunctionTok{derive}\KeywordTok{()} \OperatorTok{\textless{}\textless{}} \OperatorTok{-}\VariableTok{x}
+\end{Highlighting}
+  \end{CodeInput}
+
+  \end{minipage}%
+  %
+  \newline
+  \begin{minipage}[c]{0.5\columnwidth}
+  \[ \frac{d^2x}{dt^2} = -x \]
+  \end{minipage}%
+  \begin{minipage}[c]{0.5\columnwidth}
+  \[
+  \begin{cases}
+    \begin{aligned}
+      x(0) &= 1 \\
+      \frac{dx}{dt}(0) &= 0
+    \end{aligned}
+  \end{cases}
+  \]
+  \end{minipage}%
+
+  \subcaption{\label{fig-second-order}Second-order system of an harmonic oscillator, where the
+  variable \texttt{v} stores the derivative \(\frac{dx}{dt}\) and the rate
+  equation is specified for the derivative \texttt{v}.}
+  \end{minipage}%
+  \newline
+  \begin{minipage}[t]{\columnwidth}
+  \begin{minipage}[c]{\columnwidth}
+  \begin{CodeInput}
+\begin{Highlighting}[]
+\KeywordTok{class} \ClassTok{BigModel}\KeywordTok{(}\ClassTok{System}\KeywordTok{)}:
+  \VariableTok{x}: \ClassTok{Variable} \OperatorTok{=} \FunctionTok{initial}\KeywordTok{(\VariableTok{default}\OperatorTok{=}\ValueTok{1})}
+  \VariableTok{linked} \OperatorTok{=} \ClassTok{Decay}\KeywordTok{(\VariableTok{x}\OperatorTok{=}\VariableTok{x})}
+  \VariableTok{independent} \OperatorTok{=} \ClassTok{Decay}\KeywordTok{(\VariableTok{x}\OperatorTok{=}\ValueTok{2})}
+\end{Highlighting}
+  \end{CodeInput}
+  \end{minipage}%
+  %
+  \newline
+  \begin{minipage}[c]{0.5\columnwidth}
+  \[
+  \begin{cases}
+    \begin{aligned}
+      \frac{dx}{dt} &= -x \\
+      \frac{dx_{indep}}{dt} &= -x_{indep}
+    \end{aligned}
+  \end{cases}
+  \]
+\end{minipage}%
+\begin{minipage}[c]{0.5\columnwidth}
+  \[
+  \begin{cases}
+    \begin{aligned}
+      x(0) &= 1 \\
+      x_{indep}(0) &= 2
+    \end{aligned}
+  \end{cases}
+  \]
+  \end{minipage}%
+
+  \subcaption{\label{fig-composition}First-order system of two exponential decays by composition of
+  the \texttt{Decay} system. The subsystem \texttt{linked} has a reference
+  to the outer variable \texttt{x}, while the subsystem
+  \texttt{independent} defines a new variable with initial condition $2$,
+  which on the corresponding mathematical expression was named \(x_{ind}\).}
+  \end{minipage}%
+
+  \caption{\label{fig-poincare}Code and corresponding mathematical
+  expressions for different systems.}
+
+\end{figure}
+
+% FIGURE 2
+\begin{figure}[t]
+  \includegraphics[width=\columnwidth]{ide1.pdf}
+  \includegraphics[width=\columnwidth]{ide2.pdf}
+  \caption{
+    \label{fig-ide}
+    Screenshots of Visual Studio Code showing
+    tooltips (green arrows) and
+    highlighted type errors (red arrows).
+    On the left,
+    we show that \texttt{a},
+    a \texttt{Constant} assigned with \texttt{assign(...,\ constant=True)},
+    can be used for \texttt{Variable b}'s initial condition.
+    Instead,
+    it is flagged as a type error (red underlining)
+    when using \texttt{c}, a \texttt{Parameter},
+    for \texttt{Variable d}'s initial condition,
+    The IDE automatically recognizes \texttt{e} as an \texttt{Equation},
+    and provides autocompletion of the \texttt{Model}'s components.
+    A tooltip is shown when composing models (green arrow, right),
+    which show the expected variables and their default values.
+    The IDE highlights wrong names (\texttt{z} is not a name in \texttt{Model})
+    and mismatched types (\texttt{x} is \texttt{Variable} and \texttt{a} must be a number or a \texttt{Constant})
+  }  
+\end{figure}
+
+% FIGURE 3
+\begin{figure}[t]
+  \begin{minipage}[t]{\columnwidth}
+    \begin{CodeInput}
+\begin{Highlighting}[]
+\KeywordTok{import} \ClassTok{numpy} \KeywordTok{as} \ClassTok{np}
+\KeywordTok{from} \ClassTok{poincare} \KeywordTok{import} \ClassTok{Simulator}
+
+\VariableTok{sim} \OperatorTok{=} \ClassTok{Simulator}\KeywordTok{(\ClassTok{Oscillator})}
+\VariableTok{df} \OperatorTok{=} \VariableTok{sim}.\FunctionTok{solve}\KeywordTok{(}\VariableTok{save\_at}\OperatorTok{=}\ClassTok{np}.\FunctionTok{linspace}\KeywordTok{(}\ValueTok{0}, \ValueTok{5}, \ValueTok{101}\KeywordTok{))}
+\VariableTok{df}.\FunctionTok{head}\KeywordTok{()}
+\end{Highlighting}
+    \end{CodeInput}
+    \begin{CodeOutput}
+        \begin{tabular}[]{@{}lll@{}}
+            \toprule
+            & x & v \\
+            time & & \\
+            \midrule
+            0.00 & 1.000000 & 0.000000 \\
+            0.05 & 0.998750 & -0.049979 \\
+            0.10 & 0.995005 & -0.099835 \\
+            0.15 & 0.988772 & -0.149440 \\
+            0.20 & 0.980069 & -0.198672 \\
+            \botrule
+        \end{tabular}
+    \end{CodeOutput}
+  \end{minipage}%
+  \hfill
+  \begin{minipage}[t]{\columnwidth}
+    \begin{CodeInput}
+\begin{Highlighting}[]
+\VariableTok{df}.\FunctionTok{plot}\KeywordTok{()}
+\end{Highlighting}
+    \end{CodeInput}
+    %
+    \begin{CodeOutput}
+      \includegraphics[width=\textwidth]{cell-output}
+    \end{CodeOutput}
+  \end{minipage}%
+
+  \caption{\label{fig-sim}Simulation of the \texttt{Oscillator} system
+  from Figure~\ref{fig-second-order}. The output is a
+  \texttt{pandas.DataFrame} with a column for each variable and the time
+  as index. It is inspected and plotted with the \texttt{pandas} methods.}
+\end{figure}
+
 Poincare is a package to define and simulate dynamical systems.
@@
 
-\subsection{Extensible definition of reaction networks usingSimBio}
+\subsection{Extensible definition of reaction networks using SimBio}
 \label{extensible-definition-of-reaction-networks-using-simbio}
+
+% FIGURE 4
+\begin{figure}[t]
+
+  \begin{minipage}[t]{\columnwidth}
+    \begin{CodeInput}
+\begin{Highlighting}[]
+\KeywordTok{from} \ClassTok{simbio} \KeywordTok{import} (\ClassTok{Compartment}, \ClassTok{MassAction},
+                    \ClassTok{Species}, \ClassTok{RateLaw}, \FunctionTok{initial})
+\end{Highlighting}
+    \end{CodeInput}
+  \end{minipage}%
+  \newline
+  \begin{minipage}[t]{\columnwidth}
+    \begin{minipage}[c]{0.7\columnwidth}
+      \begin{CodeInput}
+\begin{Highlighting}[]
+\KeywordTok{class}\ClassTok{ Model}\KeywordTok{(}\ClassTok{Compartment}\KeywordTok{)}:
+  \CommentTok{"""2A {-}\textgreater{} B"""}
+\VariableTok{    A}: \ClassTok{Species }\OperatorTok{=}\FunctionTok{ initial}\KeywordTok{(}\VariableTok{default}\OperatorTok{=}\ValueTok{1}\KeywordTok{)}
+\VariableTok{    B}: \ClassTok{Species }\OperatorTok{=}\FunctionTok{ initial}\KeywordTok{(}\VariableTok{default}\OperatorTok{=}\ValueTok{0}\KeywordTok{)}
+\VariableTok{    r }\OperatorTok{=}\ClassTok{ RateLaw}\KeywordTok{(}
+\VariableTok{        reactants}\OperatorTok{=}\KeywordTok{[}\ValueTok{2} \OperatorTok{*}\VariableTok{ A}\KeywordTok{]},
+\VariableTok{        products}\OperatorTok{=}\NormalTok{\KeywordTok{[}B\KeywordTok{]},}
+\VariableTok{        rate}\OperatorTok{=}\ValueTok{1}\NormalTok{,}
+\KeywordTok{    )}
+\end{Highlighting}
+      \end{CodeInput}
+    \end{minipage}%
+    %
+    \begin{minipage}[c]{0.3\columnwidth}
+      \begin{align*}
+        & \begin{cases}
+          A(0) = 1 \\
+          B(0) = 0
+          \end{cases}
+        \\
+        & \begin{cases}
+          \frac{dA}{dt} = -2 \\
+          \frac{dB}{dt} = +1
+          \end{cases}
+      \end{align*}
+  \end{minipage}%
+  \end{minipage}%
+  \newline
+  \begin{minipage}[t]{\columnwidth}
+      \centering 
+      \begin{minipage}[c]{0.7\columnwidth}
+\centering 
+\begin{CodeInput}
+\begin{Highlighting}[]
+\KeywordTok{class}\ClassTok{ Model}\KeywordTok{(}\ClassTok{Compartment}\KeywordTok{)}:
+  \CommentTok{"""2A {-}\textgreater{} B"""}
+  \VariableTok{A}: \ClassTok{Species} \OperatorTok{=} \FunctionTok{initial}\KeywordTok{(}\VariableTok{default}\OperatorTok{=}\ValueTok{1}\KeywordTok{)}
+  \VariableTok{B}: \ClassTok{Species} \OperatorTok{=} \FunctionTok{initial}\KeywordTok{(}\VariableTok{default}\OperatorTok{=}\ValueTok{0}\KeywordTok{)}
+  \VariableTok{r} \OperatorTok{=} \ClassTok{MassAction}\KeywordTok{(}
+    \VariableTok{reactants}\OperatorTok{=}\KeywordTok{[}\ValueTok{2} \OperatorTok{*} \VariableTok{A}\KeywordTok{]},
+    \VariableTok{products}\OperatorTok{=}\NormalTok{\KeywordTok{[}B\KeywordTok{]},}
+    \VariableTok{rate}\OperatorTok{=}\ValueTok{1}\NormalTok{,}
+  \KeywordTok{)}
+          \end{Highlighting}
+          \end{CodeInput}
+      \end{minipage}%
+      %
+      \begin{minipage}[c]{0.3\columnwidth}
+        \begin{align*}
+          & \begin{cases}
+              A(0) = 1 \\
+              B(0) = 0
+          \end{cases} \\
+          & \begin{cases}
+              \frac{dA}{dt} = -2 A^2 \\
+              \frac{dB}{dt} = +1 A^2
+            \end{cases}
+        \end{align*}
+      \end{minipage}%
+  \end{minipage}%
+
+  \caption{
+    A reaction system for species \(A\) and \(B\)
+    with initial conditions \(1\) and \(0\), respectively. A single reaction
+    transforming \(2A\) into \(B\) is saved in variable \texttt{r}. The rate
+    \(1\) is specified directly for \texttt{RateLaw}, and is proportional to
+    the reactants for \texttt{MassAction}.
+  }
+  \label{fig-simbio}
+\end{figure}
+
+% FIGURE 5
+\begin{figure}[t]
+  \centering
+
+  \begin{CodeInput}
+\begin{Highlighting}[]
+\KeywordTok{from} \ClassTok{simbio.io} \KeywordTok{import} \ClassTok{biomodels}, \ClassTok{sbml}
+
+\CommentTok{\# from existing SBML file}
+\VariableTok{model} \OperatorTok{=} \ClassTok{sbml}.\FunctionTok{load}\KeywordTok{(}\StringTok{"repressilator.sbml"}\KeywordTok{)}
+\CommentTok{\# from BioModels}
+\VariableTok{model} \OperatorTok{=} \ClassTok{biomodels}.\FunctionTok{load}\KeywordTok{(}\StringTok{"BIOMD12"}\KeywordTok{)}
+\VariableTok{model}
+\end{Highlighting}
+  \end{CodeInput}
+
+  \begin{CodeOutput}
+      \begin{tabular}{@{}lll@{}}
+          \multicolumn{3}{l}{Elowitz2000 - Repressilator} \\
+          \toprule
+          type & total & names \\
+          \midrule
+          variables  &  6 & PX, PY, PZ, X, Y, Z \\
+          parameters & 17 & cell, beta, alpha0, alpha, eff, n, ... \\
+          equations  & 12 & Reaction1, Reaction2, Reaction3, ... \\
+          \botrule
+      \end{tabular}
+  \end{CodeOutput}
+
+  \caption{Creation of a model from a local SBML file or one uploaded to BioModels.}
+  \label{fig-simbio-io}
+\end{figure}
 
@@
 \label{reproducibility-and-performance}
+
+% FIGURE 6
+\begin{figure*}[t]
+  \centering
+  \includegraphics[width=\textwidth]{performance}
+  \caption{
+    Performance of different softwares to solve models from the curated section of BioModels.
+    (left) Run time for the model BIOMD3 as a function of the number of output points.
+    (right) Run time for different models for 300 output points,
+    using the geometric mean of the different softwares to order them.
+    Each point corresponds to the median of 20 runs,
+    with a neglibible errorbar given by the interquartile range.
+  }
+  \label{fig-runtime}
+\end{figure*}
 
@@
 
-%% sample for biography with author's image
-% \begin{biography}{{\color{black!20}\rule{77pt}{77pt}}}{\author{Author Name.} This is sample author biography text.
-The values provided in the optional argument are meant for sample purposes.
-There is no need to include the width and height of an image in the optional argument for live articles.
-This is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text.}
-% \end{biography}
-
-%% sample for biography without author's image
-% \begin{biography}{}{\author{Author Name.} This is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text this is sample author biography text.}
-% \end{biography}
-
-% FIGURE 1
-\begin{figure*}[t]
-  \begin{minipage}[t]{\linewidth}
-
-  {\centering
-
-  \begin{CodeInput}
-  \begin{Highlighting}[]
-  \KeywordTok{from} \ClassTok{poincare} \KeywordTok{import} \ClassTok{Derivative}, \ClassTok{System}, \ClassTok{Variable}, \FunctionTok{initial}
-  \end{Highlighting}
-  \end{CodeInput}
-
-  }
-
-  \end{minipage}%
-  \newline
-  \begin{minipage}[t]{\linewidth}
-
-  {\centering
-
-  \begin{minipage}[c]{0.6\linewidth}
-
-  {\centering
-
-  \begin{CodeInput}
-  \begin{Highlighting}[]
-  \KeywordTok{class} \ClassTok{Decay}\KeywordTok{(}\ClassTok{System}\KeywordTok{)}\OperatorTok{:}
-      \VariableTok{x}\OperatorTok{:} \ClassTok{Variable} \OperatorTok{=} \FunctionTok{initial}\KeywordTok{(\VariableTok{default}\OperatorTok{=}\ValueTok{1})}
-      \VariableTok{eq} \OperatorTok{=} \VariableTok{x}.\FunctionTok{derive}\KeywordTok{()} \OperatorTok{\textless{}\textless{}} \OperatorTok{-}\VariableTok{x}
-  \end{Highlighting}
-  \end{CodeInput}
-
-  }
-
-  \end{minipage}%
-  %
-  \begin{minipage}[c]{0.2\linewidth}
-
-  {\centering
-
-  \[
-  \frac{dx}{dt} = -x
-  \]
-
-  }
-
-  \end{minipage}%
-  %
-  \begin{minipage}[c]{0.2\linewidth}
-
-  {\centering
-
-  \[
-  x(0) = 1
-  \]
-
-  }
-
-  \end{minipage}%
-
-  \subcaption{\label{fig-first-order}First-order system of an exponential decay. The variable
-  \texttt{x} stores the initial condition for \(x\), and the variable
-  \texttt{eq} stores the rate equation for \(x\).}
-
-  }
-
-  \end{minipage}%
-  \newline
-  \begin{minipage}[t]{\linewidth}
-
-  {\centering
-
-  \begin{minipage}[c]{0.6\linewidth}
-
-  {\centering
-
-  \begin{CodeInput}
-  \begin{Highlighting}[]
-  \KeywordTok{class} \ClassTok{Oscillator}\KeywordTok{(}\ClassTok{System}\KeywordTok{)}\OperatorTok{:}
-      \VariableTok{x}: \ClassTok{Variable} \OperatorTok{=} \FunctionTok{initial}\KeywordTok{(\VariableTok{default}\OperatorTok{=}\ValueTok{1})}
-      \VariableTok{v}: \ClassTok{Derivative} \OperatorTok{=} \VariableTok{x}.\FunctionTok{derive}\KeywordTok{(\VariableTok{initial}\OperatorTok{=}\ValueTok{0})}
-      \VariableTok{eq} \OperatorTok{=} \VariableTok{v}.\FunctionTok{derive}\KeywordTok{()} \OperatorTok{\textless{}\textless{}} \OperatorTok{-}\VariableTok{x}
-  \end{Highlighting}
-  \end{CodeInput}
-
-  }
-
-  \end{minipage}%
-  %
-  \begin{minipage}[c]{0.2\linewidth}
-
-  {\centering
-
-  \[
-  \frac{d^2x}{dt^2} = -x
-  \]
-
-  }
-
-  \end{minipage}%
-  %
-  \begin{minipage}[c]{0.2\linewidth}
-
-  {\centering
-
-  \[
-  \begin{cases}
-      x(0) &= 1 \\
-      \frac{dx}{dt}(0) &= 0
-  \end{cases}
-  \]
-
-  }
-
-  \end{minipage}%
-
-  \subcaption{\label{fig-second-order}Second-order system of an harmonic oscillator, where the
-  variable \texttt{v} stores the derivative \(\frac{dx}{dt}\) and the rate
-  equation is specified for the derivative \texttt{v}.}
-
-  }
-
-  \end{minipage}%
-  \newline
-  \begin{minipage}[t]{\linewidth}
-
-  {\centering
-
-  \begin{minipage}[c]{0.6\linewidth}
-
-  {\centering
-
-  \begin{CodeInput}
-  \begin{Highlighting}[]
-  \KeywordTok{class} \ClassTok{BigModel}\KeywordTok{(}\ClassTok{System}\KeywordTok{)}:
-      \VariableTok{x}: \ClassTok{Variable} \OperatorTok{=} \FunctionTok{initial}\KeywordTok{(\VariableTok{default}\OperatorTok{=}\ValueTok{1})}
-      \VariableTok{linked} \OperatorTok{=} \ClassTok{Decay}\KeywordTok{(\VariableTok{x}\OperatorTok{=}\VariableTok{x})}
-      \VariableTok{independent} \OperatorTok{=} \ClassTok{Decay}\KeywordTok{(\VariableTok{x}\OperatorTok{=}\ValueTok{2})}
-  \end{Highlighting}
-  \end{CodeInput}
-
-  }
-
-  \end{minipage}%
-  %
-  \begin{minipage}[c]{0.2\linewidth}
-
-  {\centering
-
-  \[
-  \begin{cases}
-      \frac{dx}{dt} = -x \\
-      \frac{dx_{ind}}{dt} = -x_{ind}
-  \end{cases}
-  \]
-
-  }
-
-  \end{minipage}%
-  %
-  \begin{minipage}[c]{0.2\linewidth}
-
-  {\centering
-
-  \[
-  \begin{cases}
-      x(0) &= 1 \\
-      x_{ind}(0) &= 2
-  \end{cases}
-  \]
-
-  }
-
-  \end{minipage}%
-
-  \subcaption{\label{fig-composition}First-order system of two exponential decays by composition of
-  the \texttt{Decay} system. The subsystem \texttt{linked} has a reference
-  to the outer variable \texttt{x}, while the subsystem
-  \texttt{independent} defines a new variable with initial condition $2$,
-  which on the corresponding mathematical expression was named \(x_{ind}\).}
-
-  }
-
-  \end{minipage}%
-
-  \caption{\label{fig-poincare}Code and corresponding mathematical
-  expressions for different systems.}
-
-  \end{figure*}
-
-% FIGURE 2
-\begin{figure*}[t]
-
-  \begin{minipage}[b]{0.50\linewidth}
-  {
-      \centering 
-      \includegraphics[width=\textwidth]{ide1.pdf}
-  }
-  \end{minipage}%
-  %
-  \begin{minipage}[b]{0.50\linewidth}
-  {
-      \centering
-      \includegraphics[width=\textwidth]{ide2.pdf}
-  }
-  \end{minipage}%
-
-  \caption{
-      \label{fig-ide}
-      Screenshots of Visual Studio Code showing
-      tooltips (green arrows) and
-      highlighted type errors (red arrows).
-      On the left,
-      we show that \texttt{a},
-      a \texttt{Constant} assigned with \texttt{assign(...,\ constant=True)},
-      can be used for \texttt{Variable b}'s initial condition.
-      Instead,
-      it is flagged as a type error (red underlining)
-      when using \texttt{c}, a \texttt{Parameter},
-      for \texttt{Variable d}'s initial condition,
-      The IDE automatically recognizes \texttt{e} as an \texttt{Equation},
-      and provides autocompletion of the \texttt{Model}'s components.
-      A tooltip is shown when composing models (green arrow, right),
-      which show the expected variables and their default values.
-      The IDE highlights wrong names (\texttt{z} is not a name in \texttt{Model})
-      and mismatched types (\texttt{x} is \texttt{Variable} and \texttt{a} must be a number or a \texttt{Constant})
-  }
-
-  \end{figure*}
-
-% FIGURE 3
-\begin{figure*}[t]
-
-  \begin{minipage}[t]{\linewidth}
-
-  {\centering
-
-  \begin{CodeInput}
-  \begin{Highlighting}[]
-  \KeywordTok{import} \ClassTok{numpy} \KeywordTok{as} \ClassTok{np}
-  \KeywordTok{from} \ClassTok{poincare} \KeywordTok{import} \ClassTok{Simulator}
-
-  \VariableTok{sim} \OperatorTok{=} \ClassTok{Simulator}\KeywordTok{(\ClassTok{Oscillator})}
-  \VariableTok{df} \OperatorTok{=} \VariableTok{sim}.\FunctionTok{solve}\KeywordTok{(}\VariableTok{save\_at}\OperatorTok{=}\ClassTok{np}.\FunctionTok{linspace}\KeywordTok{(}\ValueTok{0}, \ValueTok{5}, \ValueTok{101}\KeywordTok{))}
-  \end{Highlighting}
-  \end{CodeInput}
-
-  }
-
-  \end{minipage}%
-  \newline
-  \begin{minipage}[t]{0.45\linewidth}
-
-  {\centering
-
-  \begin{CodeInput}
-  \begin{Highlighting}[]
-  \VariableTok{df}.\FunctionTok{head}\KeywordTok{()}
-  \end{Highlighting}
-  \end{CodeInput}
-
-  \begin{CodeOutput}
-      \begin{tabular}[]{@{}lll@{}}
-          \toprule
-          & x & v \\
-          time & & \\
-          \midrule
-          0.00 & 1.000000 & 0.000000 \\
-          0.05 & 0.998750 & -0.049979 \\
-          0.10 & 0.995005 & -0.099835 \\
-          0.15 & 0.988772 & -0.149440 \\
-          0.20 & 0.980069 & -0.198672 \\
-          \botrule
-      \end{tabular}
-  \end{CodeOutput}
-
-  }
-
-  \end{minipage}%
-  \hfill
-  \begin{minipage}[t]{0.45\linewidth}
-
-  {\centering
-
-  \begin{CodeInput}
-  \begin{Highlighting}[]
-  \VariableTok{df}.\FunctionTok{plot}\KeywordTok{()}\OperatorTok{;}
-  \end{Highlighting}
-  \end{CodeInput}
-
-  \begin{CodeOutput}
-    \includegraphics[width=\textwidth]{cell-output}
-  \end{CodeOutput}
-
-  }
-
-  \end{minipage}%
-
-  \caption{\label{fig-sim}Simulation of the \texttt{Oscillator} system
-  from Figure~\ref{fig-second-order}. The output is a
-  \texttt{pandas.DataFrame} with a column for each variable and the time
-  as index. It is inspected and plotted with the \texttt{pandas} methods.}
-
-  \end{figure*}
-
-% FIGURE 4
-\begin{figure*}[t]
-
-  \begin{minipage}[t]{\linewidth}
-    \centering 
-    \begin{CodeInput}
-    \begin{Highlighting}[]
-    \KeywordTok{from}\ClassTok{ simbio }\KeywordTok{import}\ClassTok{ Compartment, MassAction, Species, RateLaw, \FunctionTok{initial}}
-    \end{Highlighting}
-    \end{CodeInput}
-  \end{minipage}%
-  \newline
-  \begin{minipage}[t]{\linewidth}
-    \centering 
-    \begin{minipage}[c]{0.60\linewidth}
-      \centering 
-      \begin{CodeInput}
-      \begin{Highlighting}[]
-        \KeywordTok{class}\ClassTok{ Model}\KeywordTok{(}\ClassTok{Compartment}\KeywordTok{)}:
-            \CommentTok{"""2A {-}\textgreater{} B"""}
-
-        \VariableTok{    A}: \ClassTok{Species }\OperatorTok{=}\FunctionTok{ initial}\KeywordTok{(}\VariableTok{default}\OperatorTok{=}\ValueTok{1}\KeywordTok{)}
-        \VariableTok{    B}: \ClassTok{Species }\OperatorTok{=}\FunctionTok{ initial}\KeywordTok{(}\VariableTok{default}\OperatorTok{=}\ValueTok{0}\KeywordTok{)}
-        \VariableTok{    r }\OperatorTok{=}\ClassTok{ RateLaw}\KeywordTok{(}
-        \VariableTok{        reactants}\OperatorTok{=}\KeywordTok{[}\ValueTok{2} \OperatorTok{*}\VariableTok{ A}\KeywordTok{]},
-        \VariableTok{        products}\OperatorTok{=}\NormalTok{\KeywordTok{[}B\KeywordTok{]},}
-        \VariableTok{        rate}\OperatorTok{=}\ValueTok{1}\NormalTok{,}
-        \KeywordTok{    )}
-      \end{Highlighting}
-      \end{CodeInput}
-    \end{minipage}%
-    %
-    \begin{minipage}[c]{0.40\linewidth}
-      \centering 
-      \[
-      \begin{cases}
-          \frac{dA}{dt} = -2 \\
-          \frac{dB}{dt} = +1
-      \end{cases}
-      \begin{cases}
-          A(0) &= 1 \\
-          B(0) &= 0
-      \end{cases}
-      \]
-    \end{minipage}%
-  \end{minipage}%
-  \newline
-  \begin{minipage}[t]{\linewidth}
-      \centering 
-      \begin{minipage}[c]{0.60\linewidth}
-          \centering 
-          \begin{CodeInput}
-          \begin{Highlighting}[]
-          \KeywordTok{class}\ClassTok{ Model}\KeywordTok{(}\ClassTok{Compartment}\KeywordTok{)}:
-              \CommentTok{"""2A {-}\textgreater{} B"""}
-
-          \VariableTok{    A}: \ClassTok{Species }\OperatorTok{=}\FunctionTok{ initial}\KeywordTok{(}\VariableTok{default}\OperatorTok{=}\ValueTok{1}\KeywordTok{)}
-          \VariableTok{    B}: \ClassTok{Species }\OperatorTok{=}\FunctionTok{ initial}\KeywordTok{(}\VariableTok{default}\OperatorTok{=}\ValueTok{0}\KeywordTok{)}
-          \VariableTok{    r }\OperatorTok{=}\ClassTok{ MassAction}\KeywordTok{(}
-          \VariableTok{        reactants}\OperatorTok{=}\KeywordTok{[}\ValueTok{2} \OperatorTok{*}\VariableTok{ A}\KeywordTok{]},
-          \VariableTok{        products}\OperatorTok{=}\NormalTok{\KeywordTok{[}B\KeywordTok{]},}
-          \VariableTok{        rate}\OperatorTok{=}\ValueTok{1}\NormalTok{,}
-          \KeywordTok{    )}
-          \end{Highlighting}
-          \end{CodeInput}
-      \end{minipage}%
-      %
-      \begin{minipage}[c]{0.40\linewidth}
-          \centering 
-          \[
-          \begin{cases}
-              \frac{dA}{dt} = -2 A^2 \\
-              \frac{dB}{dt} = +1 A^2
-          \end{cases}
-          \begin{cases}
-              A(0) &= 1 \\
-              B(0) &= 0
-          \end{cases}
-          \]
-      \end{minipage}%
-  \end{minipage}%
-
-  \caption{
-    A reaction system for species \(A\) and \(B\)
-    with initial conditions \(1\) and \(0\), respectively. A single reaction
-    transforming \(2A\) into \(B\) is saved in variable \texttt{r}. The rate
-    \(1\) is specified directly for \texttt{RateLaw}, and is proportional to
-    the reactants for \texttt{MassAction}.
-  }
-  \label{fig-simbio}
-\end{figure*}
-
-% FIGURE 5
-\begin{figure*}[t]
-  \centering
-
-  \begin{CodeInput}
-  \begin{Highlighting}[]
-  \KeywordTok{from}\ClassTok{ simbio.io }\KeywordTok{import}\ClassTok{ biomodels, sbml}
-
-  \VariableTok{model }\OperatorTok{=}\ClassTok{ sbml}.\FunctionTok{load}\KeywordTok{(}\StringTok{"repressilator.sbml"}\KeywordTok{)}  \CommentTok{\# from existing SBML file}
-  \VariableTok{model }\OperatorTok{=}\ClassTok{ biomodels}.\FunctionTok{load}\KeywordTok{(}\StringTok{"BIOMD12"}\KeywordTok{)}  \CommentTok{\# from BioModels}
-  \VariableTok{model}
-  \end{Highlighting}
-  \end{CodeInput}
-
-  \begin{CodeOutput}
-      \begin{tabular}{@{}lll@{}}
-          \multicolumn{3}{l}{Elowitz2000 - Repressilator} \\
-          \toprule
-          type & total & names \\
-          \midrule
-          variables  &  6 & PX, PY, PZ, X, Y, Z \\
-          parameters & 17 & cell, beta, alpha0, alpha, eff, n, KM, tau\_mRNA, tau\_prot, ... \\
-          equations  & 12 & Reaction1, Reaction2, Reaction3, Reaction4, Reaction5, ... \\
-          \botrule
-      \end{tabular}
-  \end{CodeOutput}
-
-  \caption{Creation of a model from a local SBML file or one uploaded to BioModels.}
-  \label{fig-simbio-io}
-\end{figure*}
-
-% FIGURE 6
-\begin{figure*}[t]
-  \centering
-  \includegraphics[width=\textwidth]{performance}
-  \caption{
-    Performance of different softwares to solve models from the curated section of BioModels.
-    (left) Run time for the model BIOMD3 as a function of the number of output points.
-    (right) Run time for different models for 300 output points,
-    using the geometric mean of the different softwares to order them.
-    Each point corresponds to the median of 20 runs,
-    with a neglibible errorbar given by the interquartile range.
-  }
-  \label{fig-runtime}
-\end{figure*}
-
 \end{document}

### 2024-06-15 final/oup-authoring-template.tex
Update captions reflecting order changes in figures

@@
 
-  \subcaption{\label{fig-first-order}First-order system of an exponential decay. The variable
-  \texttt{x} stores the initial condition for \(x\), and the variable
-  \texttt{eq} stores the rate equation for \(x\).}
+  \subcaption{
+    \label{fig-first-order}
+    First-order system of an exponential decay.
+    The variable \texttt{x} stores the initial condition for \(x\),
+    and the variable \texttt{eq} stores its rate equation.
+  }
   \end{minipage}%
@@
 
-  \subcaption{\label{fig-composition}First-order system of two exponential decays by composition of
-  the \texttt{Decay} system. The subsystem \texttt{linked} has a reference
-  to the outer variable \texttt{x}, while the subsystem
-  \texttt{independent} defines a new variable with initial condition $2$,
-  which on the corresponding mathematical expression was named \(x_{ind}\).}
+  \subcaption{
+    \label{fig-composition}
+    First-order system of two exponential decays by composition of the \texttt{Decay} system.
+    The subsystem \texttt{linked} has a reference to the outer variable \texttt{x},
+    while the subsystem \texttt{independent} defines a new variable with initial condition $2$,
+    which on the corresponding mathematical expression was named \(x_{indep}\).
+  }
   \end{minipage}%
 
-  \caption{\label{fig-poincare}Code and corresponding mathematical
-  expressions for different systems.}
+  \caption{
+    \label{fig-poincare}
+    Code and corresponding mathematical expressions for different systems.
+  }
 
@@
     highlighted type errors (red arrows).
-    On the left,
+    Above,
     we show that \texttt{a},
@@
     and provides autocompletion of the \texttt{Model}'s components.
-    A tooltip is shown when composing models (green arrow, right),
+    A tooltip is shown when composing models (green arrow, below),
     which show the expected variables and their default values.
@@
 
-\CommentTok{\# from existing SBML file}
+\CommentTok{\# from a local SBML file}
 \VariableTok{model} \OperatorTok{=} \ClassTok{sbml}.\FunctionTok{load}\KeywordTok{(}\StringTok{"repressilator.sbml"}\KeywordTok{)}
@@
 
-  \caption{Creation of a model from a local SBML file or one uploaded to BioModels.}
+  \caption{Creation of a model from a local SBML file or from one uploaded to BioModels.}
   \label{fig-simbio-io}

### 2024-06-15 final/oup-authoring-template.tex
Spellchecing

@@
 As such, it is not possible to apply custom optimizations to the equations
-or use automatic differentation packages such as JAX \citep{jax2018github} to compute the model's jacobian.
+or use automatic differentiation packages such as JAX \citep{jax2018github} to compute the model's jacobian.
 
@@
 using variables, parameters and constants,
-and assiging rate equations to variables.
+and assigning rate equations to variables.
 For defining \acp{CRN},
@@
 Poincare is a package to define and simulate dynamical systems.
-Based on Python inmutable dataclasses,
+Based on Python immutable dataclasses,
 it provides a \texttt{System} class,
@@
     Each point corresponds to the median of 20 runs,
-    with a neglibible errorbar given by the interquartile range.
+    with a negligible error-bar given by the interquartile range.
   }
@@
 We speculate that this is due to memory allocation and data transfer in the Python bindings, 
-as the number of total function evaluations (i.e. including those performed by the integrator steper)
+as the number of total function evaluations (i.e. including those performed by the integrator stepper)
 is around 3000 and therefore much larger than most part of the x-axis.

### 2024-06-18 final/oup-authoring-template.tex
Add funding

@@
 
-\section{Author contributions statement}
-
-Must include all authors, identified by initials, for example:
-S.R. and D.A. conceived the experiment(s),  S.R. conducted the experiment(s), S.R. and D.A. analysed the results.  S.R. and D.A. wrote and reviewed the manuscript.
-
 \section{Acknowledgments}
-The authors thank the anonymous reviewers for their valuable suggestions. This work is supported in part by funds from the National Science Foundation (NSF: \# 1636933 and \# 1920920).
+The authors thank the anonymous reviewers for their valuable suggestions.
+This work was supported in part by funds from
+the Agencia I+D+i (PICT-2018-01516),
+and Universidad de Buenos Aires (UBAINT Doctoral, REREC-2023-593-E-UBA-REC).
 

### 2024-06-18 final/oup-authoring-template.tex
Add structured abstract

@@
 
-%\abstract{
-%\textbf{Motivation:} .\\
-%\textbf{Results:} .\\
-%\textbf{Availability:} .\\
-%\textbf{Contact:} \href{name@email.com}{name@email.com}\\
-%\textbf{Supplementary information:} Supplementary data are available at \textit{Journal Name}
-%online.}
-
 \abstract{
+\textbf{Motivation:}
   \Acfp{CRN} play a pivotal role in diverse fields
@@
   limiting their extensibility and integration with the broader Python ecosystem.
+\\ \textbf{Results:}
   In response, we developed Poincaré and SimBio,
@@
   positioning Poincaré and SimBio as valuable tools for the modelling community.
+\\ \textbf{Availability:}
+  Poincaré and SimBio are released under the MIT license.
+  their source code is available on GitHub
+  (\url{https://github.com/maurosilber/poincare}
+  and \url{https://github.com/hgrecco/simbio}).
+  and can be installed from PyPI or conda-forge.
+\\ \textbf{Contact:} \href{email:hgrecco@df.uba.ar}{hgrecco@df.uba.ar}
 }
+
 \keywords{poincare, SimBio, Python, ODE}

### 2024-06-19 final/oup-authoring-template.tex
Add EBML funding

@@
 the Agencia I+D+i (PICT-2018-01516),
-and Universidad de Buenos Aires (UBAINT Doctoral, REREC-2023-593-E-UBA-REC).
+Universidad de Buenos Aires (UBAINT Doctoral, REREC-2023-593-E-UBA-REC),
+and EMBL core funding.
 

### 2024-06-19 final/oup-authoring-template.tex
Add Sheriff changes

--- 2024-06-18 final/oup-authoring-template.tex
@@
 
-\author[1,2]{Mauro Silberberg\ORCID{0000-0002-2402-1100}}
+\author[1,2,3]{Mauro Silberberg\ORCID{0000-0002-2402-1100}}
 \author[3]{Henning Hermjakob\ORCID{0000-0001-8479-0262}}
-\author[3]{Rahuman S. Malik-Sheriff\ORCID{0000-0003-0705-9809}}
+\author[3,4]{Rahuman S. Malik-Sheriff\ORCID{0000-0003-0705-9809}}
 \author[1,2,$\ast$]{Hernán E. Grecco\ORCID{0000-0002-1165-4320}}
@@
 
-\address[1]{\orgname{Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Física}, \orgaddress{\state{Buenos Aires}, \country{Argentina}}}
-\address[2]{\orgname{CONICET - Universidad de Buenos Aires, Instituto de Física de Buenos Aires (IFIBA)}, \orgaddress{\state{Buenos Aires}, \country{Argentina}}}
-\address[3]{\orgdiv{European Bioinformatics Institute}, \orgname{European Molecular Biology Laboratory (EMBL-EBI)}, \orgaddress{\street{Wellcome Genome Campus}, \state{Cambridge}, \country{UK}}}
+\address[1]{
+  \orgname{Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Física},
+  \orgaddress{\state{Buenos Aires}, \country{Argentina}}
+}
+\address[2]{
+  \orgname{CONICET - Universidad de Buenos Aires, Instituto de Física de Buenos Aires (IFIBA)},
+  \orgaddress{\state{Buenos Aires}, \country{Argentina}}
+}
+\address[3]{
+  \orgdiv{European Bioinformatics Institute},
+  \orgname{European Molecular Biology Laboratory (EMBL-EBI)},
+  \orgaddress{\street{Wellcome Genome Campus}, \state{Cambridge}, \country{UK}}
+}
+\address[4]{
+  Secondary Affiliation:
+  \orgdiv{Department of Surgery and Cancer, Faculty of Medicine},
+  \orgname{Imperial College London},
+  \orgaddress{\state{London}, \country{UK}}
+}
 
@@
 \section{Acknowledgments}
-The authors thank the anonymous reviewers for their valuable suggestions.
 This work was supported in part by funds from
 the Agencia I+D+i (PICT-2018-01516),
-and Universidad de Buenos Aires (UBAINT Doctoral, REREC-2023-593-E-UBA-REC).
+Universidad de Buenos Aires (UBAINT Doctoral, REREC-2023-593-E-UBA-REC),
+and EMBL core funding.
 
